---
title: "AI for Oncology Trial Design"
subtitle: "Graphical Multiplicity, Group Sequential Design, and AI Skills"
author: "Keaven Anderson"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)
```

## Abstract

With recent updates and planned AI agentic skills, we explore the ability of AI to accelerate, improve and make transparent clinical trial design for oncology.
Merck LDS Oncology has developed many conventions for trial design that can be learned by AI agents.
Supporting capabilities have been developed in our internal packages: gsDesign, gsDesign2, simtrial, wpgsd, and gMCPLite.
gsDesign was first developed in 2007 and has been an internal standard for years.
Some missing capabilities such as design under non-proportional hazards and flexible options for skipping futility and efficacy analyses were initially developed in gsDesign2 and simtrial.
Recently, many of these options have been built into gsDesign; in addition, gsDesign now supports derivation of harm bounds using spending as suggested in recent FDA guidance.
Accounting for correlations in testing generated by group sequential design, testing in multiple hypotheses and testing multiple arms versus a common control are now available in wpgsd.
We will discuss the contribution of AI to many recent developments as well as the potential for building AI skills to simplify what otherwise would be tedious details required for complex design derivation.
Also, we mention possible replacement of gMCPLite with the graphicalMCP.

## Talk Roadmap

- Statistical challenge: multiple hypotheses + sequential monitoring
- Trial template: 3 endpoints x 2 populations (6 hypotheses)
- Core workflow from graphicalMCPgsd2
- Sample-size derivation cascade from one driving design
- Testing and verification workflow
- Alternative implementations: gsDesign, rpact, wpgsd
- AI contributions and AI skill architecture
- Summer intern direction

## Why This Is Hard

- Need strong family-wise Type I error control with multiple hypotheses
- Need flexibility for interim analyses with unpredictable event accrual
- Need consistency across endpoints and populations
- Need transparent and auditable testing decisions
- Need practical, reusable implementation templates

## Package Evolution

- gsDesign (2007+): internal standard for sequential design
- gsDesign2 + simtrial: extended options and design under non-proportional hazards
- gsDesign updates: broader capability, including harm-bound derivation by spending
- wpgsd: correlation-aware testing with multiple hypotheses/arms
- graphicalMCP: modern graph-first multiplicity workflow

## Trial Template (Example)

- Endpoints: OS, PFS, ORR
- Populations: BM+, All subjects
- Hypotheses:
  - H1: OS BM+
  - H2: OS All
  - H3: PFS BM+
  - H4: PFS All
  - H5: ORR BM+
  - H6: ORR All

## Multiplicity Graph Setup (Code)

```{r multiplicity-graph-code, eval=FALSE}
# Hypothesis names
nameHypotheses <- c(
  "H1: OS\n BM+",
  "H2: OS\n All subjects",
  "H3: PFS\n BM+",
  "H4: PFS\n All subjects",
  "H5: ORR\n BM+",
  "H6: ORR\n All subjects"
)

# Transition matrix for alpha reallocation
m <- matrix(c(
  0, 1, 0, 0, 0, 0,
  0, 0, .5, .5, 0, 0,
  0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, .5, .5,
  0, 0, 0, 0, 0, 1,
  .5, .5, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)

alphaHypotheses <- c(.01, .01, .004, 0.000, 0.0005, .0005)
fwer <- sum(alphaHypotheses)

g0 <- graphicalMCP::graph_create(
  hypotheses = stats::setNames(alphaHypotheses / fwer, nameHypotheses),
  transitions = m,
  hyp_names = nameHypotheses
)
```

## Sample-Size Driving Design (H1)

- H1 (OS in BM+) is the sample-size-driving design
- Design target: 90% power, one-sided alpha for H1
- Output from H1 determines BM+ enrollment profile and BM+ sample size
- All other hypotheses derive from this design

```{r h1-driving-code, eval=FALSE}
prevalence_bm <- 0.50

ossub <- gsDesign2::gs_design_ahr(
  enroll_rate = enroll_rate_bm,
  fail_rate = fail_rate_os_bm,
  alpha = alphaHypotheses[1],
  beta = 0.1,
  binding = FALSE,
  analysis_time = c(24, 36, 42),
  info_frac = NULL,
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[1]),
  test_lower = FALSE
)

n_bm <- ceiling(max(ossub$analysis$n))
n_all <- ceiling(n_bm / prevalence_bm)
```

## Sample-Size Cascade Across All Hypotheses

- Prevalence link: $n_{All} = \lceil n_{BM+} / p_{BM+} \rceil$
- H2/H4 use stratified enrollment derived from H1 and prevalence
- H3 reuses BM+ enrollment fitted from H1
- H5 uses fixed design at $n_{BM+}$
- H6 uses stratified sample sizes split by BM+ prevalence

```{r cascade-code, eval=FALSE}
# Fitted BM+ enrollment from H1
enroll_rate_bm_fitted <- ossub$enroll_rate

# Overall-population stratified enrollment derived from H1 + prevalence
enroll_rate_strat <- gsDesign2::define_enroll_rate(
  stratum = rep(c("BM+", "BM-"), each = nrow(enroll_rate_bm_fitted)),
  duration = rep(enroll_rate_bm_fitted$duration, 2),
  rate = c(
    enroll_rate_bm_fitted$rate,
    enroll_rate_bm_fitted$rate * (1 - prevalence_bm) / prevalence_bm
  )
)

# ORR split
n_bm_neg <- n_all - n_bm
```

## Analysis Timing and Spending Time

- IA1: 6 months after final patient enrolled
- IA2: 14 months after enrollment completion, with event-trigger allowance
- Final: 24 months after enrollment completion, with delay allowance
- Spending time at interims:
  - $t_k = \min(t_k^{planned}, n_k / N_K)$
- Spending time at final: $t_K = 1$

## Sequential P-values

- Group sequential hypotheses (H1-H4): compute one sequential p-value per hypothesis
- Single-look hypotheses (H5-H6): sequential p-value equals nominal p-value
- Sequential p-values are then passed to graphical multiplicity testing

```{r seqp-code, eval=FALSE}
EOCtab$seqp[EOCtabline] <- ifelse(
  is.null(gsD2list[[EOCtabline]]),
  EOCtab$nominalP[EOCtabline],
  gsDesign2::sequential_pval(
    gs_design = gsD2list[[EOCtabline]],
    event = tem$events,
    z = -stats::qnorm(tem$nominalP),
    ustime = tem$spendingTime,
    interval = c(1e-05, 0.9999)
  )
)
```

## Graphical Testing and Updates

```{r graph-test-code, eval=FALSE}
result <- graphicalMCP::graph_test_shortcut(
  graph = g0,
  p = EOCtab$seqp,
  alpha = fwer,
  verbose = TRUE
)

# Update graph through rejections
gu <- graphicalMCP::graph_update(g0, delete = rejected_order)
```

- Produces rejected hypotheses, adjusted p-values, and graph sequence
- Final graph determines maximum alpha for non-rejected hypotheses

## Verification Workflow

- For each group sequential hypothesis:
  - Update bounds with observed events and spending time
  - Compare observed nominal p-values to updated nominal boundaries
  - Compute repeated and sequential p-values by analysis
- Purpose: auditable and reproducible confirmation of decisions

```{r verify-code, eval=FALSE}
d2_upd <- gsDesign2::gs_update_ahr(
  x = d2,
  alpha = EOCtab$lastAlpha[i],
  ustime = usTime,
  event_tbl = data.frame(analysis = hresults$Analysis, event = n.I)
)
```

## Technical Problems Solved

- Integrating multiplicity graph logic with sequential testing
- Handling variable event accrual with pre-specified spending-time rules
- Supporting delayed analyses without invalidating Type I error control
- Enabling a robust verification path for every hypothesis decision
- Managing hypotheses with initial alpha equal to zero in design objects

## Alternative Implementations

- gsDesign variant: mature workflow with gMCPLite utilities
- rpact variant: alternative design engine and bound updates
- wpgsd variant: explicitly correlation-aware testing with iterative procedures
- graphicalMCP + gsDesign2: transparent modular workflow with sequential p-values

## Ease of Use and Reuse

- Reusable template structure for design, results entry, testing, and verification
- Standardized data structures reduce manual errors
- Clear separation of assumptions vs implementation logic
- Easier onboarding for statisticians through code patterns and examples

## AI Contribution So Far

- Faster generation of consistent code skeletons
- Better documentation alignment across related vignettes
- Rapid identification and correction of coding issues
- Automated support for repetitive derivations and formatting

## AI Skill Architecture for Trial Design

- Skill trigger: complex design requests with multiplicity + GSD
- Encoded workflow phases:
  - Design setup
  - Results entry
  - Hypothesis testing
  - Verification
- Reusable code patterns to reduce custom hand-coding

## Potential AI-Driven Future Workflow

1. User specifies clinical assumptions and decision priorities
2. AI generates multiplicity graph and design objects
3. AI generates analysis and verification scaffolding
4. User reviews, modifies assumptions, and reruns
5. AI regenerates with full traceability

## Summer Intern Focus

- Expand and harden AI trial-design skills
- Capture Merck LDS Oncology conventions explicitly
- Improve validation and robustness checks
- Build practical prompts/templates for day-to-day trial design work

## Discussion

- Which parts of setup are still too manual?
- Where do we want stronger guardrails?
- What should be fully automated vs always user-reviewed?

## Backup: Illness-Death Model (Single Treatment Group)

```{r illness-death-network, fig.width=8.5, fig.height=6}
par(mar = c(0, 0, 0, 0))
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))

# Diamond layout of the four states
sx <- c(0.50, 0.12, 0.88, 0.50)
sy <- c(0.88, 0.50, 0.50, 0.12)
r <- 0.10

theta <- seq(0, 2 * pi, length.out = 100)
state_colors <- c("#56B4E9", "#009E73", "#E69F00", "#999999")
state_labels <- c(
  "Alive\nNo response\nNo progression",
  "Responded\nNo progression",
  "Progressed",
  "Dead"
)

for (i in 1:4) {
  polygon(
    sx[i] + r * cos(theta),
    sy[i] + r * 0.9 * sin(theta),
    col = adjustcolor(state_colors[i], alpha.f = 0.3),
    border = state_colors[i],
    lwd = 2
  )
  text(sx[i], sy[i], state_labels[i], cex = 0.8, font = 2)
}

# Draw one transition arrow with optional curvature
arrow_trans <- function(i, j, label_expr, label_pos = 0.5,
                        label_offset = c(0, 0), curve = 0, ...) {
  dx <- sx[j] - sx[i]
  dy <- sy[j] - sy[i]
  ang <- atan2(dy, dx)

  x0 <- sx[i] + r * cos(ang)
  y0 <- sy[i] + r * 0.9 * sin(ang)
  x1 <- sx[j] - r * cos(ang)
  y1 <- sy[j] - r * 0.9 * sin(ang)

  if (curve == 0) {
    arrows(x0, y0, x1, y1, length = 0.10, lwd = 2, ...)
    lx <- x0 + label_pos * (x1 - x0) + label_offset[1]
    ly <- y0 + label_pos * (y1 - y0) + label_offset[2]
  } else {
    mx <- (x0 + x1) / 2 + curve * (y1 - y0)
    my <- (y0 + y1) / 2 - curve * (x1 - x0)
    xspline(c(x0, mx, x1), c(y0, my, y1), shape = c(0, 1, 0), border = "black", lwd = 2)
    ang2 <- atan2(y1 - my, x1 - mx)
    arrows(x1 - 0.01 * cos(ang2), y1 - 0.01 * sin(ang2), x1, y1, length = 0.10, lwd = 2)
    lx <- mx + label_offset[1]
    ly <- my + label_offset[2]
  }
  text(lx, ly, label_expr, cex = 0.95)
}

arrow_trans(1, 2, expression(lambda[resp]), label_pos = 0.45, label_offset = c(-0.06, 0.04))
arrow_trans(1, 3, expression(lambda["prog,0"]), label_pos = 0.45, label_offset = c(0.06, 0.04))
arrow_trans(1, 4, expression(lambda["death,0"]), label_pos = 0.50, label_offset = c(0.08, 0))
arrow_trans(2, 3, expression(lambda["prog,1"]), label_pos = 0.50, label_offset = c(0, 0.06))
arrow_trans(2, 4, expression(lambda["death,1"]), label_pos = 0.45, label_offset = c(-0.06, -0.04))
arrow_trans(3, 4, expression(lambda["death,2"]), label_pos = 0.45, label_offset = c(0.06, -0.04))
```

## Backup: Transition Rate Calibration

```{r transition-rates-code, eval=FALSE}
transition_rate <- build_transition_rates(
  strata = c("BM+", "BM-"),
  treatments = c("control", "experimental"),
  median_pfs = c("BM+" = pfsmedian, "BM-" = pfsmedian),
  median_os = c("BM+" = osmedian, "BM-" = osmedian),
  orr = list(
    "BM+" = c(control = orr_ctrl_bm_pos, experimental = orr_exp_sim_bm_pos),
    "BM-" = c(control = orr_ctrl_bm_neg, experimental = orr_exp_sim_bm_neg)
  ),
  hr_pfs = c("BM+" = hr_pfs_sim_bm_pos, "BM-" = hr_pfs_sim_bm_neg),
  hr_os = c("BM+" = hr_os_sim_bm_pos, "BM-" = hr_os_sim_bm_neg)
)
```

## Backup: Theoretical Curves and Full Verification

- Theoretical PFS/OS curves from illness-death CDFs can be shown before simulation
- Full verification tables (bounds, repeated p, sequential p) are retained in the vignette workflow
- Backup material available for discussion if the audience wants implementation detail
