---
title: "Iterative graphical testing for group sequential design using graphicalMCP and wpgsd"
subtitle: "Weighted parametric group sequential design with correlated tests"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
bibliography: gMCPLite.bib
vignette: >
  %\VignetteIndexEntry{Iterative graphical testing for group sequential design using graphicalMCP and wpgsd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1,
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)
```

## Overview

This document is intended to evaluate statistical significance for graphical multiplicity control when used with group sequential design [@MaurerBretz2013]. In particular, we demonstrate design and analysis of a complex oncology trial.
There are many details building on the necessarily simple example provided by @MaurerBretz2013.
The combination of tools provided by the wpgsd and gsDesign packages is non-trivial, but developed in a way that is meant to be re-used in a straightforward fashion.
This has been found to be particularly valuable to provide a prompt and verifiable conclusion in multiple trials such as @KEYNOTE048 where 14 hypotheses were evaluated using a template such as this.

This vignette focuses on combining:

- graphical multiplicity updating using **graphicalMCP** (graph specification and alpha reallocation),
- weighted parametric group sequential testing using **wpgsd**.

For comparison, see:

- [Graphical testing for group sequential design](GraphicalMultiplicity.html) (original approach),
- [Group sequential design using the graphicalMCP package using hgraph](graphicalMCPgsd.html) (graphicalMCP shortcut testing variant).

The development version of wpgsd can be installed with `remotes::install_github("keaven/wpgsd")`.

Given the complexity involved, substantial effort has been taken to provide methods to check hypothesis testing.

- Under the overall graphical procedure, the current Type I error available for each hypothesis changes as hypotheses are rejected.
- For each endpoint family (OS and PFS), weighted parametric group sequential bounds are computed and evaluated using `wpgsd`.
- After any rejections, the overall graph is updated using `graphicalMCP`, and OS and PFS families are evaluated again until no further rejections occur.

The table of contents above lays out the organization of the document.
In short, we begin with 1) design specification followed by 2) results entry which includes event counts and nominal p-values for testing, 3) carrying out hypothesis testing, and 4) verification of the hypothesis testing results.

## Design

For the template example, there are 3 endpoints and 2 populations resulting in 6 hypotheses to be tested in the trial.
The endpoints are:

- Overall survival (OS)
- Progression free survival (PFS)
- Objective response rate (ORR)

The populations to be studied are:

- The overall population (All subjects)
- A subgroup (Subgroup)

For simplicity, we design assuming the control group has an exponential time to event with a median of 12 months for OS and 5 months for PFS.
We design under a proportional hazards assumption.
ORR for the control group is assumed to be 15%.
Some of the choices here are arbitrary, but the intent is to fully specify how patients will be enrolled and followed for $\alpha$-controlled study analyses.

The following design characteristics are also specified to well-characterize outcomes for all subjects by the end of the trial:

- Enrollment is assumed to occur over 18 months. Enrollment will continue until the targeted number of subjects has been enrolled in the subgroup to ensure power as planned for that population. This means, the overall population sample size will be random and power may vary from that planned here.
The enrollment increases from 25% in the first two months to 50% in the second 2 months to 75% in the third two months of the final steady state enrollment which continues from the end of month 6 of enrollment until final enrollment expected at 18 months.
- The first interim analysis will be conducted 6 months after final patient enrolled to adequately assess ORR for all patients. Thus, the analysis is planned at 24 months after start of study enrollment, but will be adapted according to when final enrollment is completed.
This is the only analysis for ORR and is an interim analysis for PFS and OS with whatever event counts are available at the cutoff.
- The second interim analysis will be conducted 14 months after final enrollment to ensure minimum follow-up almost 3 times the assumed control median PFS for all subjects. This would be delayed up to 3 months if the final targeted event count for PFS in the subgroup is not achieved at that time. This is to ensure a complete description of tail behavior for PFS in the case a PFS curve has a plateau. PFS and OS will be analyzed. The endpoint counts for OS and for the overall population for PFS are random since the cutoff is determined by the PFS endpoint count for the subgroup.
- The final analysis will be performed 24 months after final enrollment, ensuring 2 times the assumed median control survival as minimum follow-up for all subjects. Only analysis of OS is planned. The final analysis may be delayed up to 6 months if the targeted OS event count in the subgroup is not achieved. Thus, the planned total duration of the trial for the OS endpoint is 42 months.

The sample size for the trial will be driven by an adequate sample size and targeted events in the subgroup to ensure 90% power for the OS endpoint assuming a hazard ratio of 0.65.
For group sequential designs, we assume 1-sided testing.

To reveal code blocks for the remainder of the document, press the code buttons indicated throughout.
The initial code block sets options and loads needed packages; no modification should be required by the user.

```{r, message=FALSE, warning=FALSE}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
# Prefer fixed notation
old <- options(scipen = 999)
# Colorblind palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# 2 packages used for data storage and manipulation: dplyr, tibble
library(dplyr)
library(tibble)
# 2 packages used for R Markdown capabilities: knitr, kableExtra
library(knitr)
library(kableExtra)
library(gt)
library(ggplot2) # For plotting
library(gsDesign) # Group sequential design capabilities
library(graphicalMCP) # Graph-based multiplicity evaluation
library(wpgsd) # Weighted parametric group sequential design
library(simtrial) # Simulation of clinical trial data
library(survival) # Survival analysis (logrank test)
```

### Multiplicity diagram for hypothesis testing

Following is the multiplicity graph for the trial design.
We have arbitrarily split Type I error equally between the subgroup and overall populations.
Most $\alpha$ is allocated to OS and the least to ORR, with PFS receiving an intermediate amount.
This reflects the priority of the endpoints as well as the practicality to detect clinically significant differences in each population.
Reallocation for each endpoint proceeds from the subgroup to the overall population.
If the overall population hypothesis is rejected for a given endpoint, the reallocation is split between the two populations for another endpoint.
The choice for allocation and reallocation illustrated here is to demonstrate a complex multiplicity scenario; when actually applying this method the allocation and reallocation choices should be carefully considered.

```{r, warning=FALSE, message=FALSE, fig.asp=0.9}
### THIS CODE NEEDS TO BE MODIFIED FOR YOUR STUDY
# Hypothesis names
nameHypotheses <- c(
  "H1: OS\n Subgroup",
  "H2: OS\n All subjects",
  "H3: PFS\n Subgroup",
  "H4: PFS\n All subjects",
  "H5: ORR\n Subgroup",
  "H6: ORR\n All subjects"
)
# Number of hypotheses to be tested
nHypotheses <- length(nameHypotheses)
# Transition weights for alpha reallocation (square matrix)
m <- matrix(c(
  0, 1, 0, 0, 0, 0,
  0, 0, .5, .5, 0, 0,
  0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, .5, .5,
  0, 0, 0, 0, 0, 1,
  .5, .5, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)
# Initial Type I error assigned to each hypothesis (one-sided)
alphaHypotheses <- c(.01, .01, .004, 0.000, 0.0005, .0005)
fwer <- sum(alphaHypotheses)

g0 <- graphicalMCP::graph_create(
  hypotheses = alphaHypotheses / fwer,
  transitions = m,
  hyp_names = nameHypotheses
)

layout6 <- rbind(
  c(0.28, 0.78),
  c(0.72, 0.78),
  c(0.18, 0.50),
  c(0.82, 0.50),
  c(0.28, 0.22),
  c(0.72, 0.22)
)

plot(g0,
  layout = layout6,
  vertex.size = 45,
  vertex.color = cbPalette[c(2, 2, 4, 4, 3, 3)],
  vertex.label.cex = 0.9
)
```

This testing scheme can result in what might be referred to as time travel for passing of $\alpha$.
That is, if PFS hypotheses are not rejected at a given analysis (say final PFS analysis) and OS hypotheses are rejected at the final analysis, then the previously evaluated PFS tests at the interim and final PFS analysis can be compared to updated bounds based on reallocated Type I error.
While this practice was not encouraged by @MaurerBretz2013, it was acknowledged to control Type I error as previous discussed in [@LiuAnderson2008].
Given the stringent Type I error control for multiple hypotheses, the ability to acknowledge clinically significant differences as statistically significant can be important in describing treatment benefits in regulatory labeling for a treatment.

### Group sequential designs for each hypothesis {.tabset}

For the example, we assume 1-sided testing or a non-binding futility bound as required for a @MaurerBretz2013 design using group sequential design with graphical multiplicity control.
Each is demonstrated in the example code for respective hypotheses.
Efficacy $\alpha$-spending for all group sequential designs uses the @LanDeMets spending function approximating an O'Brien-Fleming bound.

Note: `gsDesign::gsSurv()` supports multiple event-driven sample size methods, and its default is not `"Schoenfeld"`.
To align results with the `rpact`-based vignette, we set `method = "Schoenfeld"` in the designs below.

This section needs to be modified by the user to match the study design under consideration.
Those uncomfortable with coding may wish to design using the
[gsDesign Shiny app](https://gsdesign.shinyapps.io/prod/) which provides not only a point and click interface, but also a code tab that generates R code that can be copied and plugged in for designs below.

#### H1: OS, Subgroup

We assume 50% of the population is in the subgroup of interest.
A sample size of 378 is driven by overall survival (OS) in the subgroup where we assume a hazard ratio of 0.65.
Here we assume a one-sided group sequential design with no futility bound.

```{r, results='asis'}
osmedian <- 12 # Median control survival
# Derive group sequential design for OS in the targeted subgroup
ossub <- gsDesign::gsSurv(
  k = 3, # 3 analyses for OS
  test.type = 1, # Efficacy bound only (no futility)
  alpha = alphaHypotheses[1], # Allocated alpha from design hypothesis group
  beta = 0.1, # Type 2 error (1 - power)
  hr = 0.65, # Assumed hazard ratio for power calculation
  timing = c(0.61, 0.82), # Choose these to match targeted calendar timing of analyses
  sfu = sfLDOF, # Spending function to approximate O'Brien-Fleming bound
  lambdaC = log(2) / osmedian, # Exponential control failure rate
  eta = 0.001, # Exponential dropout rate
  gamma = c(2.5, 5, 7.5, 10), # Relative enrollment rates by time period
  R = c(2, 2, 2, 12), # Duration of time periods for rates in gamma
  T = 42, # Planned study duration for OS
  minfup = 24, # Planned minimum follow-up after end of enrollment
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(ossub)
rownames(tab) <- seq_len(nrow(tab))
cat(summary(ossub))
```

The above text was automatically generated and could be edited appropriately for description of the design.
Following is a summary table describing study bounds.

```{r}
# tab %>% kable(caption = "Design for OS in the subgroup.") %>% kable_styling()
tab %>%
  gt() %>%
  tab_header(title = "Design for OS in the Subgroup") %>%
  cols_align(align = "left", columns = Value) %>%
  tab_footnote(
    footnote = "Cumulative boundary crossing probability includes crossing probability at earlier analysis.",
    locations = cells_body(columns = "Value", rows = c(9, 10, 14, 15))
  ) %>%
  tab_footnote(
    footnote = "Approximate hazard ratio at bound.",
    locations = cells_body(columns = "Value", rows = c(3, 8, 13))
  )
```

#### H2: OS, All

The total sample size is assumed to be twice the above, N=756.
The power and hazard ratio can be adjusted to appropriately size the trial rather than starting with adjusting sample size to reach a targeted power.
For this example, we consider altering power (`beta`) while fixing the hazard ratio at 0.75, representing an increase in median OS from 12 months in the control group to 16 months in the experimental group.
For this design, we consider a non-binding futility bound where the trial may be stopped early in the overall population if the bound is crossed.
We use a @HwangShihDeCani bound with $\gamma = -3.25$.
Study designers should carefully consider implication for parameter choices, particularly if the futility bounds provide sensible guidance for stopping the trial.
Since the futility bounds are non-binding, the efficacy bound is computed assuming the futility bound is ignored which will control Type I error at the targeted level even if a futility bound is crossed and the trial is continued.

```{r, results = 'asis'}
hr <- .75
beta <- .14
os <- gsDesign::gsSurv(
  k = 3, test.type = 4, alpha = 0.01, beta = beta, hr = hr,
  timing = c(0.62, 0.83), sfu = sfLDOF,
  sfl = sfHSD, sflpar = -3.25,
  lambdaC = log(2) / 12, eta = 0.001, S = NULL,
  gamma = c(2.5, 5, 7.5, 10), R = c(2, 2, 2, 12),
  T = 42, minfup = 24,
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(os)
rownames(tab) <- seq_len(nrow(tab))
cat(summary(os))
```

```{r}
tab %>%
  kable(caption = "Design for OS in all subjects") %>%
  kable_styling()
```

We can also plot different design characteristics. Here, we plot the approximate hazard ratio to cross each bound which may be helpful for design team discussions.

```{r}
plot(os, plottype = "HR", xlab = "Events")
```

#### H3: PFS, Subgroup

For progression free survival (PFS) we assume a shorter median time to event of 5 months.
With an assumed hazard ratio of 0.65, we adjust `beta` and `timing` to match the targeted sample size and interim analysis timing.
We assume a larger dropout rate for PFS than we did for OS.
Here we set up a futility bound for safety.
This is an asymmetric 2-sided design with both futility and efficacy boundary crossing probabilities under the null hypothesis. The parameter `astar = 0.1` specifies total lower bound spending of 10%.
The lower @HwangShihDeCani spending bound with $\gamma = -8$ is intended to be conservative in terms of futility at the interim, but still provide a safety bound for the PFS result in this targeted population.
This would have to be carefully evaluated by the study design team at the time of design.

```{r, results='asis'}
hr <- .65
beta <- .149
pfssub <- gsDesign::gsSurv(
  k = 2, test.type = 6, astar = 0.1, alpha = 0.004, beta = beta, hr = hr,
  timing = .87, sfu = sfLDOF,
  sfl = sfHSD, sflpar = -8,
  lambdaC = log(2) / 5, eta = 0.02, S = NULL,
  gamma = c(2.5, 5, 7.5, 10), R = c(2, 2, 2, 12),
  T = 32, minfup = 14,
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(pfssub)
rownames(tab) <- seq_len(nrow(tab))
cat(summary(pfssub))
```

```{r}
tab %>%
  kable(caption = "Design for PFS in the subgroup") %>%
  kable_styling()
```

#### H4: PFS, All

Finally, we design for PFS in all subjects.
In this case, we simplify to a one-sided design.
A futility bound could be considered, if appropriate.

```{r}
hr <- .74
beta <- .15
pfs <- gsDesign::gsSurv(
  k = 2, test.type = 1, alpha = 0.004, beta = beta, hr = hr,
  timing = .86, sfu = sfLDOF,
  lambdaC = log(2) / 5, eta = 0.02, S = NULL,
  gamma = c(2.5, 5, 7.5, 10), R = c(2, 2, 2, 12),
  T = 32, minfup = 14,
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(pfs)
rownames(tab) <- seq_len(nrow(tab))
tab %>%
  kable(caption = "Design for PFS in the overall population") %>%
  kable_styling()
```

#### H5 and H6: ORR

For objective response rate (ORR), we assume an underlying control rate of 15%.
In the subgroup population, we have almost 90% power to detect a 20% improvement.

```{r}
nBinomial(p1 = .35, p2 = .15, alpha = .0005, n = 378)
```

In the all subjects population, we have approximately 95% power to detect an improvement in ORR from 15% to 30%.

```{r}
nBinomial(p1 = .3, p2 = .15, alpha = .0005, n = 756)
```

### Design list

Now we associate designs with hypotheses in an ordered list corresponding to the order in the multiplicity graph setup.
Since ORR designs are not group sequential, we enter `NULL` values for those in the last 2 entries of the design list; hit code button to reveal code for this.

```{r}
### THIS NEEDS TO BE MODIFIED TO MATCH STUDY
gsDlist <- list(ossub, os, pfssub, pfs, NULL, NULL)
```

### Spending plan and spending time

While it was relatively straightforward above to set up timing of analyses to match for the different hypotheses, accumulation of endpoints can vary from plan in a variety of ways.
Planning on how to deal with this is critical at the time of protocol development to avoid later amendments or inappropriate $\alpha$-allocation to early analyses.
Before going into examples, we review the concept of $\alpha$-spending and what we will refer to as _spending time_.

For a given hypothesis, we will assign a non-decreasing spending function $f(t)$ defined for $t\ge 0$ with $f(0)=0$ and $f(t)=\alpha$ for $t\ge 1$.
We will assume $K$ analyses with observed event counts $n_k$ at analysis $k=1,2,\ldots,K$ and a targeted final event count of $N_k$.
The $\alpha$-spending at analysis $k$ was originally defined [@LanDeMets] as $f(t_k=n_k/N_K)$.
The values $n_k/N_K$ will be referred to as the _information fraction_, $k=1,\ldots,K$.
This is used to pre-specify the cumulative amount of Type I error for a hypothesis at each analysis.
In @LanDeMets1989 they noted that calendar time was another option for $t_k$ values, $k=1,\ldots,K.$
@PLWBook noted further that as long as $t_k$ is increasing with $k$, it can be used to define spending; this is subject to the requirement that under the null hypothesis, the timing must be selected in a way that is not correlated with the test statistic (e.g., blinded).
We will refer to $t_k$, regardless of its definition, as the _spending time_ for a hypothesis.
Note that the joint distribution of interim and final tests for a hypothesis is driven by $n_k$, $k=1,\ldots,K$.
This is equivalent to basing correlation on the information fraction $n_k^{(actual)}/n_K^{(planned)}$, $1\le k\le K$.
Thus, both spending time and information fraction are required to compute bounds for group sequential testing.
Our general objectives here will be to:

- Spend all Type I error for each hypothesis in its combined interim and final analyses; this requires the spending time to be 1 for the final analysis of a hypothesis.
- Ensure spending time is well defined for each analysis of each hypothesis.
- We will assume that both follow-up duration and event counts may be of interest in determining timing of analyses; e.g., for immuno-oncology therapies there have been delayed treatment effects and the tail of the time-to-event distribution has been important to establish benefit. Thus, we will assume here that over-spending at interim analysis is to be avoided.

Here we assume that the subgroup prevalence was over-estimated in the study design and indicating how spending time can be used to deal with this deviation from plan.

## Simulated trial data

We simulate a single trial instance using an illness-death model with response.
The model has four states: alive without response or progression (0),
responded without progression (1), progressed (2), and dead (3).

```{r, message=FALSE}
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "sim_illness_death.R"))
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "cut_illness_death.R"))
```

```{r}
prevalence_sub <- 0.50
pfsmedian <- 5

transition_rate <- build_transition_rates(
  strata = c("Subgroup", "Complement"),
  treatments = c("control", "experimental"),
  median_pfs = c("Subgroup" = pfsmedian, "Complement" = pfsmedian),
  median_os = c("Subgroup" = osmedian, "Complement" = osmedian),
  orr = list(
    "Subgroup" = c(control = 0.15, experimental = 0.35),
    "Complement" = c(control = 0.15, experimental = 0.25)
  ),
  hr_pfs = c("Subgroup" = 0.65, "Complement" = 0.85),
  hr_os = c("Subgroup" = 0.65, "Complement" = 0.85)
)
```

### Simulate one trial

```{r}
set.seed(20250328)
n_sub <- ceiling(ossub$n.I[ossub$k])  # Final targeted subgroup N from H1 design
n_total <- ceiling(n_sub / prevalence_sub)
steady_rate <- n_total / 15  # 18-month enrollment: 0.25*2 + 0.5*2 + 0.75*2 + 1*12 = 15 rate-months

sim_data <- sim_illness_death(
  n = n_total,
  stratum = data.frame(stratum = c("Subgroup", "Complement"),
                       p = c(prevalence_sub, 1 - prevalence_sub)),
  block = c("control", "control", "experimental", "experimental"),
  enroll_rate = data.frame(
    rate = steady_rate * c(0.25, 0.50, 0.75, 1.00),
    duration = c(2, 2, 2, 12)
  ),
  transition_rate = transition_rate,
  dropout_rate = 0.001
)

fpe <- max(sim_data$ENRLTIME)
cat("Total N:", nrow(sim_data), "\n")
cat("N Subgroup:", sum(sim_data$STRATUM == "Subgroup"),
    "N Complement:", sum(sim_data$STRATUM == "Complement"), "\n")
cat("FPE:", round(fpe, 1), "months\n")
```

### Determine analysis cut dates and compute p-values

```{r}
target_pfs_events_sub <- ceiling(max(pfssub$n.I))
target_os_events_sub <- ceiling(max(ossub$n.I))

analyses <- list(
  list(min_followup = 6, endpoint = NULL, event_target = NULL,
       target_stratum = NULL, max_followup = NULL),
  list(min_followup = 14, endpoint = "PFS", event_target = target_pfs_events_sub,
       target_stratum = "Subgroup", max_followup = 17),
  list(min_followup = 24, endpoint = "OS", event_target = target_os_events_sub,
       target_stratum = "Subgroup", max_followup = 30)
)

cut_dates <- get_analysis_dates(sim_data, analyses)

cat("Analysis cut dates:\n")
for (i in seq_along(cut_dates)) {
  cat(sprintf("  Analysis %d: %.1f months (FPE + %.1f)\n",
              i, cut_dates[i], cut_dates[i] - fpe))
}
```

```{r}
adtte <- lapply(seq_along(cut_dates), function(i) {
  d <- cut_illness_death(sim_data, cut_dates[i])
  d$ANALYSIS <- i
  d
})

# Helper: 1-sided logrank p-value
logrank_pval <- function(data, stratified = FALSE) {
  data$event <- 1L - data$CNSR
  if (stratified) {
    fit <- survdiff(Surv(AVAL, event) ~ TRT + strata(STRATUM), data = data)
  } else {
    fit <- survdiff(Surv(AVAL, event) ~ TRT, data = data)
  }
  obs_ctrl <- if (is.matrix(fit$obs)) sum(fit$obs[1, ]) else fit$obs[1]
  exp_ctrl <- if (is.matrix(fit$exp)) sum(fit$exp[1, ]) else fit$exp[1]
  z <- sign(obs_ctrl - exp_ctrl) * sqrt(fit$chisq)
  pnorm(-z)
}

# Helper: 1-sided rate difference p-value
rd_pval <- function(data, stratified = FALSE) {
  if (!stratified) {
    n_exp <- sum(data$TRT == "experimental")
    n_ctrl <- sum(data$TRT == "control")
    p_exp <- sum(data$AVAL[data$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(data$AVAL[data$TRT == "control"]) / n_ctrl
    p_pool <- (p_exp * n_exp + p_ctrl * n_ctrl) / (n_exp + n_ctrl)
    se <- sqrt(p_pool * (1 - p_pool) * (1 / n_exp + 1 / n_ctrl))
    if (se == 0) return(0.5)
    return(pnorm(-(p_exp - p_ctrl) / se))
  }
  strata <- unique(data$STRATUM)
  n_s <- rd_s <- var_s <- numeric(length(strata))
  for (k in seq_along(strata)) {
    ds <- data[data$STRATUM == strata[k], ]
    n_exp <- sum(ds$TRT == "experimental")
    n_ctrl <- sum(ds$TRT == "control")
    n_s[k] <- n_exp + n_ctrl
    p_exp <- sum(ds$AVAL[ds$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(ds$AVAL[ds$TRT == "control"]) / n_ctrl
    rd_s[k] <- p_exp - p_ctrl
    var_s[k] <- p_exp * (1 - p_exp) / n_exp + p_ctrl * (1 - p_ctrl) / n_ctrl
  }
  w <- n_s / sum(n_s)
  z <- sum(w * rd_s) / sqrt(sum(w^2 * var_s))
  pnorm(-z)
}

pvals <- list()
for (i in seq_along(cut_dates)) {
  d <- adtte[[i]]
  pvals[[paste0("H1_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS", STRATUM == "Subgroup"))
  pvals[[paste0("H2_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS"), stratified = TRUE)
  if (i <= 2) {
    pvals[[paste0("H3_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS", STRATUM == "Subgroup"))
    pvals[[paste0("H4_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS"), stratified = TRUE)
  }
  if (i == 1) {
    pvals[[paste0("H5_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR", STRATUM == "Subgroup"))
    pvals[[paste0("H6_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR"), stratified = TRUE)
  }
}
```

## Results entry at time of analysis

### Build results table

Event counts and p-values are assembled from the simulated trial data.
Spending time uses the subgroup information fraction for each hypothesis.

```{r,warnings=FALSE,message=FALSE}
get_events <- function(analysis_idx, paramcd, stratum = NULL) {
  d <- adtte[[analysis_idx]]
  if (!is.null(stratum)) d <- d %>% filter(STRATUM == stratum)
  sum(d$PARAMCD == paramcd & d$CNSR == 0, na.rm = TRUE)
}

events_os_sub <- sapply(1:3, function(i) get_events(i, "OS", "Subgroup"))
events_os_all <- sapply(1:3, function(i) get_events(i, "OS"))
events_pfs_sub <- sapply(1:2, function(i) get_events(i, "PFS", "Subgroup"))
events_pfs_all <- sapply(1:2, function(i) get_events(i, "PFS"))

# Update design objects with observed event counts
ossub$n.I <- events_os_sub
os$n.I <- events_os_all
pfssub$n.I <- events_pfs_sub
pfs$n.I <- events_pfs_all

inputResults <- tibble(
  H = c(rep(1, 3), rep(2, 3), rep(3, 2), rep(4, 2), 5, 6),
  Pop = c(
    rep("Subgroup", 3), rep("All", 3),
    rep("Subgroup", 2), rep("All", 2),
    "Subgroup", "All"
  ),
  Endpoint = c(rep("OS", 6), rep("PFS", 4), rep("ORR", 2)),
  nominalP = c(
    pvals$H1_1, pvals$H1_2, pvals$H1_3,
    pvals$H2_1, pvals$H2_2, pvals$H2_3,
    pvals$H3_1, pvals$H3_2,
    pvals$H4_1, pvals$H4_2,
    pvals$H5_1,
    pvals$H6_1
  ),
  Analysis = c(1:3, 1:3, 1:2, 1:2, 1, 1),
  events = c(ossub$n.I, os$n.I, pfssub$n.I, pfs$n.I, NA, NA),
  spendingTime = c(
    ossub$n.I / max(ossub$n.I),
    ossub$n.I / max(ossub$n.I),
    pfssub$n.I / max(pfssub$n.I),
    pfssub$n.I / max(pfssub$n.I),
    NA, NA
  )
)
kable(inputResults, caption = "Results from simulated trial data") %>%
  kable_styling()
```

## Testing hypotheses (iterative approach)

In this vignette we test OS and PFS hypotheses separately using `wpgsd` under the *current* Type I error allocation given by the overall graphical procedure. Whenever one or more hypotheses are rejected, we update the overall graph using `graphicalMCP` and repeat. This continues until no additional hypotheses can be rejected.

### Endpoint-specific setup for wpgsd

We treat OS (H1/H2) and PFS (H3/H4) as two separate 2-hypothesis graphical procedures, each with subgroup nested in overall population.

```{r,message=FALSE,warning=FALSE}
# Helper: correlation matrix for 2 nested populations (subgroup nested in overall)
make_corr_nested_2hyp <- function(n_sub, n_all) {
  stopifnot(length(n_sub) == length(n_all))
  n_sub <- as.numeric(n_sub)
  n_all <- as.numeric(n_all)
  k <- length(n_sub)

  # Test ordering matches wpgsd examples: H1_A1, H2_A1, H1_A2, H2_A2, ...
  n_vec <- as.vector(rbind(n_sub, n_all))
  overlap <- matrix(0, nrow = 2 * k, ncol = 2 * k)

  for (a in seq_len(k)) {
    for (b in seq_len(k)) {
      amin <- min(a, b)
      # Indices in the (H1,H2) x analyses ordering
      ia1 <- 2 * (a - 1) + 1
      ia2 <- 2 * (a - 1) + 2
      ib1 <- 2 * (b - 1) + 1
      ib2 <- 2 * (b - 1) + 2

      # subgroup-subgroup
      overlap[ia1, ib1] <- n_sub[amin]
      # overall-overall
      overlap[ia2, ib2] <- n_all[amin]
      # subgroup-overall (and overall-subgroup)
      overlap[ia1, ib2] <- n_sub[amin]
      overlap[ia2, ib1] <- n_sub[amin]
    }
  }

  corr <- overlap / sqrt(outer(n_vec, n_vec))
  # Ensure numerical symmetry
  corr <- (corr + t(corr)) / 2
  corr
}

# Helper: run wpgsd for a 2-hypothesis graph with H1 -> H2 recycling
run_wpgsd_2hyp <- function(alpha_total, w_local, p_obs, corr, info_frac) {
  if (is.na(alpha_total) || alpha_total <= 0) {
    return(list(bounds = NULL, closed_test = NULL, rejected = c(FALSE, FALSE)))
  }

  w_local <- as.numeric(w_local)
  if (length(w_local) != 2) stop("w_local must have length 2")
  if (any(w_local <= 0)) stop("w_local must be strictly positive for wpgsd")
  w_local <- w_local / sum(w_local)

  m_local <- matrix(c(
    0, 1,
    0, 0
  ), nrow = 2, byrow = TRUE)

  # Overall spending time (type = 2) uses the minimum information fraction
  # across hypotheses at each analysis
  t_overall <- pmin(info_frac[[1]], info_frac[[2]])

  bounds <- wpgsd::generate_bounds(
    type = 2,
    k = nrow(p_obs),
    w = w_local,
    m = m_local,
    corr = corr,
    alpha = alpha_total,
    sf = gsDesign::sfLDOF,
    sfparm = 0,
    t = t_overall
  )

  ct <- wpgsd::closed_test(bounds, p_obs)
  ct <- as.data.frame(ct)

  rejected <- c(
    any(as.character(ct$H1) == "Success"),
    any(as.character(ct$H2) == "Success")
  )

  list(bounds = bounds, closed_test = ct, rejected = rejected)
}

# Helper: run endpoint family under current alpha allocation.
# If one hypothesis has alpha = 0, fall back to single-hypothesis group sequential testing.
run_family_nested_2hyp <- function(alpha_vec, p_obs, corr, info_frac) {
  alpha_vec <- as.numeric(alpha_vec)
  if (length(alpha_vec) != 2) stop("alpha_vec must have length 2")

  if (any(alpha_vec < 0)) stop("alpha_vec must be nonnegative")
  if (sum(alpha_vec) <= 0) {
    return(list(bounds = NULL, closed_test = NULL, rejected = c(FALSE, FALSE)))
  }

  active <- alpha_vec > 0

  if (sum(active) == 2) {
    return(run_wpgsd_2hyp(
      alpha_total = sum(alpha_vec),
      w_local = alpha_vec,
      p_obs = p_obs,
      corr = corr,
      info_frac = info_frac
    ))
  }

  # Single-hypothesis fallback
  i <- which(active)[1]
  k <- nrow(p_obs)
  timing <- if (k > 1) info_frac[[i]][1:(k - 1)] else numeric(0)
  gd <- gsDesign::gsDesign(
    k = k,
    test.type = 1,
    alpha = alpha_vec[i],
    sfu = gsDesign::sfLDOF,
    timing = timing
  )
  p_bound <- 1 - stats::pnorm(gd$upper$bound)
  p_col <- if (i == 1) p_obs$H1 else p_obs$H2

  ct <- data.frame(
    H1 = rep(NA_character_, k),
    H2 = rep(NA_character_, k),
    Analysis = paste("Analysis", p_obs$Analysis),
    stringsAsFactors = FALSE
  )
  ct[[if (i == 1) "H1" else "H2"]] <- ifelse(p_col <= p_bound, "Success", "Fail")

  rejected <- c(FALSE, FALSE)
  rejected[i] <- any(p_col <= p_bound)

  list(bounds = NULL, closed_test = ct, rejected = rejected)
}

# OS inputs (H1/H2)
n_os_sub <- ossub$n.I
n_os_all <- os$n.I
p_obs_os <- tibble::tibble(
  Analysis = 1:3,
  H1 = dplyr::filter(inputResults, H == 1) %>% dplyr::arrange(Analysis) %>% dplyr::pull(nominalP),
  H2 = dplyr::filter(inputResults, H == 2) %>% dplyr::arrange(Analysis) %>% dplyr::pull(nominalP)
)
corr_os <- make_corr_nested_2hyp(n_os_sub, n_os_all)
info_os <- list(n_os_sub / max(n_os_sub), n_os_all / max(n_os_all))

# PFS inputs (H3/H4) encoded as local (H1/H2) for wpgsd
n_pfs_sub <- pfssub$n.I
n_pfs_all <- pfs$n.I
p_obs_pfs <- tibble::tibble(
  Analysis = 1:2,
  H1 = dplyr::filter(inputResults, H == 3) %>% dplyr::arrange(Analysis) %>% dplyr::pull(nominalP),
  H2 = dplyr::filter(inputResults, H == 4) %>% dplyr::arrange(Analysis) %>% dplyr::pull(nominalP)
)
corr_pfs <- make_corr_nested_2hyp(n_pfs_sub, n_pfs_all)
info_pfs <- list(n_pfs_sub / max(n_pfs_sub), n_pfs_all / max(n_pfs_all))

# ORR inputs (H5/H6): fixed tests at the current allocation
p_orr <- c(
  dplyr::filter(inputResults, H == 5) %>% dplyr::pull(nominalP),
  dplyr::filter(inputResults, H == 6) %>% dplyr::pull(nominalP)
)
```

### Correlation matrices for test statistics

The WPGSD methods use the correlation structure of test statistics across analyses. Below we show the correlation matrices implied by the event-overlap tables for the OS and PFS families. For convenience, we also show a combined correlation matrix with OS and PFS blocks (and 0 correlation between OS and PFS blocks).

```{r,message=FALSE,warning=FALSE,echo=FALSE}
fmt_corr <- function(x, title) {
  xt <- as.data.frame(round(x, 2))
  xt$Test <- if (!is.null(rownames(xt))) rownames(xt) else as.character(seq_len(nrow(xt)))
  xt <- xt %>% dplyr::select(.data$Test, dplyr::everything())
  gt::gt(xt) %>%
    gt::tab_header(title = title) %>%
    gt::tab_options(
      table.width = gt::pct(100),
      table.font.size = gt::px(10),
      data_row.padding = gt::px(2),
      heading.title.font.size = gt::px(12)
    ) %>%
    gt::sub_missing(missing_text = "")
}

# Make consistent test-statistic names when wpgsd does not set dimnames
make_names_2hyp <- function(h1, h2, k) {
  as.vector(rbind(paste0(h1, "_A", seq_len(k)), paste0(h2, "_A", seq_len(k))))
}

k_os <- length(n_os_sub)
k_pfs <- length(n_pfs_sub)

nm_os <- make_names_2hyp("H1", "H2", k_os)
nm_pfs <- make_names_2hyp("H3", "H4", k_pfs)

# Use named copies for display (do not modify the matrices used for computation)
corr_os_disp <- corr_os
rownames(corr_os_disp) <- nm_os
colnames(corr_os_disp) <- nm_os

corr_pfs_disp <- corr_pfs
rownames(corr_pfs_disp) <- nm_pfs
colnames(corr_pfs_disp) <- nm_pfs

# Combined block-diagonal correlation matrix (OS + PFS)
corr_all <- matrix(
  0,
  nrow = nrow(corr_os_disp) + nrow(corr_pfs_disp),
  ncol = ncol(corr_os_disp) + ncol(corr_pfs_disp)
)
corr_all[seq_len(nrow(corr_os_disp)), seq_len(ncol(corr_os_disp))] <- corr_os_disp
idx_pfs <- (nrow(corr_os_disp) + 1):(nrow(corr_os_disp) + nrow(corr_pfs_disp))
jdx_pfs <- (ncol(corr_os_disp) + 1):(ncol(corr_os_disp) + ncol(corr_pfs_disp))
corr_all[idx_pfs, jdx_pfs] <- corr_pfs_disp

rn_all <- c(paste0("OS_", nm_os), paste0("PFS_", nm_pfs))
colnames(corr_all) <- rn_all
rownames(corr_all) <- rn_all

# Display-only: show off-block OS–PFS correlations as missing/blank
corr_all_disp <- corr_all
corr_all_disp[seq_len(nrow(corr_os_disp)), jdx_pfs] <- NA_real_
corr_all_disp[idx_pfs, seq_len(ncol(corr_os_disp))] <- NA_real_
invisible(NULL)
```

```{r,message=FALSE,warning=FALSE,echo=FALSE}
fmt_corr(corr_os_disp, "OS family: correlation matrix for Z statistics (H1/H2 across analyses)")
```

```{r,message=FALSE,warning=FALSE,echo=FALSE}
fmt_corr(corr_pfs_disp, "PFS family: correlation matrix for Z statistics (H3/H4 across analyses; encoded as local H1/H2)")
```

```{r,message=FALSE,warning=FALSE,echo=FALSE}
fmt_corr(corr_all_disp, "Combined correlation matrix (block-diagonal OS + PFS)")
```

### Iterative testing and graph updates

```{r,message=FALSE,warning=FALSE}
max_iter <- 20

g_cur <- g0
graphs_iter <- list(g_cur)

rejected_global <- rep(FALSE, nHypotheses)
iter_log <- NULL

for (iter in 1:max_iter) {
  alpha_cur <- fwer * as.numeric(g_cur$hypotheses)

  # OS: alpha allocated to H1/H2 in the current overall graph
  os_res <- run_family_nested_2hyp(alpha_cur[1:2], p_obs = p_obs_os, corr = corr_os, info_frac = info_os)
  rej_os <- os_res$rejected

  # PFS: alpha allocated to H3/H4 in the current overall graph
  pfs_res <- run_family_nested_2hyp(alpha_cur[3:4], p_obs = p_obs_pfs, corr = corr_pfs, info_frac = info_pfs)
  rej_pfs <- pfs_res$rejected

  # ORR: fixed tests under current allocation
  rej_orr <- (p_orr <= alpha_cur[5:6]) & (alpha_cur[5:6] > 0)

  newly_rejected <- which(c(rej_os, rej_pfs, rej_orr) & !rejected_global)

  iter_log <- dplyr::bind_rows(
    iter_log,
    tibble::tibble(
      iteration = iter,
      alpha_H1 = alpha_cur[1],
      alpha_H2 = alpha_cur[2],
      alpha_H3 = alpha_cur[3],
      alpha_H4 = alpha_cur[4],
      alpha_H5 = alpha_cur[5],
      alpha_H6 = alpha_cur[6],
      newly_rejected = if (length(newly_rejected) == 0) "" else paste(newly_rejected, collapse = ",")
    )
  )

  if (length(newly_rejected) == 0) break

  rejected_global[newly_rejected] <- TRUE
  gu <- graphicalMCP::graph_update(g_cur, delete = newly_rejected)
  g_cur <- gu$updated_graph
  graphs_iter[[length(graphs_iter) + 1]] <- g_cur
}

iter_log %>%
  dplyr::mutate(
    across(starts_with("alpha_"), ~ round(.x, 6))
  ) %>%
  gt::gt() %>%
  gt::tab_header(title = "Iteration log: current alpha allocation and new rejections")
```

### Closed testing results under the final allocation

```{r,message=FALSE,warning=FALSE}
# Recompute under the final allocation from the last iteration
alpha_final <- fwer * as.numeric(g_cur$hypotheses)

os_final <- run_family_nested_2hyp(alpha_final[1:2], p_obs = p_obs_os, corr = corr_os, info_frac = info_os)
pfs_final <- run_family_nested_2hyp(alpha_final[3:4], p_obs = p_obs_pfs, corr = corr_pfs, info_frac = info_pfs)

if (!is.null(os_final$closed_test)) {
  os_final$closed_test %>%
    gt::gt() %>%
    gt::tab_header(title = "OS (H1/H2): closed testing results (wpgsd)")
}

if (!is.null(pfs_final$closed_test)) {
  pfs_final$closed_test %>%
    gt::gt() %>%
    gt::tab_header(title = "PFS (H3/H4): closed testing results (wpgsd; encoded as local H1/H2)")
}
```

### Fully adjusted p-values for each hypothesis

Here we compute fully adjusted p-values as the smallest familywise error rate (overall \(\alpha\)) at which a hypothesis would be rejected under the full iterative procedure described above. This can be computed regardless of whether a hypothesis is rejected under the current allocation.

```{r,message=FALSE,warning=FALSE}
iterate_rejections <- function(fwer_val) {
  g_cur0 <- g0
  rejected <- rep(FALSE, nHypotheses)

  for (iter in 1:20) {
    alpha_cur <- fwer_val * as.numeric(g_cur0$hypotheses)

    os_res <- run_family_nested_2hyp(alpha_cur[1:2], p_obs = p_obs_os, corr = corr_os, info_frac = info_os)
    pfs_res <- run_family_nested_2hyp(alpha_cur[3:4], p_obs = p_obs_pfs, corr = corr_pfs, info_frac = info_pfs)
    rej_orr <- (p_orr <= alpha_cur[5:6]) & (alpha_cur[5:6] > 0)

    newly <- which(c(os_res$rejected, pfs_res$rejected, rej_orr) & !rejected)
    if (length(newly) == 0) break

    rejected[newly] <- TRUE
    g_cur0 <- graphicalMCP::graph_update(g_cur0, delete = newly)$updated_graph
  }

  rejected
}

# Cache to avoid re-running the full procedure repeatedly
rej_cache <- new.env(parent = emptyenv())
get_rej <- function(fwer_val) {
  key <- sprintf("%.12f", fwer_val)
  if (exists(key, envir = rej_cache, inherits = FALSE)) {
    return(get(key, envir = rej_cache, inherits = FALSE))
  }
  val <- tryCatch(
    iterate_rejections(fwer_val),
    error = function(e) rep(FALSE, nHypotheses)
  )
  assign(key, val, envir = rej_cache)
  val
}

# Binary search for the minimal overall alpha leading to rejection
find_adj_p <- function(h, tol_iter = 15) {
  # Search upward from the current fwer, avoiding extreme alpha values
  hi <- fwer
  if (hi <= 0) hi <- 0.001
  while (hi < 0.5 && !get_rej(hi)[h]) hi <- hi * 2
  if (!get_rej(hi)[h]) return(NA_real_)
  lo <- 0
  for (i in seq_len(tol_iter)) {
    mid <- (lo + hi) / 2
    if (get_rej(mid)[h]) {
      hi <- mid
    } else {
      lo <- mid
    }
  }
  hi
}

adj_table <- tibble::tibble(
  Hypothesis = nameHypotheses,
  adjusted_p = vapply(seq_len(nHypotheses), find_adj_p, numeric(1)),
  rejected_at_fwer = rejected_global
)

# Add nominal p-value + analysis (using first success if rejected, else last analysis)
analysis_os_h1 <- if (!is.null(os_final$closed_test) && any(os_final$closed_test$H1 == "Success", na.rm = TRUE)) {
  min(which(os_final$closed_test$H1 == "Success"))
} else {
  nrow(p_obs_os)
}
analysis_os_h2 <- if (!is.null(os_final$closed_test) && any(os_final$closed_test$H2 == "Success", na.rm = TRUE)) {
  min(which(os_final$closed_test$H2 == "Success"))
} else {
  nrow(p_obs_os)
}
analysis_pfs_h3 <- if (!is.null(pfs_final$closed_test) && any(pfs_final$closed_test$H1 == "Success", na.rm = TRUE)) {
  min(which(pfs_final$closed_test$H1 == "Success"))
} else {
  nrow(p_obs_pfs)
}
analysis_pfs_h4 <- if (!is.null(pfs_final$closed_test) && any(pfs_final$closed_test$H2 == "Success", na.rm = TRUE)) {
  min(which(pfs_final$closed_test$H2 == "Success"))
} else {
  nrow(p_obs_pfs)
}

nom_p <- c(
  p_obs_os$H1[analysis_os_h1],
  p_obs_os$H2[analysis_os_h2],
  p_obs_pfs$H1[analysis_pfs_h3],
  p_obs_pfs$H2[analysis_pfs_h4],
  p_orr[1],
  p_orr[2]
)

analysis_used <- c(
  analysis_os_h1,
  analysis_os_h2,
  analysis_pfs_h3,
  analysis_pfs_h4,
  1,
  1
)

# Sequential p-values as minimal within-family alpha needed for rejection
# at the specified analysis, holding within-family weights fixed.
seq_p_family_2hyp <- function(h_local, analysis, w_local, p_obs_full, corr_full, info_frac_full) {
  w_local <- as.numeric(w_local)
  w_local <- pmax(w_local, 1e-12)
  w_local <- w_local / sum(w_local)

  # Restrict to the analyses up to `analysis`
  p_obs <- p_obs_full %>% dplyr::filter(analysis <= !!analysis)
  corr <- corr_full[seq_len(2 * analysis), seq_len(2 * analysis), drop = FALSE]
  info_frac <- list(
    info_frac_full[[1]][seq_len(analysis)],
    info_frac_full[[2]][seq_len(analysis)]
  )

  rejects_at <- function(alpha_total) {
    res <- tryCatch(
      run_wpgsd_2hyp(
        alpha_total = alpha_total,
        w_local = w_local,
        p_obs = p_obs,
        corr = corr,
        info_frac = info_frac
      ),
      error = function(e) NULL
    )
    if (is.null(res)) return(FALSE)
    isTRUE(res$rejected[h_local])
  }

  # If we can't reject even with alpha = 1, report 1
  if (!rejects_at(1)) return(1)

  lo <- 0
  hi <- 1
  for (i in 1:20) {
    mid <- (lo + hi) / 2
    if (rejects_at(mid)) {
      hi <- mid
    } else {
      lo <- mid
    }
  }
  hi
}

w_os_local <- alpha_final[1:2]
w_os_local <- if (sum(w_os_local) > 0) w_os_local / sum(w_os_local) else c(0.5, 0.5)
p_obs_os_w <- p_obs_os %>% dplyr::transmute(analysis = Analysis, H1 = H1, H2 = H2)

w_pfs_local <- alpha_final[3:4]
w_pfs_local <- if (sum(w_pfs_local) > 0) w_pfs_local / sum(w_pfs_local) else c(0.5, 0.5)
p_obs_pfs_w <- p_obs_pfs %>% dplyr::transmute(analysis = Analysis, H1 = H1, H2 = H2)

seq_p <- c(
  seq_p_family_2hyp(1, analysis_os_h1, w_os_local, p_obs_os_w, corr_os, info_os),
  seq_p_family_2hyp(2, analysis_os_h2, w_os_local, p_obs_os_w, corr_os, info_os),
  seq_p_family_2hyp(1, analysis_pfs_h3, w_pfs_local, p_obs_pfs_w, corr_pfs, info_pfs),
  seq_p_family_2hyp(2, analysis_pfs_h4, w_pfs_local, p_obs_pfs_w, corr_pfs, info_pfs),
  p_orr[1],
  p_orr[2]
)

adj_table2 <- adj_table %>%
  dplyr::mutate(
    Analysis = analysis_used,
    nominal_p = nom_p,
    sequential_p = seq_p
  )

adj_table2 %>%
  gt::gt() %>%
  gt::fmt_number(columns = c("nominal_p", "sequential_p", "adjusted_p"), decimals = 6) %>%
  gt::tab_header(title = "Fully adjusted p-values (minimum familywise error rate for rejection)")
```

### Compare to graphicalMCP adjusted p-values and weighted Bonferroni

The fully adjusted p-values above are based on the full iterative procedure (graph updates + WPGSD family testing). For comparison, we can also:

- apply the graphicalMCP shortcut procedure to the sequential p-values, and
- compute a weighted Bonferroni-style adjusted p-value based on the maximum \(\alpha\) allocated to each hypothesis during the shortcut rejection sequence.

```{r,message=FALSE,warning=FALSE}
# Shortcut graphicalMCP procedure based on sequential p-values
res_short <- graphicalMCP::graph_test_shortcut(
  graph = g0,
  p = seq_p,
  alpha = fwer,
  verbose = FALSE
)

adj_p_short <- res_short$outputs$adjusted_p
rej_short <- res_short$outputs$rejected

if (is.numeric(rej_short)) {
  rej_short_logical <- rep(FALSE, nHypotheses)
  rej_short_logical[rej_short] <- TRUE
} else {
  rej_short_logical <- as.logical(rej_short)
}

# Track the maximum alpha allocated to each hypothesis in the shortcut rejection sequence
rejected_order_short <- which(rej_short_logical)[order(adj_p_short[rej_short_logical])]

graphs_short <- list(g0)
if (length(rejected_order_short) > 0) {
  gu_short <- graphicalMCP::graph_update(g0, delete = rejected_order_short)
  graphs_short <- gu_short$intermediate_graphs
}
ng_short <- length(graphs_short)

lastWeights_short <- as.numeric(graphs_short[[ng_short]]$hypotheses)
lastGraph_short <- rep(ng_short, nHypotheses)

if (length(rejected_order_short) > 0) {
  for (j in seq_along(rejected_order_short)) {
    h <- rejected_order_short[j]
    lastWeights_short[h] <- as.numeric(graphs_short[[j]]$hypotheses[h])
    lastGraph_short[h] <- j
  }
}

lastAlpha_short <- fwer * lastWeights_short
adj_p_wbonf <- ifelse(lastWeights_short > 0, pmin(1, seq_p / lastWeights_short), NA_real_)
rej_wbonf <- (seq_p <= lastAlpha_short) & (lastWeights_short > 0)

comp_tab <- tibble::tibble(
  Hypothesis = nameHypotheses,
  sequential_p = seq_p,
  rejected_full_iter = rejected_global,
  adjusted_p_full_iter = adj_table$adjusted_p,
  rejected_shortcut = rej_short_logical,
  adjusted_p_shortcut = adj_p_short,
  max_alpha_alloc_shortcut = lastAlpha_short,
  adjusted_p_weighted_bonf = adj_p_wbonf,
  rejected_weighted_bonf = rej_wbonf,
  last_graph_shortcut = lastGraph_short
)

comp_tab %>%
  gt::gt() %>%
  gt::fmt_number(
    columns = c(
      "sequential_p",
      "adjusted_p_full_iter",
      "adjusted_p_shortcut",
      "max_alpha_alloc_shortcut",
      "adjusted_p_weighted_bonf"
    ),
    decimals = 6
  ) %>%
  gt::tab_header(title = "Comparison: full iterative vs shortcut graphicalMCP vs weighted Bonferroni")
```

### Overall multiplicity graph by iteration {.tabset}

```{r,message=FALSE,warning=FALSE,results='asis', fig.asp=0.9}
for (i in seq_along(graphs_iter)) {
  cat("####", paste("Iteration", i - 1, "\n\n"))
  plot(graphs_iter[[i]],
    layout = layout6,
    vertex.size = 45,
    vertex.color = cbPalette[c(2, 2, 4, 4, 3, 3)],
    vertex.label.cex = 0.9
  )
  cat("\n\n")
}
```

```{r}
# Restore default options
options(old)
```

## Session information

You can use `sessionInfo()` to document the versions of R and R packages used to render this document.
Note, in particular, that version 3.1 or later of the gsDesign package is needed.

## References

```{r, echo=FALSE, results='asis'}
cat("<script>\n")
cat("// Fix dynamic tabsets code block rendering issue under pkgdown\n")
cat("(function () {\n")
cat("  // Fix the tweaked element class\n")
cat("  unstyledCode = document.querySelectorAll(\"button code\");\n\n")
cat("  // Quit if there are no elements matched (when not rendered by pkgdown)\n")
cat("  if (!unstyledCode.length) {\n")
cat("    return;\n")
cat("  }\n\n")
cat("  for (let i = 0; i < unstyledCode.length; i++) {\n")
cat("    blockOld = unstyledCode[i].outerHTML;\n")
cat("    blockNew =\n")
cat("      \"<pre class='downlit sourceCode r' style='text-align:left;white-space:pre-wrap;'>\" +\n")
cat("      blockOld +\n")
cat("      \"</pre>\";\n")
cat("    unstyledCode[i].outerHTML = blockNew;\n")
cat("  }\n\n")
cat("  // Activate the first real tab\n")
cat("  var tabElements = [\n")
cat("    \"ul#group-sequential-designs-for-each-hypothesis li:nth-child(2) button\",\n")
cat("    \"ul#overall-multiplicity-graph-by-iteration li:nth-child(2) button\",\n")
cat("  ];\n\n")
cat("  for (let i = 0; i < tabElements.length; i++) {\n")
cat("    var el = document.querySelector(tabElements[i]);\n")
cat("    var tab = new bootstrap.Tab(el);\n")
cat("    tab.show();\n")
cat("  }\n")
cat("})();\n")
cat("</script>\n")
```
