---
title: "Group sequential design using the graphicalMCP package using hgraph"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
bibliography: gMCPLite.bib
vignette: >
  %\VignetteIndexEntry{Group sequential design using the graphicalMCP package using hgraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1,
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)
```

## Overview

This document is intended to evaluate statistical significance for graphical multiplicity control when used with group sequential design [@MaurerBretz2013]. In particular, we demonstrate design and analysis of a complex oncology trial.
There are many details building on the necessarily simple example provided by @MaurerBretz2013.
The combination of tools provided by the gMCPLite and gsDesign packages is non-trivial, but developed in a way that is meant to be re-used in a straightforward fashion.
This has been found to be particularly valuable to provide a prompt and verifiable conclusion in multiple trials such as @KEYNOTE048 where 14 hypotheses were evaluated using a template such as this.

This vignette is a variant of the original group sequential design vignette,
[Graphical testing for group sequential design](GraphicalMultiplicity.html).
The only substantive difference is that the section **"Evaluate hypothesis rejection using gMCPLite"** and subsequent sections have been updated to use the **graphicalMCP** package (while still using `hGraph()` for plotting).

Given the complexity involved, substantial effort has been taken to provide methods to check hypothesis testing.

- The initial testing is done by using sequential p-values [@LiuAnderson2008] which can then be plugged into standard graphical hypothesis testing R package, gMCP [@Bretz2009].
- The graphical testing produces a sequence including the original multiplicity graph, followed by updated multiplicity graphs, each with a single hypothesis rejected from the previous graph.
- The final graph, assuming not all hypotheses were rejected, provides the final Type I error available for testing each hypothesis that was not rejected.
- Updated group sequential bounds for each hypothesis at the largest alpha-level it was evaluated can be checked vs. nominal p-values at each analysis to verify the testing conclusions reached with the above methods.

The table of contents above lays out the organization of the document.
In short, we begin with 1) design specification followed by 2) results entry which includes event counts and nominal p-values for testing, 3) carrying out hypothesis testing, and 4) verification of the hypothesis testing results.

## Design

For the template example, there are 3 endpoints and 2 populations resulting in 6 hypotheses to be tested in the trial.
The endpoints are:

- Overall survival (OS)
- Progression free survival (PFS)
- Objective response rate (ORR)

The populations to be studied are:

- The overall population (All subjects)
- A biomarker-positive subgroup (BM+)

For simplicity, we design assuming the control group has an exponential time to event with a median of 12 months for OS and 5 months for PFS.
We design under a proportional hazards assumption.
ORR for the control group is assumed to be 15%.
In the BM+ subgroup, we assume hazard ratios of 0.65 for both OS and PFS and an experimental ORR of 35%.
In the BM- population, the assumed treatment effect is attenuated: hazard ratios of 0.85 for both OS and PFS and an experimental ORR of 25%.
Some of the choices here are arbitrary, but the intent is to fully specify how patients will be enrolled and followed for $\alpha$-controlled study analyses.

The following design characteristics are also specified to well-characterize outcomes for all subjects by the end of the trial:

- Enrollment is assumed to occur over 18 months. Enrollment will continue until the targeted number of subjects has been enrolled in the BM+ subgroup to ensure power as planned for that population. This means, the overall population sample size will be random and power may vary from that planned here.
The enrollment increases from 25% in the first two months to 50% in the second 2 months to 75% in the third two months of the final steady state enrollment which continues from the end of month 6 of enrollment until final enrollment expected at 18 months.
- The first interim analysis will be conducted 6 months after final patient enrolled to adequately assess ORR for all patients. Thus, the analysis is planned at 24 months after start of study enrollment, but will be adapted according to when final enrollment is completed.
This is the only analysis for ORR and is an interim analysis for PFS and OS with whatever event counts are available at the cutoff.
- The second interim analysis will be conducted 14 months after final enrollment to ensure minimum follow-up almost 3 times the assumed control median PFS for all subjects. This would be delayed up to 3 months if the final targeted event count for PFS in the BM+ subgroup is not achieved at that time. This is to ensure a complete description of tail behavior for PFS in the case a PFS curve has a plateau. PFS and OS will be analyzed. The endpoint counts for OS and for the overall population for PFS are random since the cutoff is determined by the PFS endpoint count for the BM+ subgroup.
- The final analysis will be performed 24 months after final enrollment, ensuring 2 times the assumed median control survival as minimum follow-up for all subjects. Only analysis of OS is planned. The final analysis may be delayed up to 6 months if the targeted OS event count in the BM+ subgroup is not achieved. Thus, the planned total duration of the trial for the OS endpoint is 42 months.

The sample size for the trial will be driven by an adequate sample size and targeted events in the BM+ subgroup to ensure 90% power for the OS endpoint assuming a hazard ratio of 0.65.
For group sequential designs, we assume 1-sided testing.

To reveal code blocks for the remainder of the document, press the code buttons indicated throughout.
The initial code block sets options and loads needed packages; no modification should be required by the user.

```{r, message=FALSE, warning=FALSE}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
# Prefer fixed notation
old <- options(scipen = 999)
# Colorblind palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# 2 packages used for data storage and manipulation: dplyr, tibble
library(dplyr)
library(tibble)
# 2 packages used for R Markdown capabilities: knitr, kableExtra
library(knitr)
library(kableExtra)
library(gt)
library(ggplot2) # For plotting
library(gsDesign) # Group sequential design capabilities (>= 3.9.0.9003 for gsSurvPower)
library(gMCPLite) # Multiplicity evaluation
library(simtrial) # Simulation of clinical trial data
library(survival) # Survival analysis (logrank test)
```

### Multiplicity diagram for hypothesis testing

Following is the multiplicity graph for the trial design.
We have arbitrarily split Type I error equally between the BM+ and overall populations.
Most $\alpha$ is allocated to OS and the least to ORR, with PFS receiving an intermediate amount.
This reflects the priority of the endpoints as well as the practicality to detect clinically significant differences in each population.
Reallocation for each endpoint proceeds from the BM+ to the overall population.
If the overall population hypothesis is rejected for a given endpoint, the reallocation is split between the two populations for another endpoint.
The choice for allocation and reallocation illustrated here is to demonstrate a complex multiplicity scenario; when actually applying this method the allocation and reallocation choices should be carefully considered.

```{r, warning=FALSE, message=FALSE}
### THIS CODE NEEDS TO BE MODIFIED FOR YOUR STUDY
# If needed, see help file for gMCPLite::hGraph() for explanation of parameters below
# Hypothesis names
nameHypotheses <- c(
  "H1: OS\n BM+",
  "H2: OS\n All subjects",
  "H3: PFS\n BM+",
  "H4: PFS\n All subjects",
  "H5: ORR\n BM+",
  "H6: ORR\n All subjects"
)
# Number of hypotheses to be tested
nHypotheses <- length(nameHypotheses)
# Transition weights for alpha reallocation (square matrix)
m <- matrix(c(
  0, 1, 0, 0, 0, 0,
  0, 0, .5, .5, 0, 0,
  0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, .5, .5,
  0, 0, 0, 0, 0, 1,
  .5, .5, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)
# Initial Type I error assigned to each hypothesis (one-sided)
alphaHypotheses <- c(.01, .01, .004, 0.000, 0.0005, .0005)
fwer <- sum(alphaHypotheses)
# Make a ggplot representation of the above specification and display it
g <- gMCPLite::hGraph(6,
  alphaHypotheses = alphaHypotheses, m = m, nameHypotheses = nameHypotheses,
  palette = cbPalette,
  halfWid = 1, halfHgt = .35, xradius = 2.5, yradius = 1, offset = 0, trhw = .15,
  x = c(-1.25, 1.25, -2.5, 2.5, -1.25, 1.25), y = c(2, 2, 1, 1, 0, 0),
  trprop = 0.4, fill = as.character(c(2, 2, 4, 4, 3, 3))
) 
print(g)
```

This testing scheme can result in what might be referred to as time travel for passing of $\alpha$.
That is, if PFS hypotheses are not rejected at a given analysis (say final PFS analysis) and OS hypotheses are rejected at the final analysis, then the previously evaluated PFS tests at the interim and final PFS analysis can be compared to updated bounds based on reallocated Type I error.
While this practice was not encouraged by @MaurerBretz2013, it was acknowledged to control Type I error as previous discussed in [@LiuAnderson2008].
Given the stringent Type I error control for multiple hypotheses, the ability to acknowledge clinically significant differences as statistically significant can be important in describing treatment benefits in regulatory labeling for a treatment.

### Group sequential designs for each hypothesis {.tabset}

For the example, we assume 1-sided testing or a non-binding futility bound as required for a @MaurerBretz2013 design using group sequential design with graphical multiplicity control.
Each is demonstrated in the example code for respective hypotheses.
Efficacy $\alpha$-spending for all group sequential designs uses the @LanDeMets spending function approximating an O'Brien-Fleming bound.

Note: `gsDesign::gsSurv()` supports multiple event-driven sample size methods, and its default is not `"Schoenfeld"`.
To align results with the `rpact`-based vignette, we set `method = "Schoenfeld"` in the designs below.

This section needs to be modified by the user to match the study design under consideration.
Those uncomfortable with coding may wish to design using the
[gsDesign Shiny app](https://gsdesign.shinyapps.io/prod/) which provides not only a point and click interface, but also a code tab that generates R code that can be copied and plugged in for designs below.

#### H1: OS, BM+

We assume 50% of the population is biomarker-positive (BM+).
The sample size is driven by overall survival (OS) in the BM+ population where we assume a hazard ratio of 0.65, targeting 90% power.
Here we assume a one-sided group sequential design with no futility bound.

```{r, results='asis'}
osmedian <- 12 # Median control survival
# Derive group sequential design for OS in the BM+ population
ossub <- gsDesign::gsSurv(
  k = 3, # 3 analyses for OS
  test.type = 1, # Efficacy bound only (no futility)
  alpha = alphaHypotheses[1], # Allocated alpha from design hypothesis group
  beta = 0.1, # Type 2 error (1 - power)
  hr = 0.65, # Assumed hazard ratio for power calculation
  timing = c(0.61, 0.82), # Choose these to match targeted calendar timing of analyses
  sfu = sfLDOF, # Spending function to approximate O'Brien-Fleming bound
  lambdaC = log(2) / osmedian, # Exponential control failure rate
  eta = 0.001, # Exponential dropout rate
  gamma = c(2.5, 5, 7.5, 10), # Relative enrollment rates by time period
  R = c(2, 2, 2, 12), # Duration of time periods for rates in gamma
  T = 42, # Planned study duration for OS
  minfup = 24, # Planned minimum follow-up after end of enrollment
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(ossub)
rownames(tab) <- 1:nrow(tab)
cat(summary(ossub))
```

The above text was automatically generated and could be edited appropriately for description of the design.
Following is a summary table describing study bounds.

```{r}
# tab %>% kable(caption = "Design for OS in the BM+ population.") %>% kable_styling()
tab %>%
  gt() %>%
  tab_header(title = "Design for OS in the BM+ population") %>%
  cols_align(align = "left", columns = Value) %>%
  tab_footnote(
    footnote = "Cumulative boundary crossing probability includes crossing probability at earlier analysis.",
    locations = cells_body(columns = "Value", rows = c(9, 10, 14, 15))
  ) %>%
  tab_footnote(
    footnote = "Approximate hazard ratio at bound.",
    locations = cells_body(columns = "Value", rows = c(3, 8, 13))
  )
```

The BM+ enrollment rate from H1 determines all other designs.
The BM- enrollment rate is derived from the BM+ prevalence.
We also define design parameters for the overall population hypotheses.
For the overall population, we use geometric mean hazard ratios across BM+ and BM- strata.

```{r}
### Derived design parameters from H1
prevalence_bm <- 0.50
pfsmedian <- 5 # Median control PFS

# BM+ sample size (both arms) from H1
n_bm <- ceiling(sum(ossub$gamma * ossub$R))
# Overall sample size driven by prevalence
n_total <- ceiling(n_bm / prevalence_bm)

# BM- hazard ratios (attenuated treatment effect)
hr_os_bm_neg <- 0.85
hr_pfs_bm_neg <- 0.85

# Geometric mean HR for overall population designs
hr_os_avg <- sqrt(0.65 * hr_os_bm_neg)   # ~0.743
hr_pfs_avg <- sqrt(0.65 * hr_pfs_bm_neg) # ~0.743

# ORR assumptions
orr_ctrl <- 0.15
orr_exp_bm_pos <- 0.35
orr_exp_bm_neg <- 0.25
orr_exp_avg <- prevalence_bm * orr_exp_bm_pos + (1 - prevalence_bm) * orr_exp_bm_neg # 0.30

# Stratified enrollment rate matrix for overall population designs
# (2-column: BM+ / BM-)
gamma_strat <- cbind(
  ossub$gamma,                                                  # BM+ rates
  ossub$gamma * (1 - prevalence_bm) / prevalence_bm            # BM- rates
)

# Stratified control-arm hazard rates
lambdaC_os_strat <- matrix(log(2) / osmedian, nrow = 1, ncol = 2,
                           dimnames = list("0-Inf", c("Stratum 1", "Stratum 2")))
lambdaC_pfs_strat <- matrix(log(2) / pfsmedian, nrow = 1, ncol = 2,
                            dimnames = list("0-Inf", c("Stratum 1", "Stratum 2")))

# Stratified dropout rates
eta_os_strat <- matrix(0.001, nrow = 1, ncol = 2,
                       dimnames = list("0-Inf", c("Stratum 1", "Stratum 2")))
eta_pfs_strat <- matrix(0.02, nrow = 1, ncol = 2,
                        dimnames = list("0-Inf", c("Stratum 1", "Stratum 2")))

cat("BM+ N:", n_bm, "\nTotal N:", n_total, "\n")
cat("Geometric mean HR (OS):", round(hr_os_avg, 4), "\n")
cat("Geometric mean HR (PFS):", round(hr_pfs_avg, 4), "\n")
```

#### H2: OS, All

The total sample size is `r n_total`, approximately twice the BM+ sample size.
For the overall population, we use `gsSurvPower()` to compute power with BM+-derived enrollment and the geometric mean hazard ratio across strata.
For this design, we consider a non-binding futility bound where the trial may be stopped early in the overall population if the bound is crossed.
We use a @HwangShihDeCani bound with $\gamma = -3.25$.
Study designers should carefully consider implication for parameter choices, particularly if the futility bounds provide sensible guidance for stopping the trial.
Since the futility bounds are non-binding, the efficacy bound is computed assuming the futility bound is ignored which will control Type I error at the targeted level even if a futility bound is crossed and the trial is continued.

```{r, results = 'asis'}
os <- gsDesign::gsSurvPower(
  x = ossub,
  test.type = 4, # Non-binding futility
  alpha = alphaHypotheses[2],
  sfu = sfLDOF,
  sfl = sfHSD, sflpar = -3.25,
  testLower = c(TRUE, TRUE, FALSE),
  lambdaC = lambdaC_os_strat, # Stratified control hazard (BM+/BM-)
  hr = hr_os_avg, # Geometric mean HR across strata
  eta = eta_os_strat, # Stratified dropout
  etaE = eta_os_strat,
  gamma = gamma_strat, # Stratified enrollment rates
  targetN = n_total, # Rescale enrollment to overall population N
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(os)
rownames(tab) <- 1:nrow(tab)
cat(summary(os))
```

```{r}
tab %>%
  kable(caption = "Design for OS in all subjects") %>%
  kable_styling()
```

We can also plot different design characteristics. Here, we plot the approximate hazard ratio to cross each bound which may be helpful for design team discussions.

```{r}
plot(os, plottype = "HR", xlab = "Events")
```

#### H3: PFS, BM+

For progression free survival (PFS) we assume a shorter median time to event of 5 months.
With an assumed hazard ratio of 0.65 in the BM+ population, we use `gsSurvPower()` with enrollment rates from H1 and `targetN = n_bm`.
We assume a larger dropout rate for PFS than we did for OS.
Here we set up a futility bound for safety.
This is an asymmetric 2-sided design with both futility and efficacy boundary crossing probabilities under the null hypothesis. The parameter `astar = 0.1` specifies total lower bound spending of 10%.
The lower @HwangShihDeCani spending bound with $\gamma = -8$ is intended to be conservative in terms of futility at the interim, but still provide a safety bound for the PFS result in this targeted population.
This would have to be carefully evaluated by the study design team at the time of design.

```{r, results='asis'}
pfssub <- gsDesign::gsSurvPower(
  x = ossub,
  k = 2, test.type = 6, astar = 0.1,
  alpha = alphaHypotheses[3],
  sfu = sfLDOF,
  sfl = sfHSD, sflpar = -8,
  testUpper = c(TRUE, TRUE),
  testLower = c(TRUE, FALSE),
  lambdaC = log(2) / pfsmedian, # BM+ control PFS hazard
  hr = 0.65,
  eta = 0.02,
  gamma = ossub$gamma,
  targetN = n_bm,
  plannedCalendarTime = c(32, 36),
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(pfssub)
rownames(tab) <- 1:nrow(tab)
cat(summary(pfssub))
```

```{r}
tab %>%
  kable(caption = "Design for PFS in the BM+ population") %>%
  kable_styling()
```

#### H4: PFS, All

Finally, we design for PFS in all subjects using `gsSurvPower()` with stratified control hazard rates and the geometric mean hazard ratio.
In this case, we simplify to a one-sided design.
A futility bound could be considered, if appropriate.

```{r}
pfs <- gsDesign::gsSurvPower(
  x = ossub,
  k = 2, test.type = 1,
  alpha = alphaHypotheses[3], # Use same alpha as H3 (H4 starts with 0 but receives alpha via reallocation)
  sfu = sfLDOF,
  lambdaC = lambdaC_pfs_strat, # Stratified control PFS hazard (BM+/BM-)
  hr = hr_pfs_avg, # Geometric mean HR across strata
  eta = eta_pfs_strat, # Stratified dropout
  etaE = eta_pfs_strat,
  gamma = gamma_strat, # Stratified enrollment rates
  targetN = n_total,
  plannedCalendarTime = c(32, 36),
  method = "Schoenfeld"
)
tab <- gsDesign::gsBoundSummary(pfs)
rownames(tab) <- 1:nrow(tab)
tab %>%
  kable(caption = "Design for PFS in the overall population") %>%
  kable_styling()
```

#### H5 and H6: ORR

For objective response rate (ORR), we assume an underlying control rate of `r orr_ctrl * 100`%.
In the BM+ population, we compute power to detect a 20% improvement using the H1-derived BM+ sample size.

```{r}
nBinomial(p1 = orr_exp_bm_pos, p2 = orr_ctrl, alpha = alphaHypotheses[5], n = n_bm)
```

In the all subjects population, we use prevalence-weighted average response rates and the overall sample size.

```{r}
nBinomial(p1 = orr_exp_avg, p2 = orr_ctrl, alpha = alphaHypotheses[6], n = n_total)
```

### Design list

Now we associate designs with hypotheses in an ordered list corresponding to the order in the multiplicity graph setup.
Since ORR designs are not group sequential, we enter `NULL` values for those in the last 2 entries of the design list; hit code button to reveal code for this.

```{r}
### THIS NEEDS TO BE MODIFIED TO MATCH STUDY
gsDlist <- list(ossub, os, pfssub, pfs, NULL, NULL)
```

### Spending plan and spending time

While it was relatively straightforward above to set up timing of analyses to match for the different hypotheses, accumulation of endpoints can vary from plan in a variety of ways.
Planning on how to deal with this is critical at the time of protocol development to avoid later amendments or inappropriate $\alpha$-allocation to early analyses.
Before going into examples, we review the concept of $\alpha$-spending and what we will refer to as _spending time_.

For a given hypothesis, we will assign a non-decreasing spending function $f(t)$ defined for $t\ge 0$ with $f(0)=0$ and $f(t)=\alpha$ for $t\ge 1$.
We will assume $K$ analyses with observed event counts $n_k$ at analysis $k=1,2,\ldots,K$ and a targeted final event count of $N_k$.
The $\alpha$-spending at analysis $k$ was originally defined [@LanDeMets] as $f(t_k=n_k/N_K)$.
The values $n_k/N_K$ will be referred to as the _information fraction_, $k=1,\ldots,K$.
This is used to pre-specify the cumulative amount of Type I error for a hypothesis at each analysis.
In @LanDeMets1989 they noted that calendar time was another option for $t_k$ values, $k=1,\ldots,K.$
@PLWBook noted further that as long as $t_k$ is increasing with $k$, it can be used to define spending; this is subject to the requirement that under the null hypothesis, the timing must be selected in a way that is not correlated with the test statistic (e.g., blinded).
We will refer to $t_k$, regardless of its definition, as the _spending time_ for a hypothesis.
Note that the joint distribution of interim and final tests for a hypothesis is driven by $n_k$, $k=1,\ldots,K$.
This is equivalent to basing correlation on the information fraction $n_k^{(actual)}/n_K^{(planned)}$, $1\le k\le K$.
Thus, both spending time and information fraction are required to compute bounds for group sequential testing.
Our general objectives here will be to:

- Spend all Type I error for each hypothesis in its combined interim and final analyses; this requires the spending time to be 1 for the final analysis of a hypothesis.
- Ensure spending time is well defined for each analysis of each hypothesis.
- We will assume that both follow-up duration and event counts may be of interest in determining timing of analyses; e.g., for immuno-oncology therapies there have been delayed treatment effects and the tail of the time-to-event distribution has been important to establish benefit. Thus, we will assume here that over-spending at interim analysis is to be avoided.

## Simulated trial data

We simulate a single trial instance using an illness-death model with response.
The model has four states: alive without response or progression (0),
responded without progression (1), progressed (2), and dead (3).

```{r, message=FALSE}
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "sim_illness_death.R"))
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "cut_illness_death.R"))
```

```{r}
transition_rate <- build_transition_rates(
  strata = c("BM+", "BM-"),
  treatments = c("control", "experimental"),
  median_pfs = c("BM+" = pfsmedian, "BM-" = pfsmedian),
  median_os = c("BM+" = osmedian, "BM-" = osmedian),
  orr = list(
    "BM+" = c(control = 0.15, experimental = 0.35),
    "BM-" = c(control = 0.15, experimental = 0.25)
  ),
  hr_pfs = c("BM+" = 0.65, "BM-" = 0.85),
  hr_os = c("BM+" = 0.65, "BM-" = 0.85)
)
```

### Simulate one trial

```{r}
set.seed(20250328)
steady_rate <- n_total / 15  # 18-month enrollment: 0.25*2 + 0.5*2 + 0.75*2 + 1*12 = 15 rate-months

sim_data <- sim_illness_death(
  n = n_total,
  stratum = data.frame(stratum = c("BM+", "BM-"),
                       p = c(prevalence_bm, 1 - prevalence_bm)),
  block = c("control", "control", "experimental", "experimental"),
  enroll_rate = data.frame(
    rate = steady_rate * c(0.25, 0.50, 0.75, 1.00),
    duration = c(2, 2, 2, 12)
  ),
  transition_rate = transition_rate,
  dropout_rate = 0.001
)

fpe <- max(sim_data$ENRLTIME)
cat("Total N:", nrow(sim_data), "\n")
cat("N BM+:", sum(sim_data$STRATUM == "BM+"),
    "N BM-:", sum(sim_data$STRATUM == "BM-"), "\n")
cat("FPE:", round(fpe, 1), "months\n")
```

### Determine analysis cut dates and compute p-values

```{r}
target_pfs_events_sub <- ceiling(max(pfssub$n.I))
target_os_events_sub <- ceiling(max(ossub$n.I))

analyses <- list(
  list(min_followup = 6, endpoint = NULL, event_target = NULL,
       target_stratum = NULL, max_followup = NULL),
  list(min_followup = 14, endpoint = "PFS", event_target = target_pfs_events_sub,
       target_stratum = "BM+", max_followup = 17),
  list(min_followup = 24, endpoint = "OS", event_target = target_os_events_sub,
       target_stratum = "BM+", max_followup = 30)
)

cut_dates <- get_analysis_dates(sim_data, analyses)

cat("Analysis cut dates:\n")
for (i in seq_along(cut_dates)) {
  cat(sprintf("  Analysis %d: %.1f months (FPE + %.1f)\n",
              i, cut_dates[i], cut_dates[i] - fpe))
}
```

```{r}
adtte <- lapply(seq_along(cut_dates), function(i) {
  d <- cut_illness_death(sim_data, cut_dates[i])
  d$ANALYSIS <- i
  d
})

# Helper: 1-sided logrank p-value
logrank_pval <- function(data, stratified = FALSE) {
  data$event <- 1L - data$CNSR
  if (stratified) {
    fit <- survdiff(Surv(AVAL, event) ~ TRT + strata(STRATUM), data = data)
  } else {
    fit <- survdiff(Surv(AVAL, event) ~ TRT, data = data)
  }
  obs_ctrl <- if (is.matrix(fit$obs)) sum(fit$obs[1, ]) else fit$obs[1]
  exp_ctrl <- if (is.matrix(fit$exp)) sum(fit$exp[1, ]) else fit$exp[1]
  z <- sign(obs_ctrl - exp_ctrl) * sqrt(fit$chisq)
  pnorm(-z)
}

# Helper: 1-sided rate difference p-value
rd_pval <- function(data, stratified = FALSE) {
  if (!stratified) {
    n_exp <- sum(data$TRT == "experimental")
    n_ctrl <- sum(data$TRT == "control")
    p_exp <- sum(data$AVAL[data$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(data$AVAL[data$TRT == "control"]) / n_ctrl
    p_pool <- (p_exp * n_exp + p_ctrl * n_ctrl) / (n_exp + n_ctrl)
    se <- sqrt(p_pool * (1 - p_pool) * (1 / n_exp + 1 / n_ctrl))
    if (se == 0) return(0.5)
    return(pnorm(-(p_exp - p_ctrl) / se))
  }
  strata <- unique(data$STRATUM)
  n_s <- rd_s <- var_s <- numeric(length(strata))
  for (k in seq_along(strata)) {
    ds <- data[data$STRATUM == strata[k], ]
    n_exp <- sum(ds$TRT == "experimental")
    n_ctrl <- sum(ds$TRT == "control")
    n_s[k] <- n_exp + n_ctrl
    p_exp <- sum(ds$AVAL[ds$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(ds$AVAL[ds$TRT == "control"]) / n_ctrl
    rd_s[k] <- p_exp - p_ctrl
    var_s[k] <- p_exp * (1 - p_exp) / n_exp + p_ctrl * (1 - p_ctrl) / n_ctrl
  }
  w <- n_s / sum(n_s)
  z <- sum(w * rd_s) / sqrt(sum(w^2 * var_s))
  pnorm(-z)
}

pvals <- list()
for (i in seq_along(cut_dates)) {
  d <- adtte[[i]]
  pvals[[paste0("H1_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS", STRATUM == "BM+"))
  pvals[[paste0("H2_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS"), stratified = TRUE)
  if (i <= 2) {
    pvals[[paste0("H3_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS", STRATUM == "BM+"))
    pvals[[paste0("H4_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS"), stratified = TRUE)
  }
  if (i == 1) {
    pvals[[paste0("H5_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR", STRATUM == "BM+"))
    pvals[[paste0("H6_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR"), stratified = TRUE)
  }
}
```

## Results entry at time of analysis

### Build results table

Event counts and p-values are assembled from the simulated trial data.
Spending time uses the BM+ information fraction for each hypothesis.

```{r,warnings=FALSE,message=FALSE}
get_events <- function(analysis_idx, paramcd, stratum = NULL) {
  d <- adtte[[analysis_idx]]
  if (!is.null(stratum)) d <- d %>% filter(STRATUM == stratum)
  sum(d$PARAMCD == paramcd & d$CNSR == 0, na.rm = TRUE)
}

events_os_sub <- sapply(1:3, function(i) get_events(i, "OS", "BM+"))
events_os_all <- sapply(1:3, function(i) get_events(i, "OS"))
events_pfs_sub <- sapply(1:2, function(i) get_events(i, "PFS", "BM+"))
events_pfs_all <- sapply(1:2, function(i) get_events(i, "PFS"))

# Update design objects with observed event counts
ossub$n.I <- events_os_sub
os$n.I <- events_os_all
pfssub$n.I <- events_pfs_sub
pfs$n.I <- events_pfs_all

inputResults <- tibble(
  H = c(rep(1, 3), rep(2, 3), rep(3, 2), rep(4, 2), 5, 6),
  Pop = c(
    rep("BM+", 3), rep("All", 3),
    rep("BM+", 2), rep("All", 2),
    "BM+", "All"
  ),
  Endpoint = c(rep("OS", 6), rep("PFS", 4), rep("ORR", 2)),
  nominalP = c(
    pvals$H1_1, pvals$H1_2, pvals$H1_3,
    pvals$H2_1, pvals$H2_2, pvals$H2_3,
    pvals$H3_1, pvals$H3_2,
    pvals$H4_1, pvals$H4_2,
    pvals$H5_1,
    pvals$H6_1
  ),
  Analysis = c(1:3, 1:3, 1:2, 1:2, 1, 1),
  events = c(ossub$n.I, os$n.I, pfssub$n.I, pfs$n.I, NA, NA),
  spendingTime = c(
    ossub$n.I / max(ossub$n.I),
    ossub$n.I / max(ossub$n.I),
    pfssub$n.I / max(pfssub$n.I),
    pfssub$n.I / max(pfssub$n.I),
    NA, NA
  )
)
kable(inputResults, caption = "Results from simulated trial data") %>%
  kable_styling()
```

## Testing hypotheses

### Compute sequential p-values for each hypothesis

Sequential p-value computation is done in one loop in an attempt to minimize chances for coding errors.
We delay showing these until after display of the sequence of multiplicity graphs generated by hypothesis rejection is shown.

```{r,message=FALSE}
### USER SHOULD NOT NEED TO MODIFY THIS CODE
EOCtab <- NULL
EOCtab <- inputResults %>%
  group_by(H) %>%
  slice(1) %>%
  ungroup() %>%
  select("H", "Pop", "Endpoint", "nominalP")
EOCtab$seqp <- .9999
for (EOCtabline in 1:nHypotheses) {
  EOCtab$seqp[EOCtabline] <-
    ifelse(is.null(gsDlist[[EOCtabline]]), EOCtab$nominalP[EOCtabline], {
      tem <- filter(inputResults, H == EOCtabline)
      sequentialPValue(
        gsD = gsDlist[[EOCtabline]], interval = c(.0001, .9999),
        n.I = tem$events,
        Z = -qnorm(tem$nominalP),
        usTime = tem$spendingTime
      )
    })
}
EOCtab <- EOCtab %>% select(-"nominalP")
# kable(EOCtab,caption="Sequential p-values as initially placed in EOCtab") %>% kable_styling()
```

### Evaluate hypothesis rejection using graphicalMCP

We need to set up an `initial_graph` object as implemented in the graphicalMCP package.

```{r,message=FALSE,warning=FALSE}
library(graphicalMCP)

g0 <- graphicalMCP::graph_create(
  hypotheses = alphaHypotheses / fwer,
  transitions = m,
  hyp_names = nameHypotheses
)
```

Now we add the sequential p-values and evaluate which hypotheses have been rejected.

```{r}
result <- graphicalMCP::graph_test_shortcut(
  graph = g0,
  p = EOCtab$seqp,
  alpha = fwer,
  verbose = TRUE
)

adj_p <- result$outputs$adjusted_p
rej <- result$outputs$rejected

if (is.numeric(rej)) {
  rej_logical <- rep(FALSE, nHypotheses)
  rej_logical[rej] <- TRUE
} else {
  rej_logical <- as.logical(rej)
}

EOCtab$Rejected <- rej_logical
EOCtab$adjPValues <- adj_p
```

## Verification of hypotheses rejected

```{r}
rejected_order <- which(EOCtab$Rejected)[order(EOCtab$adjPValues[EOCtab$Rejected])]

graphs <- list(g0)
if (length(rejected_order) > 0) {
  gu <- graphicalMCP::graph_update(g0, delete = rejected_order)
  # `intermediate_graphs` includes the initial graph as its first element
  graphs <- gu$intermediate_graphs
}
ngraphs <- length(graphs)

# Final weights for hypotheses not rejected come from the final graph
lastWeights <- as.numeric(graphs[[ngraphs]]$hypotheses)
lastGraph <- rep(ngraphs, nrow(EOCtab))

# For rejected hypotheses, take the hypothesis weight from the graph at which it was rejected
if (length(rejected_order) > 0) {
  for (j in seq_along(rejected_order)) {
    h <- rejected_order[j]
    lastWeights[h] <- as.numeric(graphs[[j]]$hypotheses[h])
    lastGraph[h] <- j
  }
}

EOCtab$lastAlpha <- fwer * lastWeights
EOCtab$lastGraph <- lastGraph
EOCtabx <- EOCtab
names(EOCtabx) <- c(
  "Hypothesis", "Population", "Endpoint", "Sequential p",
  "Rejected", "Adjusted p", "Max alpha allocated", "Last Graph"
)
# Display table with desired column order
# Delayed following until after multiplicity graph sequence
# EOCtabx %>% select(c(1:4,7,5:6,8)) %>% kable() %>% kable_styling()
```

### Multiplicity graph sequence from graphicalMCP {.tabset}

```{r,message=FALSE,warning=FALSE,results='asis'}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
for (i in 1:ngraphs) {
  mx <- graphs[[i]]$transitions
  rownames(mx) <- NULL
  colnames(mx) <- NULL
  g <- gMCPLite::hGraph(
    nHypotheses = nHypotheses,
    alphaHypotheses = graphs[[i]]$hypotheses * fwer,
    m = mx,
    nameHypotheses = nameHypotheses,
    palette = cbPalette,
    halfWid = 1, halfHgt = .35, xradius = 2.5, yradius = 1, offset = 0, trhw = .15,
    x = c(-1.25, 1.25, -2.5, 2.5, -1.25, 1.25), y = c(2, 2, 1, 1, 0, 0),
    trprop = .4, fill = as.character(c(2, 2, 4, 4, 3, 3))
  ) 
  cat(" \n")
  cat("####", paste(" Graph", as.character(i), " \n\n"))
  print(g)
  cat(" \n\n\n")
}
```

### Comparison of sequential p-values to multiplicity graphs

We can compare sequential p-values to available $\alpha$ in each graph.
In the column 'Last Graph' we can see one of 2 things:

1) For rejected hypotheses, the maximum $\alpha$ allocated to the hypothesis. For example, hypothesis one was allocated $\alpha=0.01$ in the first graph above (select using first tab). We see that the sequential p-value of 0.0001 is smaller than $\alpha=0.01$ and thus the hypothesis is rejected. We can then proceed to the second graph and see that hypothesis 5 was rejected. The last hypothesis rejected is hypothesis 3 in the third graph.
2) For the remaining hypotheses (H2, H4, H6) the maximum $\alpha$ allocated is in the fourth graph; since each sequential p-value is greater than the allocated $\alpha$ for the corresponding hypothesis, none of these hypotheses were rejected.

```{r}
EOCtabx %>%
  select(c(1:4, 7, 5:6, 8)) %>%
  kable() %>%
  kable_styling()
```

### Bounds at final $\alpha$ allocated for group sequential tests {.tabset}

As a separate validation, we examine group sequential bounds for each hypothesis updated with 1) the maximum $\alpha$ allocated above, 2) the number of events at each analysis, and 3) the cumulative spending at each analysis above. The nominal p-value for at least one of the analyses performed for each rejected hypotheses should be less than or equal to the nominal p-value in the group sequential design.
For each hypothesis not rejected, all nominal p-values are greater than the its corresponding bound.
For hypotheses tested without a group sequential design, the nominal p-value for the test of that hypothesis can be compared to the maximum alpha allocated in the above table.

```{r,results='asis'}
for (i in 1:nHypotheses) {
  # Set up tab for hypothesis in output
  cat("####", paste(" Hypothesis", as.character(i), " \n"))
  # Get results for hypothesis
  hresults <- inputResults %>% filter(H == i)
  # Print out max alpha allocated
  xx <- paste("Max alpha allocated from above table: ",
    as.character(EOCtab$lastAlpha[i]),
    sep = ""
  )
  d <- gsDlist[[i]]
  # If not group sequential for this hypothesis, print the max alpha allocated
  # and the nominal p-value
  if (is.null(d)) {
    cat("Maximum alpha allocated: ")
    cat(EOCtab$lastAlpha[i])
    cat("\n\n")
    cat("Nominal p-value for hypothesis test: ")
    cat(hresults$nominalP)
    cat("\n\n")
  }
  # For group sequential tests, print max alpha allocated and
  # corresponding group sequential bounds
  if (!is.null(gsDlist[[i]])) {
    cat("Nominal p-values at each analysis for comparison to bounds in table below:\n\n")
    cat(hresults$nominalP)
    cat("\n\n")
    # Get other info for current hypothesis
    n.I <- hresults$events
    usTime <- hresults$spendingTime
    n.Iplan <- max(d$n.I)
    if (length(n.I) == 1) {
      n.I <- c(n.I, n.Iplan)
      usTime <- c(usTime, 1)
    }
    # If no alpha allocated, just print text line to note this along with the 0 alpha allocated
    if (EOCtab$lastAlpha[i] == 0) {
      cat("Maximum alpha allocated: 0\n\n")
      cat("No testing required\n\n")
    }
    if (EOCtab$lastAlpha[i] > 0) {
      dupdate <- gsDesign::gsDesign(
        alpha = EOCtab$lastAlpha[i],
        k = length(n.I),
        n.I = n.I,
        usTime = usTime,
        maxn.IPlan = n.Iplan,
        n.fix = d$n.fix,
        test.type = 1,
        sfu = d$upper$sf,
        sfupar = d$upper$param
      )
      tabl <- gsDesign::gsBoundSummary(dupdate,
        Nname = "Events",
        exclude = c(
          "B-value", "CP", "CP H1", "Spending",
          "~delta at bound", "P(Cross) if delta=0",
          "PP", "P(Cross) if delta=1"
        )
      )
      kable(tabl, caption = xx, row.names = FALSE) %>%
        kable_styling() %>%
        cat()
      cat("\n\n")
    }
  }
}
```

```{r}
# Restore default options
options(old)
```

## Session information

You can use `sessionInfo()` to document the versions of R and R packages used to render this document.
Note, in particular, that version 3.1 or later of the gsDesign package is needed.

## References

```{js,echo=FALSE}
// Fix dynamic tabsets code block rendering issue under pkgdown
(function () {
  // Fix the tweaked element class
  unstyledCode = document.querySelectorAll("button code");

  // Quit if there are no elements matched (when not rendered by pkgdown)
  if (!unstyledCode.length) {
    return;
  }

  for (let i = 0; i < unstyledCode.length; i++) {
    blockOld = unstyledCode[i].outerHTML;
    blockNew =
      "<pre class='downlit sourceCode r' style='text-align:left;white-space:pre-wrap;'>" +
      blockOld +
      "</pre>";
    unstyledCode[i].outerHTML = blockNew;
  }

  // Activate the first real tab
  var tabElements = [
    "ul#multiplicity-graph-sequence-from-graphicalmcp li:nth-child(2) button",
    "ul#bounds-at-final-alpha-allocated-for-group-sequential-tests li:nth-child(2) button",
  ];

  for (let i = 0; i < tabElements.length; i++) {
    var el = document.querySelector(tabElements[i]);
    var tab = new bootstrap.Tab(el);
    tab.show();
  }
})();
```
