---
title: "Group sequential design using the graphicalMCP and rpact packages"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
bibliography: gMCPLite.bib
vignette: >
  %\VignetteIndexEntry{Group sequential design using the graphicalMCP and rpact packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1,
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)
```

## Overview

NOTE: THIS VIGNETTE IS NOT YET COMPLETE. IN PARTICULAR, SPENDING TIME AND SEQUENTIAL P-VALUES NEED TO BE CHECKED.

This document is intended to evaluate statistical significance for graphical multiplicity control when used with group sequential design [@MaurerBretz2013]. 
In particular, we demonstrate design and analysis of a complex oncology trial using the **rpact** package.
There are many details building on the necessarily simple example provided by @MaurerBretz2013.
The combination of tools provided by the graphicalMCP and rpact packages is non-trivial, but developed in a way that is meant to be re-used in a straightforward fashion.
Here we have to add a function to produce sequential p-values for a rpact design object.
Otherwise, the vignette is essentially the same as the original group sequential design vignettes, but implemented using **rpact** for all group sequential computations.
This has been found to be particularly valuable to provide a prompt and verifiable conclusion in multiple trials such as @KEYNOTE048 where 14 hypotheses were evaluated using a template such as this.

This vignette is a variant of the original group sequential design vignette,
[Graphical testing for group sequential design](GraphicalMultiplicity.html).
The key differences in this vignette are:

- Graph updates and multiple testing are performed using **graphicalMCP**.
- Group sequential boundaries are computed using **rpact**, and we add a small helper to compute **sequential p-values** for an rpact group sequential design.

Given the complexity involved, substantial effort has been taken to provide methods to check hypothesis testing.

- The initial testing is done by using sequential p-values [@LiuAnderson2008] which can then be plugged into standard graphical hypothesis testing R package, gMCP [@Bretz2009].
- The graphical testing produces a sequence including the original multiplicity graph, followed by updated multiplicity graphs, each with a single hypothesis rejected from the previous graph.
- The final graph, assuming not all hypotheses were rejected, provides the final Type I error available for testing each hypothesis that was not rejected.
- Updated group sequential bounds for each hypothesis at the largest alpha-level it was evaluated can be checked vs. nominal p-values at each analysis to verify the testing conclusions reached with the above methods.

The table of contents above lays out the organization of the document.
In short, we begin with 1) design specification followed by 2) results entry which includes event counts and nominal p-values for testing, 3) carrying out hypothesis testing, and 4) verification of the hypothesis testing results.

## Design

For the template example, there are 3 endpoints and 2 populations resulting in 6 hypotheses to be tested in the trial.
The endpoints are:

- Overall survival (OS)
- Progression free survival (PFS)
- Objective response rate (ORR)

The populations to be studied are:

- The overall population (All subjects)
- A subgroup (Subgroup)

For simplicity, we design assuming the control group has an exponential time to event with a median of 12 months for OS and 5 months for PFS.
We design under a proportional hazards assumption.
ORR for the control group is assumed to be 15%.
Some of the choices here are arbitrary, but the intent is to fully specify how patients will be enrolled and followed for $\alpha$-controlled study analyses.

The following design characteristics are also specified to well-characterize outcomes for all subjects by the end of the trial:

- Enrollment is assumed to occur over 18 months. Enrollment will continue until the targeted number of subjects has been enrolled in the subgroup to ensure power as planned for that population. This means, the overall population sample size will be random and power may vary from that planned here.
The enrollment increases from 25% in the first two months to 50% in the second 2 months to 75% in the third two months of the final steady state enrollment which continues from the end of month 6 of enrollment until final enrollment expected at 18 months.
- The first interim analysis will be conducted 6 months after final patient enrolled to adequately assess ORR for all patients. Thus, the analysis is planned at 24 months after start of study enrollment, but will be adapted according to when final enrollment is completed.
This is the only analysis for ORR and is an interim analysis for PFS and OS with whatever event counts are available at the cutoff.
- The second interim analysis will be conducted 14 months after final enrollment to ensure minimum follow-up almost 3 times the assumed control median PFS for all subjects. This would be delayed up to 6 months if the final targeted event count for PFS in either the subgroup or the overall population is not achieved at that time. This is to ensure a complete description of tail behavior for PFS in the case a PFS curve has a plateau. PFS and OS will be analyzed.
- The final analysis will be performed 24 months after final enrollment, ensuring 2 times the assumed median control survival as minimum follow-up for all subjects. Only analysis of OS is planned. The final analysis may be delayed up to 9 months if the targeted OS event count in either the subgroup or the overall population is not achieved. Thus, the planned total duration of the trial for the OS endpoint is 45 months.

The sample size for the trial will be driven by an adequate sample size and targeted events in the subgroup to ensure 90% power for the OS endpoint assuming a hazard ratio of 0.65.
For group sequential designs, we assume 1-sided testing.

To reveal code blocks for the remainder of the document, press the code buttons indicated throughout.
The initial code block sets options and loads needed packages; no modification should be required by the user.

```{r, message=FALSE, warning=FALSE}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
# Prefer fixed notation
old <- options(scipen = 999)
# Colorblind palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# 2 packages used for data storage and manipulation: dplyr, tibble
library(dplyr)
library(tibble)
# 2 packages used for R Markdown capabilities: knitr, kableExtra
library(knitr)
library(kableExtra)
library(gt)
library(ggplot2) # For plotting
suppressPackageStartupMessages(library(rpact)) # Group sequential design capabilities
library(graphicalMCP) # Multiplicity evaluation and graph updates
library(simtrial) # Simulation of clinical trial data
library(survival) # Survival analysis (logrank test)
```

### Multiplicity diagram for hypothesis testing

Following is the multiplicity graph for the trial design.
We have arbitrarily split Type I error equally between the subgroup and overall populations.
Most $\alpha$ is allocated to OS and the least to ORR, with PFS receiving an intermediate amount.
This reflects the priority of the endpoints as well as the practicality to detect clinically significant differences in each population.
Reallocation for each endpoint proceeds from the subgroup to the overall population.
If the overall population hypothesis is rejected for a given endpoint, the reallocation is split between the two populations for another endpoint.
The choice for allocation and reallocation illustrated here is to demonstrate a complex multiplicity scenario; when actually applying this method the allocation and reallocation choices should be carefully considered.

```{r, warning=FALSE, message=FALSE, fig.asp=0.9}
### THIS CODE NEEDS TO BE MODIFIED FOR YOUR STUDY
# Hypothesis names
nameHypotheses <- c(
  "H1: OS\n Subgroup",
  "H2: OS\n All subjects",
  "H3: PFS\n Subgroup",
  "H4: PFS\n All subjects",
  "H5: ORR\n Subgroup",
  "H6: ORR\n All subjects"
)
# Number of hypotheses to be tested
nHypotheses <- length(nameHypotheses)
# Transition weights for alpha reallocation (square matrix)
m <- matrix(c(
  0, 1, 0, 0, 0, 0,
  0, 0, .5, .5, 0, 0,
  0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, .5, .5,
  0, 0, 0, 0, 0, 1,
  .5, .5, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)
# Initial Type I error assigned to each hypothesis (one-sided)
alphaHypotheses <- c(.01, .01, .004, 0.000, 0.0005, .0005)
fwer <- sum(alphaHypotheses)

fill_idx <- c(2, 2, 4, 4, 3, 3)
vertex_color_map <- stats::setNames(cbPalette[fill_idx], nameHypotheses)

g0 <- graphicalMCP::graph_create(
  hypotheses = stats::setNames(alphaHypotheses / fwer, nameHypotheses),
  transitions = m,
  hyp_names = nameHypotheses
)

layout6 <- rbind(
  c(0.28, 0.74),
  c(0.72, 0.74),
  c(0.18, 0.50),
  c(0.82, 0.50),
  c(0.28, 0.26),
  c(0.72, 0.26)
)

plot(g0,
  layout = layout6,
  vertex.size = 45,
  vertex.label.cex = 0.9,
  vertex.color = unname(vertex_color_map[names(g0$hypotheses)]),
  margin = 0.25
)
```

This testing scheme can result in what might be referred to as time travel for passing of $\alpha$.
That is, if PFS hypotheses are not rejected at a given analysis (say final PFS analysis) and OS hypotheses are rejected at the final analysis, then the previously evaluated PFS tests at the interim and final PFS analysis can be compared to updated bounds based on reallocated Type I error.
While this practice was not encouraged by @MaurerBretz2013, it was acknowledged to control Type I error as previous discussed in [@LiuAnderson2008].
Given the stringent Type I error control for multiple hypotheses, the ability to acknowledge clinically significant differences as statistically significant can be important in describing treatment benefits in regulatory labeling for a treatment.

### Group sequential designs for each hypothesis {.tabset}

For the example, we assume 1-sided testing or a non-binding futility bound as required for a @MaurerBretz2013 design using group sequential design with graphical multiplicity control.
Each is demonstrated in the example code for respective hypotheses.
Efficacy $\alpha$-spending for all group sequential designs uses the @LanDeMets spending function approximating an O'Brien-Fleming bound.

This section needs to be modified by the user to match the study design under consideration.
Those uncomfortable with coding may wish to design using the
[rpact Shiny app](https://www.rpact.org/) or other validated tooling which provides not only a point and click interface, but also reproducible R code.

#### H1: OS, Subgroup

We assume 50% of the population is in the subgroup of interest.
A sample size of 378 is driven by overall survival (OS) in the subgroup where we assume a hazard ratio of 0.65.
Here we assume a one-sided group sequential design with no futility bound.

```{r, results='asis'}
osmedian <- 12 # Median control survival
# Group sequential design for OS in the subgroup (rpact)
ossub <- rpact::getDesignGroupSequential(
  kMax = 3,
  alpha = alphaHypotheses[1],
  beta = 0.1,
  sided = 1,
  informationRates = c(0.61, 0.82, 1),
  typeOfDesign = "asOF"
)
print(ossub)
```

```{r}
as.data.frame(ossub) %>%
  select(
    stages, informationRates, criticalValues,
    stageLevels, alphaSpent
  ) %>%
  kable(caption = "Design for OS in the Subgroup (rpact)") %>%
  kable_styling()
```

#### H2: OS, All

The total sample size is assumed to be twice the above, N=756.
The power and hazard ratio can be adjusted to appropriately size the trial rather than starting with adjusting sample size to reach a targeted power.
For this example, we consider altering power (`beta`) while fixing the hazard ratio at 0.75, representing an increase in median OS from 12 months in the control group to 16 months in the experimental group.
For this design, we consider a non-binding futility bound where the trial may be stopped early in the overall population if the bound is crossed.
We use a @HwangShihDeCani bound with $\gamma = -3.25$.
Study designers should carefully consider implication for parameter choices, particularly if the futility bounds provide sensible guidance for stopping the trial.
Since the futility bounds are non-binding, the efficacy bound is computed assuming the futility bound is ignored which will control Type I error at the targeted level even if a futility bound is crossed and the trial is continued.

```{r, results = 'asis'}
hr <- .75
beta <- .14

os <- rpact::getDesignGroupSequential(
  kMax = 3,
  alpha = 0.01,
  beta = beta,
  sided = 1,
  informationRates = c(0.62, 0.83, 1),
  typeOfDesign = "asOF",
  typeBetaSpending = "bsHSD",
  gammaB = -3.25,
  bindingFutility = FALSE,
  futilityStops = c(TRUE, FALSE)
)
print(os)
```

```{r}
as.data.frame(os) %>%
  select(
    stages, informationRates, criticalValues,
    futilityBounds, stageLevels, alphaSpent
  ) %>%
  kable(caption = "Design for OS in all subjects (rpact)") %>%
  kable_styling()
```

We can also plot different design characteristics. Here, we plot the approximate hazard ratio to cross each bound which may be helpful for design team discussions.

```{r}
plot(os, plottype = "HR", xlab = "Events")
```

#### H3: PFS, Subgroup

For progression free survival (PFS) we assume a shorter median time to event of 5 months.
With an assumed hazard ratio of 0.65, we adjust `beta` and `timing` to match the targeted sample size and interim analysis timing.
We assume a larger dropout rate for PFS than we did for OS.
Here we set up a futility bound for safety.
This is an asymmetric 2-sided design with both futility and efficacy boundary crossing probabilities under the null hypothesis. The parameter `astar = 0.1` specifies total lower bound spending of 10%.
The lower @HwangShihDeCani spending bound with $\gamma = -8$ is intended to be conservative in terms of futility at the interim, but still provide a safety bound for the PFS result in this targeted population.
This would have to be carefully evaluated by the study design team at the time of design.

```{r, results='asis'}
hr <- .65
beta <- .149

pfssub <- rpact::getDesignGroupSequential(
  kMax = 2,
  alpha = 0.004,
  beta = beta,
  sided = 1,
  informationRates = c(0.87, 1),
  typeOfDesign = "asOF",
  typeBetaSpending = "bsHSD",
  gammaB = -8,
  bindingFutility = FALSE,
  futilityStops = TRUE
)
print(pfssub)
```

```{r}
as.data.frame(pfssub) %>%
  select(
    stages, informationRates, criticalValues,
    futilityBounds, stageLevels, alphaSpent
  ) %>%
  kable(caption = "Design for PFS in the subgroup (rpact)") %>%
  kable_styling()
```

#### H4: PFS, All

Finally, we design for PFS in all subjects.
In this case, we simplify to a one-sided design.
A futility bound could be considered, if appropriate.

```{r}
hr <- .74
beta <- .15

pfs <- rpact::getDesignGroupSequential(
  kMax = 2,
  alpha = 0.004,
  beta = beta,
  sided = 1,
  informationRates = c(0.86, 1),
  typeOfDesign = "asOF"
)

as.data.frame(pfs) %>%
  select(
    stages, informationRates, criticalValues,
    stageLevels, alphaSpent
  ) %>%
  kable(caption = "Design for PFS in the overall population (rpact)") %>%
  kable_styling()
```

#### H5 and H6: ORR

For objective response rate (ORR), we assume an underlying control rate of 15%.
In the subgroup population, we have almost 90% power to detect a 20% improvement.

```{r}
# Approximate power for ORR in subgroup (normal approximation)
rpact::getPowerRates(
  groups = 2,
  pi1 = 0.35,
  pi2 = 0.15,
  riskRatio = FALSE,
  alpha = 0.0005,
  maxNumberOfSubjects = 378
)
```

In the all subjects population, we have approximately 95% power to detect an improvement in ORR from 15% to 30%.

```{r}
rpact::getPowerRates(
  groups = 2,
  pi1 = 0.30,
  pi2 = 0.15,
  riskRatio = FALSE,
  alpha = 0.0005,
  maxNumberOfSubjects = 756
)
```

### Design list

Now we associate designs with hypotheses in an ordered list corresponding to the order in the multiplicity graph setup.
Since ORR designs are not group sequential, we enter `NULL` values for those in the last 2 entries of the design list; hit code button to reveal code for this.

```{r}
### THIS NEEDS TO BE MODIFIED TO MATCH STUDY
gsDlist <- list(ossub, os, pfssub, pfs, NULL, NULL)
```

### Spending plan and spending time

While it was relatively straightforward above to set up timing of analyses to match for the different hypotheses, accumulation of endpoints can vary from plan in a variety of ways.
Planning on how to deal with this is critical at the time of protocol development to avoid later amendments or inappropriate $\alpha$-allocation to early analyses.
Before going into examples, we review the concept of $\alpha$-spending and what we will refer to as _spending time_.

For a given hypothesis, we will assign a non-decreasing spending function $f(t)$ defined for $t\ge 0$ with $f(0)=0$ and $f(t)=\alpha$ for $t\ge 1$.
We will assume $K$ analyses with observed event counts $n_k$ at analysis $k=1,2,\ldots,K$ and a targeted final event count of $N_k$.
The $\alpha$-spending at analysis $k$ was originally defined [@LanDeMets] as $f(t_k=n_k/N_K)$.
The values $n_k/N_K$ will be referred to as the _information fraction_, $k=1,\ldots,K$.
This is used to pre-specify the cumulative amount of Type I error for a hypothesis at each analysis.
In @LanDeMets1989 they noted that calendar time was another option for $t_k$ values, $k=1,\ldots,K.$
@PLWBook noted further that as long as $t_k$ is increasing with $k$, it can be used to define spending; this is subject to the requirement that under the null hypothesis, the timing must be selected in a way that is not correlated with the test statistic (e.g., blinded).
We will refer to $t_k$, regardless of its definition, as the _spending time_ for a hypothesis.
Note that the joint distribution of interim and final tests for a hypothesis is driven by $n_k$, $k=1,\ldots,K$.
This is equivalent to basing correlation on the information fraction $n_k^{(actual)}/n_K^{(planned)}$, $1\le k\le K$.
Thus, both spending time and information fraction are required to compute bounds for group sequential testing.
Our general objectives here will be to:

- Spend all Type I error for each hypothesis in its combined interim and final analyses; this requires the information fraction to be 1 for the final analysis of a hypothesis.
- Ensure the information fraction is well defined for each analysis of each hypothesis.
- We will assume that both follow-up duration and event counts may be of interest in determining timing of analyses; e.g., for immuno-oncology therapies there have been delayed treatment effects and the tail of the time-to-event distribution has been important to establish benefit. Thus, we will assume here that over-spending at interim analysis is to be avoided.

Since rpact does not support a separate spending time, we use the actual information fraction $n_k/n_K$ for each hypothesis to drive $\alpha$-spending.
This means we need each hypothesis to reach its targeted final event count so that the information fraction reaches 1 at the final analysis.
Consequently, analysis timing targets both the subgroup and overall population event counts, and we allow 3 additional months of delay at each analysis (compared to other vignettes in this series) to ensure both populations accumulate sufficient events.

## Simulated trial data

We simulate a single trial instance using an illness-death model with response.
The model has four states: alive without response or progression (0),
responded without progression (1), progressed (2), and dead (3).

```{r, message=FALSE}
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "sim_illness_death.R"))
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "cut_illness_death.R"))
```

```{r}
prevalence_sub <- 0.50
pfsmedian <- 5

transition_rate <- build_transition_rates(
  strata = c("Subgroup", "Complement"),
  treatments = c("control", "experimental"),
  median_pfs = c("Subgroup" = pfsmedian, "Complement" = pfsmedian),
  median_os = c("Subgroup" = osmedian, "Complement" = osmedian),
  orr = list(
    "Subgroup" = c(control = 0.15, experimental = 0.35),
    "Complement" = c(control = 0.15, experimental = 0.25)
  ),
  hr_pfs = c("Subgroup" = 0.65, "Complement" = 0.85),
  hr_os = c("Subgroup" = 0.65, "Complement" = 0.85)
)
```

### Simulate one trial

```{r}
set.seed(20250328)
n_sub <- 378  # Final targeted subgroup N from design narrative
n_total <- ceiling(n_sub / prevalence_sub)
steady_rate <- n_total / 15  # 18-month enrollment: 0.25*2 + 0.5*2 + 0.75*2 + 1*12 = 15 rate-months

sim_data <- sim_illness_death(
  n = n_total,
  stratum = data.frame(stratum = c("Subgroup", "Complement"),
                       p = c(prevalence_sub, 1 - prevalence_sub)),
  block = c("control", "control", "experimental", "experimental"),
  enroll_rate = data.frame(
    rate = steady_rate * c(0.25, 0.50, 0.75, 1.00),
    duration = c(2, 2, 2, 12)
  ),
  transition_rate = transition_rate,
  dropout_rate = 0.001
)

fpe <- max(sim_data$ENRLTIME)
cat("Total N:", nrow(sim_data), "\n")
cat("N Subgroup:", sum(sim_data$STRATUM == "Subgroup"),
    "N Complement:", sum(sim_data$STRATUM == "Complement"), "\n")
cat("FPE:", round(fpe, 1), "months\n")
```

### Determine analysis cut dates and compute p-values

```{r}
target_pfs_events_sub <- 297  # From design narrative (pfssub final events)
target_os_events_sub <- 284   # From design narrative (ossub final events)
target_pfs_events_all <- 594  # Approximate 2x subgroup for overall population
target_os_events_all <- 568   # Approximate 2x subgroup for overall population

# Determine cut dates targeting subgroup event counts
analyses_sub <- list(
  list(min_followup = 6, endpoint = NULL, event_target = NULL,
       target_stratum = NULL, max_followup = NULL),
  list(min_followup = 14, endpoint = "PFS", event_target = target_pfs_events_sub,
       target_stratum = "Subgroup", max_followup = 20),
  list(min_followup = 24, endpoint = "OS", event_target = target_os_events_sub,
       target_stratum = "Subgroup", max_followup = 33)
)

# Determine cut dates targeting overall population event counts
analyses_all <- list(
  list(min_followup = 6, endpoint = NULL, event_target = NULL,
       target_stratum = NULL, max_followup = NULL),
  list(min_followup = 14, endpoint = "PFS", event_target = target_pfs_events_all,
       target_stratum = NULL, max_followup = 20),
  list(min_followup = 24, endpoint = "OS", event_target = target_os_events_all,
       target_stratum = NULL, max_followup = 33)
)

cut_dates_sub <- get_analysis_dates(sim_data, analyses_sub)
cut_dates_all <- get_analysis_dates(sim_data, analyses_all)
# Use the later of the two dates at each analysis
cut_dates <- pmax(cut_dates_sub, cut_dates_all)

cat("Analysis cut dates:\n")
for (i in seq_along(cut_dates)) {
  cat(sprintf("  Analysis %d: %.1f months (FPE + %.1f)\n",
              i, cut_dates[i], cut_dates[i] - fpe))
}
```

```{r}
adtte <- lapply(seq_along(cut_dates), function(i) {
  d <- cut_illness_death(sim_data, cut_dates[i])
  d$ANALYSIS <- i
  d
})

# Helper: 1-sided logrank p-value
logrank_pval <- function(data, stratified = FALSE) {
  data$event <- 1L - data$CNSR
  if (stratified) {
    fit <- survdiff(Surv(AVAL, event) ~ TRT + strata(STRATUM), data = data)
  } else {
    fit <- survdiff(Surv(AVAL, event) ~ TRT, data = data)
  }
  obs_ctrl <- if (is.matrix(fit$obs)) sum(fit$obs[1, ]) else fit$obs[1]
  exp_ctrl <- if (is.matrix(fit$exp)) sum(fit$exp[1, ]) else fit$exp[1]
  z <- sign(obs_ctrl - exp_ctrl) * sqrt(fit$chisq)
  pnorm(-z)
}

# Helper: 1-sided rate difference p-value
rd_pval <- function(data, stratified = FALSE) {
  if (!stratified) {
    n_exp <- sum(data$TRT == "experimental")
    n_ctrl <- sum(data$TRT == "control")
    p_exp <- sum(data$AVAL[data$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(data$AVAL[data$TRT == "control"]) / n_ctrl
    p_pool <- (p_exp * n_exp + p_ctrl * n_ctrl) / (n_exp + n_ctrl)
    se <- sqrt(p_pool * (1 - p_pool) * (1 / n_exp + 1 / n_ctrl))
    if (se == 0) return(0.5)
    return(pnorm(-(p_exp - p_ctrl) / se))
  }
  strata <- unique(data$STRATUM)
  n_s <- rd_s <- var_s <- numeric(length(strata))
  for (k in seq_along(strata)) {
    ds <- data[data$STRATUM == strata[k], ]
    n_exp <- sum(ds$TRT == "experimental")
    n_ctrl <- sum(ds$TRT == "control")
    n_s[k] <- n_exp + n_ctrl
    p_exp <- sum(ds$AVAL[ds$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(ds$AVAL[ds$TRT == "control"]) / n_ctrl
    rd_s[k] <- p_exp - p_ctrl
    var_s[k] <- p_exp * (1 - p_exp) / n_exp + p_ctrl * (1 - p_ctrl) / n_ctrl
  }
  w <- n_s / sum(n_s)
  z <- sum(w * rd_s) / sqrt(sum(w^2 * var_s))
  pnorm(-z)
}

pvals <- list()
for (i in seq_along(cut_dates)) {
  d <- adtte[[i]]
  pvals[[paste0("H1_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS", STRATUM == "Subgroup"))
  pvals[[paste0("H2_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS"), stratified = TRUE)
  if (i <= 2) {
    pvals[[paste0("H3_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS", STRATUM == "Subgroup"))
    pvals[[paste0("H4_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS"), stratified = TRUE)
  }
  if (i == 1) {
    pvals[[paste0("H5_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR", STRATUM == "Subgroup"))
    pvals[[paste0("H6_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR"), stratified = TRUE)
  }
}
```

## Results entry at time of analysis

### Build results table

Event counts and p-values are assembled from the simulated trial data.
Note that rpact does not support spending time; we use the information fraction instead.

```{r,warnings=FALSE,message=FALSE}
get_events <- function(analysis_idx, paramcd, stratum = NULL) {
  d <- adtte[[analysis_idx]]
  if (!is.null(stratum)) d <- d %>% filter(STRATUM == stratum)
  sum(d$PARAMCD == paramcd & d$CNSR == 0, na.rm = TRUE)
}

events_os_sub <- sapply(1:3, function(i) get_events(i, "OS", "Subgroup"))
events_os_all <- sapply(1:3, function(i) get_events(i, "OS"))
events_pfs_sub <- sapply(1:2, function(i) get_events(i, "PFS", "Subgroup"))
events_pfs_all <- sapply(1:2, function(i) get_events(i, "PFS"))

inputResults <- tibble(
  H = c(rep(1, 3), rep(2, 3), rep(3, 2), rep(4, 2), 5, 6),
  Pop = c(
    rep("Subgroup", 3), rep("All", 3),
    rep("Subgroup", 2), rep("All", 2),
    "Subgroup", "All"
  ),
  Endpoint = c(rep("OS", 6), rep("PFS", 4), rep("ORR", 2)),
  nominalP = c(
    pvals$H1_1, pvals$H1_2, pvals$H1_3,
    pvals$H2_1, pvals$H2_2, pvals$H2_3,
    pvals$H3_1, pvals$H3_2,
    pvals$H4_1, pvals$H4_2,
    pvals$H5_1,
    pvals$H6_1
  ),
  Analysis = c(1:3, 1:3, 1:2, 1:2, 1, 1),
  events = c(events_os_sub, events_os_all, events_pfs_sub, events_pfs_all, NA, NA),
  # Use information fraction as spending time for each hypothesis
  spendingTime = c(
    events_os_sub / max(events_os_sub),
    events_os_all / max(events_os_all),
    events_pfs_sub / max(events_pfs_sub),
    events_pfs_all / max(events_pfs_all),
    NA, NA
  )
)
kable(inputResults, caption = "Results from simulated trial data") %>%
  kable_styling()
```

## Testing hypotheses

### Compute sequential p-values for each hypothesis

Sequential p-value computation is done in one loop in an attempt to minimize chances for coding errors.
We delay showing these until after display of the sequence of multiplicity graphs generated by hypothesis rejection is shown.
NOTE: THIS NEEDS TO BE CHECKED AS SPENDING TIME MAY NOT BE AN OPTION IN rpact.

```{r,message=FALSE}
### USER SHOULD NOT NEED TO MODIFY THIS CODE
sequentialPValue <- function(
  design_type,
  sided,
  informationRates,
  events,
  nominalP,
  futility = NULL
) {
  kMax <- length(informationRates)
  if (length(events) != kMax || length(nominalP) != kMax) {
    stop("`events`, `nominalP`, and `informationRates` must have same length.")
  }

  # Build rpact design; repeated p-values are invariant to `alpha` here
  if (is.null(futility)) {
    design <- rpact::getDesignGroupSequential(
      kMax = kMax,
      alpha = 0.025,
      beta = 0.2,
      sided = sided,
      informationRates = informationRates,
      typeOfDesign = design_type
    )
  } else {
    design <- rpact::getDesignGroupSequential(
      kMax = kMax,
      alpha = 0.025,
      beta = futility$beta,
      sided = sided,
      informationRates = informationRates,
      typeOfDesign = design_type,
      typeBetaSpending = futility$typeBetaSpending,
      gammaB = futility$gammaB,
      bindingFutility = isTRUE(futility$bindingFutility),
      futilityStops = futility$futilityStops
    )
  }

  # 1-sided z-scores (upper direction) from nominal p-values
  z <- stats::qnorm(1 - nominalP)
  dat <- rpact::getDataset(
    cumEvents = events,
    cumLogRanks = z,
    cumAllocationRatios = rep(1, kMax)
  )
  sr <- rpact::getStageResults(design = design, dataInput = dat, stage = kMax)
  as.numeric(rpact::getRepeatedPValues(sr))[kMax]
}

EOCtab <- NULL
EOCtab <- inputResults %>%
  group_by(H) %>%
  slice(1) %>%
  ungroup() %>%
  select("H", "Pop", "Endpoint", "nominalP")
EOCtab$seqp <- .9999
for (EOCtabline in 1:nHypotheses) {
  EOCtab$seqp[EOCtabline] <-
    ifelse(is.null(gsDlist[[EOCtabline]]), EOCtab$nominalP[EOCtabline], {
      tem <- filter(inputResults, H == EOCtabline)
      # Sequential p-value = repeated p-value at last available analysis
      d <- gsDlist[[EOCtabline]]

      fut <- NULL
      if (EOCtabline == 2) {
        fut <- list(
          beta = beta,
          typeBetaSpending = "bsHSD",
          gammaB = -3.25,
          bindingFutility = FALSE,
          futilityStops = c(TRUE, FALSE)
        )
      }
      if (EOCtabline == 3) {
        fut <- list(
          beta = beta,
          typeBetaSpending = "bsHSD",
          gammaB = -8,
          bindingFutility = FALSE,
          futilityStops = TRUE
        )
      }

      sequentialPValue(
        design_type = as.data.frame(d)$typeOfDesign[1],
        sided = as.data.frame(d)$sided[1],
        informationRates = tem$spendingTime,
        events = tem$events,
        nominalP = tem$nominalP,
        futility = fut
      )
    })
}
EOCtab <- EOCtab %>% select(-"nominalP")
# kable(EOCtab,caption="Sequential p-values as initially placed in EOCtab") %>% kable_styling()
```

### Evaluate hypothesis rejection using graphicalMCP

We need to set up an `initial_graph` object as implemented in the graphicalMCP package.

```{r,message=FALSE,warning=FALSE}
library(graphicalMCP)

g0 <- graphicalMCP::graph_create(
  hypotheses = alphaHypotheses / fwer,
  transitions = m,
  hyp_names = nameHypotheses
)
```

Now we add the sequential p-values and evaluate which hypotheses have been rejected.

```{r}
result <- graphicalMCP::graph_test_shortcut(
  graph = g0,
  p = EOCtab$seqp,
  alpha = fwer,
  verbose = TRUE
)

adj_p <- result$outputs$adjusted_p
rej <- result$outputs$rejected

if (is.numeric(rej)) {
  rej_logical <- rep(FALSE, nHypotheses)
  rej_logical[rej] <- TRUE
} else {
  rej_logical <- as.logical(rej)
}

EOCtab$Rejected <- rej_logical
EOCtab$adjPValues <- adj_p
```

## Verification of hypotheses rejected

```{r}
rejected_order <- which(EOCtab$Rejected)[order(EOCtab$adjPValues[EOCtab$Rejected])]

graphs <- list(g0)
if (length(rejected_order) > 0) {
  gu <- graphicalMCP::graph_update(g0, delete = rejected_order)
  # `intermediate_graphs` includes the initial graph as its first element
  graphs <- gu$intermediate_graphs
}
ngraphs <- length(graphs)

# Final weights for hypotheses not rejected come from the final graph
lastWeights <- as.numeric(graphs[[ngraphs]]$hypotheses)
lastGraph <- rep(ngraphs, nrow(EOCtab))

# For rejected hypotheses, take the hypothesis weight from the graph at which it was rejected
if (length(rejected_order) > 0) {
  for (j in seq_along(rejected_order)) {
    h <- rejected_order[j]
    lastWeights[h] <- as.numeric(graphs[[j]]$hypotheses[h])
    lastGraph[h] <- j
  }
}

EOCtab$lastAlpha <- fwer * lastWeights
EOCtab$lastGraph <- lastGraph
EOCtabx <- EOCtab
names(EOCtabx) <- c(
  "Hypothesis", "Population", "Endpoint", "Sequential p",
  "Rejected", "Adjusted p", "Max alpha allocated", "Last Graph"
)
# Display table with desired column order
# Delayed following until after multiplicity graph sequence
# EOCtabx %>% select(c(1:4,7,5:6,8)) %>% kable() %>% kable_styling()
```

### Multiplicity graph sequence from graphicalMCP {.tabset}

```{r,message=FALSE,warning=FALSE,results='asis',fig.asp=0.95}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
for (i in 1:ngraphs) {
  cat(" \n")
  cat("####", paste(" Graph", as.character(i), " \n\n"))
  plot(graphs[[i]],
    layout = layout6,
    vertex.size = 45,
    vertex.label.cex = 0.9,
    vertex.color = unname(vertex_color_map[names(graphs[[i]]$hypotheses)]),
    margin = 0.25
  )
  cat(" \n\n\n")
}
```

### Comparison of sequential p-values to multiplicity graphs

We can compare sequential p-values to available $\alpha$ in each graph.
In the column 'Last Graph' we can see one of 2 things:

1) For rejected hypotheses, the maximum $\alpha$ allocated to the hypothesis. For example, hypothesis one was allocated $\alpha=0.01$ in the first graph above (select using first tab). We see that the sequential p-value of 0.0001 is smaller than $\alpha=0.01$ and thus the hypothesis is rejected. We can then proceed to the second graph and see that hypothesis 5 was rejected. The last hypothesis rejected is hypothesis 3 in the third graph.
2) For the remaining hypotheses (H2, H4, H6) the maximum $\alpha$ allocated is in the fourth graph; since each sequential p-value is greater than the allocated $\alpha$ for the corresponding hypothesis, none of these hypotheses were rejected.

```{r}
EOCtabx %>%
  select(c(1:4, 7, 5:6, 8)) %>%
  kable() %>%
  kable_styling()
```

### Bounds at final $\alpha$ allocated for group sequential tests {.tabset}

As a separate validation, we examine group sequential bounds for each hypothesis updated with 1) the maximum $\alpha$ allocated above, 2) the number of events at each analysis, and 3) the cumulative spending at each analysis above. The nominal p-value for at least one of the analyses performed for each rejected hypotheses should be less than or equal to the nominal p-value in the group sequential design.
For each hypothesis not rejected, all nominal p-values are greater than the its corresponding bound.
For hypotheses tested without a group sequential design, the nominal p-value for the test of that hypothesis can be compared to the maximum alpha allocated in the above table.

```{r,results='asis'}
for (i in 1:nHypotheses) {
  # Set up tab for hypothesis in output
  cat("####", paste(" Hypothesis", as.character(i), " \n"))
  # Get results for hypothesis
  hresults <- inputResults %>% filter(H == i)
  # Print out max alpha allocated
  xx <- paste("Max alpha allocated from above table: ",
    as.character(EOCtab$lastAlpha[i]),
    sep = ""
  )
  d <- gsDlist[[i]]
  # If not group sequential for this hypothesis, print the max alpha allocated
  # and the nominal p-value
  if (is.null(d)) {
    cat("Maximum alpha allocated: ")
    cat(EOCtab$lastAlpha[i])
    cat("\n\n")
    cat("Nominal p-value for hypothesis test: ")
    cat(hresults$nominalP)
    cat("\n\n")
  }
  # For group sequential tests, print max alpha allocated and
  # corresponding group sequential bounds
  if (!is.null(gsDlist[[i]])) {
    cat("Nominal p-values at each analysis for comparison to bounds in table below:\n\n")
    cat(hresults$nominalP)
    cat("\n\n")
    # If no alpha allocated, just print text line to note this along with the 0 alpha allocated
    if (EOCtab$lastAlpha[i] == 0) {
      cat("Maximum alpha allocated: 0\n\n")
      cat("No testing required\n\n")
    }
    if (EOCtab$lastAlpha[i] > 0) {
      # Update design at the maximum allocated alpha and observed spending time
      informationRates <- hresults$spendingTime
      kMax <- length(informationRates)

      fut <- NULL
      if (i == 2) {
        fut <- list(
          beta = beta,
          typeBetaSpending = "bsHSD",
          gammaB = -3.25,
          bindingFutility = FALSE,
          futilityStops = c(TRUE, FALSE)
        )
      }
      if (i == 3) {
        fut <- list(
          beta = beta,
          typeBetaSpending = "bsHSD",
          gammaB = -8,
          bindingFutility = FALSE,
          futilityStops = TRUE
        )
      }

      if (is.null(fut)) {
        dupdate <- rpact::getDesignGroupSequential(
          kMax = kMax,
          alpha = EOCtab$lastAlpha[i],
          beta = 0.2,
          sided = 1,
          informationRates = informationRates,
          typeOfDesign = as.data.frame(d)$typeOfDesign[1]
        )
      } else {
        dupdate <- rpact::getDesignGroupSequential(
          kMax = kMax,
          alpha = EOCtab$lastAlpha[i],
          beta = fut$beta,
          sided = 1,
          informationRates = informationRates,
          typeOfDesign = as.data.frame(d)$typeOfDesign[1],
          typeBetaSpending = fut$typeBetaSpending,
          gammaB = fut$gammaB,
          bindingFutility = isTRUE(fut$bindingFutility),
          futilityStops = fut$futilityStops
        )
      }

      as.data.frame(dupdate) %>%
        select(
          stages,
          informationRates,
          criticalValues,
          stageLevels,
          alphaSpent,
          futilityBounds
        ) %>%
        kable(caption = xx, row.names = FALSE) %>%
        kable_styling() %>%
        cat()
      cat("\n\n")
    }
  }
}
```

```{r}
# Restore default options
options(old)
```

## Session information

You can use `sessionInfo()` to document the versions of R and R packages used to render this document.
Note, in particular, that the **rpact** package is required.

## References

```{js,echo=FALSE}
// Fix dynamic tabsets code block rendering issue under pkgdown
(function () {
  // Fix the tweaked element class
  unstyledCode = document.querySelectorAll("button code");

  // Quit if there are no elements matched (when not rendered by pkgdown)
  if (!unstyledCode.length) {
    return;
  }

  for (let i = 0; i < unstyledCode.length; i++) {
    blockOld = unstyledCode[i].outerHTML;
    blockNew =
      "<pre class='downlit sourceCode r' style='text-align:left;white-space:pre-wrap;'>" +
      blockOld +
      "</pre>";
    unstyledCode[i].outerHTML = blockNew;
  }

  // Activate the first real tab
  var tabElements = [
    "ul#multiplicity-graph-sequence-from-graphicalmcp li:nth-child(2) button",
    "ul#bounds-at-final-alpha-allocated-for-group-sequential-tests li:nth-child(2) button",
  ];

  for (let i = 0; i < tabElements.length; i++) {
    var el = document.querySelector(tabElements[i]);
    var tab = new bootstrap.Tab(el);
    tab.show();
  }
})();
```
