---
title: "New Trial Design: Group sequential design using graphicalMCP and gsDesign2"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
bibliography: gMCPLite.bib
vignette: >
  %\VignetteIndexEntry{New Trial Design: Group sequential design using graphicalMCP and gsDesign2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1,
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)
```

## Overview

This document evaluates statistical significance for graphical multiplicity control with group sequential design following the Maurer-Bretz (2013) framework. The combination of graphicalMCP and gsDesign2 packages is used for design and analysis.

The trial tests 3 endpoints across 2 populations, yielding 6 hypotheses:

| Hypothesis | Endpoint | Population | Analyses |
|------------|----------|------------|----------|
| H1 | OS | BM+ | 3 |
| H2 | OS | Overall | 3 |
| H3 | PFS | BM+ | 2 |
| H4 | PFS | Overall | 2 |
| H5 | ORR | BM+ | 1 |
| H6 | ORR | Overall | 1 |

Key design features:

- Enrollment over 14 months with a 6-month ramp-up.
- BM+ prevalence assumed at 60%.

### Analysis timing rules

All timing is relative to final patient enrolled (FPE). Enrollment is planned over 14 months.

| Analysis | Timing rule (after FPE) | Max extension | Endpoints assessed |
|----------|------------------------|---------------|-------------------|
| IA1 | 6 months after FPE | None | ORR (final for both populations), PFS and OS (interim) |
| IA2 | 14 months after FPE AND targeted PFS event count in BM+ reached | Up to 17 months after FPE (+3 months) | PFS (final for both populations), OS (interim) |
| Final | 24 months after FPE AND targeted OS event count in BM+ reached | Up to 30 months after FPE (+6 months) | OS (final for both populations) |

Notes on timing rules:

- IA1 is calendar-driven to ensure 6 months minimum follow-up for ORR assessment in all patients. PFS and OS are analyzed with whatever event counts are available at the cutoff.
- IA2 is both calendar- and event-driven: we require at least 14 months after FPE to ensure minimum follow-up of nearly 3x the assumed control median PFS, but also require the targeted BM+ PFS event count. If the targeted events are not reached by 14 months after FPE, the analysis may be delayed up to 3 months.
- The final analysis is event-driven for OS in BM+. If the targeted event count is not reached by 24 months after FPE, the analysis may be delayed up to 6 months (30 months after FPE).

## Design

```{r, message=FALSE, warning=FALSE}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
old <- options(scipen = 999)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
trt_colors <- c("control" = "#4E79A7", "experimental" = "#E15759")
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(gt)
library(ggplot2)
library(gsDesign)
library(gsDesign2)
library(graphicalMCP)
library(simtrial)
library(survival)

if (!"sequential_pval" %in% getNamespaceExports("gsDesign2")) {
  stop(
    "This vignette requires a version of gsDesign2 that exports `sequential_pval()`.\n",
    "Install the GitHub version from `Merck/gsDesign2`."
  )
}
```

### Multiplicity diagram for hypothesis testing

Alpha is split equally between BM+ and Overall populations for OS, with smaller allocations to PFS and ORR. Reallocation flows from BM+ to Overall within each endpoint, and from Overall to the next endpoint's populations.

```{r, warning=FALSE, message=FALSE, fig.asp=0.9}
### MODIFY HYPOTHESES, ALPHA, AND TRANSITIONS FOR YOUR STUDY
nameHypotheses <- c(
  "H1: OS\n BM+",
  "H2: OS\n Overall",
  "H3: PFS\n BM+",
  "H4: PFS\n Overall",
  "H5: ORR\n BM+",
  "H6: ORR\n Overall"
)
nHypotheses <- length(nameHypotheses)

# Transition weights for alpha reallocation
m <- matrix(c(
  0, 1, 0, 0, 0, 0,
  0, 0, .5, .5, 0, 0,
  0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, .5, .5,
  0, 0, 0, 0, 0, 1,
  .5, .5, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)

# Initial Type I error assigned to each hypothesis (one-sided)
alphaHypotheses <- c(.01, .01, .004, 0.000, 0.0005, .0005)
fwer <- sum(alphaHypotheses)

fill_idx <- c(2, 2, 4, 4, 3, 3)
vertex_color_map <- stats::setNames(cbPalette[fill_idx], nameHypotheses)

g0 <- graphicalMCP::graph_create(
  hypotheses = stats::setNames(alphaHypotheses / fwer, nameHypotheses),
  transitions = m,
  hyp_names = nameHypotheses
)

layout6 <- rbind(
  c(0.28, 0.74),
  c(0.72, 0.74),
  c(0.18, 0.50),
  c(0.82, 0.50),
  c(0.28, 0.26),
  c(0.72, 0.26)
)

# Build vertex labels with alpha character
alphaLabels <- sprintf("\u03b1 = %s", format(alphaHypotheses, scientific = FALSE))
vertexLabels <- paste(nameHypotheses, alphaLabels, sep = "\n")

plot(g0,
  layout = layout6,
  vertex.size = 45,
  vertex.label = vertexLabels,
  vertex.label.cex = 0.9,
  vertex.color = unname(vertex_color_map[names(g0$hypotheses)]),
  margin = 0.25
)
```

### Common design parameters

BM+ prevalence is 60%. Enrollment occurs over 14 months with a 6-month ramp-up.

```{r}
### MODIFY THESE PARAMETERS FOR YOUR STUDY
bm_prev <- 0.60
osmedian <- 12  # Control median OS (months)
pfsmedian <- 5  # Control median PFS (months)
```

### Group sequential designs for each hypothesis {.tabset}

H1 (OS in BM+) is the sample-size-driving hypothesis, designed with `gs_design_ahr()` targeting 90% power.
All other hypotheses use the enrollment rates derived from H1 and compute power with `gs_power_ahr()` (time-to-event) or `fixed_design_rd()` (ORR).

#### H1: OS, BM+ (sample-size-driving)

BM+ subgroup only. Control median OS = `r osmedian` months, HR = 0.65. Target 90% power.
This design determines the BM+ enrollment rate and sample size for all other hypotheses.

```{r}
fail_rate_os_sub <- gsDesign2::define_fail_rate(
  duration = Inf,
  fail_rate = log(2) / osmedian,
  hr = 0.65,
  dropout_rate = 0.001
)

# Initial BM+ enrollment rate for design (will be updated by gs_design_ahr)
enroll_rate_sub_init <- gsDesign2::define_enroll_rate(
  duration = c(2, 2, 2, 8),
  rate = c(0.25, 0.50, 0.75, 1.00)  # Relative rates; gs_design_ahr will scale
)

ossub <- gsDesign2::gs_design_ahr(
  enroll_rate = enroll_rate_sub_init,
  fail_rate = fail_rate_os_sub,
  alpha = alphaHypotheses[1],
  beta = 0.1,
  binding = FALSE,
  analysis_time = c(20, 28, 38),  # FPE+6, FPE+14, FPE+24 (14 mo enrollment + follow-up)
  info_frac = NULL,                # Derived from analysis_time + enroll/fail rates
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[1]),
  test_lower = FALSE
)

gsDesign2::gs_bound_summary(ossub) %>%
  gt::gt() %>%
  gt::tab_header(title = "Design for OS in BM+ (gsDesign2)")
```

#### Derive enrollment rates and sample sizes from H1

The BM+ enrollment rate from H1 determines all other designs.
The BM- enrollment rate is derived from the BM+ prevalence.

```{r}
# Extract BM+ enrollment rate from H1 design
enroll_rate_sub <- ossub$enroll_rate
n_bm_pos <- sum(enroll_rate_sub$rate * enroll_rate_sub$duration)
n_bm_neg <- n_bm_pos * (1 - bm_prev) / bm_prev
n_total <- n_bm_pos + n_bm_neg

cat("BM+ sample size (per arm):", round(n_bm_pos), "\n")
cat("BM- sample size (per arm):", round(n_bm_neg), "\n")
cat("Total sample size (per arm):", round(n_total), "\n")

# Build stratified enrollment rate for overall population designs
enroll_rate_overall <- gsDesign2::define_enroll_rate(
  stratum = rep(c("BM+", "BM-"), each = nrow(enroll_rate_sub)),
  duration = rep(enroll_rate_sub$duration, 2),
  rate = c(
    enroll_rate_sub$rate,                                    # BM+ rates from H1
    enroll_rate_sub$rate * (1 - bm_prev) / bm_prev          # BM- rates scaled by prevalence
  )
)
```

#### H2: OS, Overall

Stratified power calculation using enrollment from H1.

- BM+ (60%): Control median OS = `r osmedian` months, HR = 0.65.
- BM- (40%): Control median OS = `r osmedian` months, HR = 0.85.

```{r}
fail_rate_os_overall <- gsDesign2::define_fail_rate(
  stratum = c("BM+", "BM-"),
  duration = c(Inf, Inf),
  fail_rate = log(2) / osmedian,
  hr = c(0.65, 0.85),
  dropout_rate = 0.001
)

os <- gsDesign2::gs_power_ahr(
  enroll_rate = enroll_rate_overall,
  fail_rate = fail_rate_os_overall,
  ratio = 1,
  event = NULL,
  analysis_time = c(20, 28, 38),  # FPE+6, FPE+14, FPE+24
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[2]),
  test_lower = FALSE,
  binding = FALSE
)

gsDesign2::gs_bound_summary(os) %>%
  gt::gt() %>%
  gt::tab_header(title = "Power for OS in Overall population (gsDesign2)")
```

#### H3: PFS, BM+

Power calculation for BM+ subgroup using enrollment from H1. Control median PFS = `r pfsmedian` months, HR = 0.70.

```{r}
fail_rate_pfs_sub <- gsDesign2::define_fail_rate(
  duration = Inf,
  fail_rate = log(2) / pfsmedian,
  hr = 0.70,
  dropout_rate = 0.001
)

pfssub <- gsDesign2::gs_power_ahr(
  enroll_rate = enroll_rate_sub,
  fail_rate = fail_rate_pfs_sub,
  ratio = 1,
  event = NULL,
  analysis_time = c(20, 28),  # FPE+6, FPE+14
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[3]),
  test_lower = FALSE,
  binding = FALSE
)

gsDesign2::gs_bound_summary(pfssub) %>%
  gt::gt() %>%
  gt::tab_header(title = "Power for PFS in BM+ (gsDesign2)")
```

#### H4: PFS, Overall

Stratified power calculation using enrollment from H1.

- BM+ (60%): Control median PFS = `r pfsmedian` months, HR = 0.70.
- BM- (40%): Control median PFS = `r pfsmedian` months, HR = 0.90.

```{r}
fail_rate_pfs_overall <- gsDesign2::define_fail_rate(
  stratum = c("BM+", "BM-"),
  duration = c(Inf, Inf),
  fail_rate = log(2) / pfsmedian,
  hr = c(0.70, 0.90),
  dropout_rate = 0.001
)

# H4 starts with alpha = 0 but may receive alpha through reallocation.
# Use PFS BM+ alpha for bounds structure; actual testing uses reallocated alpha.
pfs <- gsDesign2::gs_power_ahr(
  enroll_rate = enroll_rate_overall,
  fail_rate = fail_rate_pfs_overall,
  ratio = 1,
  event = NULL,
  analysis_time = c(20, 28),  # FPE+6, FPE+14
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[3]),  # Use H3's alpha for bounds
  test_lower = FALSE,
  binding = FALSE
)

gsDesign2::gs_bound_summary(pfs) %>%
  gt::gt() %>%
  gt::tab_header(title = "Power for PFS in Overall population (gsDesign2)")
```

#### H5: ORR, BM+

Fixed design for rate difference in BM+ subgroup. Control ORR = 30%, experimental ORR = 45%.

```{r}
orr_ctrl_bm_pos <- 0.30
orr_exp_bm_pos <- 0.45

orr_sub <- gsDesign2::fixed_design_rd(
  alpha = alphaHypotheses[5],
  power = NULL,
  p_c = orr_ctrl_bm_pos,
  p_e = orr_exp_bm_pos,
  rd0 = 0,
  n = ceiling(n_bm_pos) * 2  # Total N (both arms) in BM+
)

summary(orr_sub) %>%
  gt::gt() %>%
  gt::fmt_number(columns = "Bound", decimals = 2) %>%
  gt::fmt_number(columns = "Power", decimals = 3) %>%
  gt::tab_header(title = "Power for ORR in BM+ (fixed_design_rd)")
```

#### H6: ORR, Overall

Fixed design for rate difference in overall population, stratified by biomarker status.

- BM+ (60%): Control ORR = 30%, experimental ORR = 45%.
- BM- (40%): Control ORR = 20%, experimental ORR = 24%.

```{r}
orr_ctrl_bm_neg <- 0.20
orr_exp_bm_neg <- 0.24

orr_ctrl_overall <- orr_ctrl_bm_pos * bm_prev + orr_ctrl_bm_neg * (1 - bm_prev)
orr_exp_overall <- orr_exp_bm_pos * bm_prev + orr_exp_bm_neg * (1 - bm_prev)

orr_all <- gsDesign2::fixed_design_rd(
  alpha = alphaHypotheses[6],
  power = NULL,
  p_c = orr_ctrl_overall,
  p_e = orr_exp_overall,
  rd0 = 0,
  n = ceiling(n_total) * 2  # Total N (both arms) overall
)

summary(orr_all) %>%
  gt::gt() %>%
  gt::fmt_number(columns = "Bound", decimals = 2) %>%
  gt::fmt_number(columns = "Power", decimals = 3) %>%
  gt::tab_header(title = "Power for ORR in Overall population (fixed_design_rd)")
```

### Design list

```{r}
gsD2list <- list(ossub, os, pfssub, pfs, NULL, NULL)
```

### Enrollment and analysis timing

Enrollment occurs over 14 months with a 6-month ramp-up:

- Months 1-2: 25% of steady-state rate
- Months 3-4: 50% of steady-state rate
- Months 5-6: 75% of steady-state rate
- Months 7-14: Full steady-state enrollment

Analysis timing relative to final patient enrolled (FPE):

- **IA1 (FPE + 6 months)**: ORR assessed for all patients. Interim PFS and OS.
- **IA2 (FPE + 14 months)**: Final PFS analysis, requires targeted BM+ PFS events (max extension to FPE + 17 months). Interim OS.
- **Final (FPE + 24 months)**: Final OS analysis, requires targeted BM+ OS events (max extension to FPE + 30 months).

## Simulated trial data

We simulate a single trial instance using an illness-death model with response.
The model has four states: alive without response or progression (0),
responded without progression (1), progressed (2), and dead (3).
Piecewise exponential distributions drive all state transitions.

```{r illness-death-diagram, fig.asp=0.8, fig.width=6, out.width="70%"}
# --- Illness-death model diagram ---
par(mar = c(0, 0, 0, 0))
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))

# State positions (diamond layout)
sx <- c(0.50, 0.12, 0.88, 0.50)  # State 0, 1, 2, 3
sy <- c(0.88, 0.50, 0.50, 0.12)
r <- 0.10  # circle radius in user coords

# Draw state circles
theta <- seq(0, 2 * pi, length.out = 100)
state_colors <- c("#56B4E9", "#009E73", "#E69F00", "#999999")
state_labels <- c(
  "Alive\nNo response\nNo progression",
  "Responded\nNo progression",
  "Progressed",
  "Dead"
)

for (i in 1:4) {
  polygon(sx[i] + r * cos(theta), sy[i] + r * 0.9 * sin(theta),
          col = adjustcolor(state_colors[i], alpha.f = 0.3),
          border = state_colors[i], lwd = 2)
  text(sx[i], sy[i], state_labels[i], cex = 0.7, font = 2)
}

# Arrow helper: draw from edge of circle i to edge of circle j
draw_trans <- function(i, j, label_expr, label_pos = 0.5,
                       label_offset = c(0, 0), curve = 0, ...) {
  dx <- sx[j] - sx[i]
  dy <- sy[j] - sy[i]
  ang <- atan2(dy, dx)

  # Start and end at circle edge
  x0 <- sx[i] + r * cos(ang)
  y0 <- sy[i] + r * 0.9 * sin(ang)
  x1 <- sx[j] - r * cos(ang)
  y1 <- sy[j] - r * 0.9 * sin(ang)

  if (curve == 0) {
    arrows(x0, y0, x1, y1, length = 0.10, lwd = 2, ...)
    lx <- x0 + label_pos * (x1 - x0) + label_offset[1]
    ly <- y0 + label_pos * (y1 - y0) + label_offset[2]
  } else {
    # Curved arrow using xspline
    mx <- (x0 + x1) / 2 + curve * (y1 - y0)
    my <- (y0 + y1) / 2 - curve * (x1 - x0)
    xspline(c(x0, mx, x1), c(y0, my, y1), shape = c(0, 1, 0),
            border = "black", lwd = 2, ...)
    # Arrowhead at end
    ang2 <- atan2(y1 - my, x1 - mx)
    arrows(x1 - 0.01 * cos(ang2), y1 - 0.01 * sin(ang2),
           x1, y1, length = 0.10, lwd = 2, ...)
    lx <- mx + label_offset[1]
    ly <- my + label_offset[2]
  }
  text(lx, ly, label_expr, cex = 0.85)
}

# Transitions
draw_trans(1, 2, expression(lambda[resp]), label_pos = 0.45, label_offset = c(-0.06, 0.04))
draw_trans(1, 3, expression(lambda["prog,0"]), label_pos = 0.45, label_offset = c(0.06, 0.04))
draw_trans(1, 4, expression(lambda["death,0"]), label_pos = 0.5, label_offset = c(0.08, 0))
draw_trans(2, 3, expression(lambda["prog,1"]), label_pos = 0.5, label_offset = c(0, 0.06))
draw_trans(2, 4, expression(lambda["death,1"]), label_pos = 0.45, label_offset = c(-0.06, -0.04))
draw_trans(3, 4, expression(lambda["death,2"]), label_pos = 0.45, label_offset = c(0.06, -0.04))
```

```{r, message=FALSE}
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "sim_illness_death.R"))
source(file.path(rprojroot::find_root("DESCRIPTION"), "inst", "simulation", "cut_illness_death.R"))
```

### Transition rates

Transition rates are calibrated from the design assumptions. Control arm medians
are matched numerically; experimental arm rates are derived by scaling with
the specified HRs and solving for the target ORR.

```{r}
transition_rate <- build_transition_rates(
  strata = c("BM+", "BM-"),
  treatments = c("control", "experimental"),
  median_pfs = c("BM+" = pfsmedian, "BM-" = pfsmedian),
  median_os = c("BM+" = osmedian, "BM-" = osmedian),
  orr = list(
    "BM+" = c(control = orr_ctrl_bm_pos, experimental = orr_exp_bm_pos),
    "BM-" = c(control = orr_ctrl_bm_neg, experimental = orr_exp_bm_neg)
  ),
  hr_pfs = c("BM+" = 0.70, "BM-" = 0.90),
  hr_os = c("BM+" = 0.65, "BM-" = 0.85)
)

kable(transition_rate, digits = 4, caption = "Illness-death transition rates") %>%
  kable_styling()
```

### Theoretical survival curves

The illness-death model implies marginal PFS and OS distributions that are mixtures of
exponential and convolution-of-exponential densities. We plot these theoretical curves
to visualize the assumed treatment effects before simulating any data.

```{r}
# Compute theoretical survival for a given stratum and treatment
get_theoretical_surv <- function(transition_rate, stratum_val, trt_val, t_grid) {
  tr <- transition_rate[transition_rate$stratum == stratum_val &
                          transition_rate$treatment == trt_val, ]
  get_rate <- function(trans) tr$rate[tr$transition == trans]

  lambda_resp <- get_rate("response")
  lambda_prog0 <- get_rate("prog_0")
  lambda_death0 <- get_rate("death_0")
  lambda_prog1 <- get_rate("prog_1")
  lambda_death1 <- get_rate("death_1")
  lambda_death2 <- get_rate("death_2")

  lambda_0 <- lambda_resp + lambda_prog0 + lambda_death0
  orr_param <- lambda_resp / lambda_0
  lambda_1 <- lambda_prog1 + lambda_death1

  pfs_surv <- 1 - sapply(t_grid, function(t)
    .pfs_cdf(t, lambda_0, orr_param, lambda_1))
  os_surv <- 1 - sapply(t_grid, function(t)
    .os_cdf(t, lambda_0, orr_param, lambda_1,
            lambda_death0, lambda_prog0, lambda_death1,
            lambda_prog1, lambda_death2))

  tibble(
    time = rep(t_grid, 2),
    surv = c(pfs_surv, os_surv),
    endpoint = rep(c("PFS", "OS"), each = length(t_grid)),
    stratum = stratum_val,
    treatment = trt_val
  )
}

t_grid <- seq(0, 38, by = 0.25)  # Limit to trial duration (final analysis at 38 months)

# By-stratum curves (BM+ and BM-)
stratum_curves <- bind_rows(
  get_theoretical_surv(transition_rate, "BM+", "control", t_grid),
  get_theoretical_surv(transition_rate, "BM+", "experimental", t_grid),
  get_theoretical_surv(transition_rate, "BM-", "control", t_grid),
  get_theoretical_surv(transition_rate, "BM-", "experimental", t_grid)
) %>% mutate(population = "By stratum")

# Overall population: weighted average of BM+ and BM- survival
bm_pos_curves <- bind_rows(
  get_theoretical_surv(transition_rate, "BM+", "control", t_grid),
  get_theoretical_surv(transition_rate, "BM+", "experimental", t_grid)
)
bm_neg_curves <- bind_rows(
  get_theoretical_surv(transition_rate, "BM-", "control", t_grid),
  get_theoretical_surv(transition_rate, "BM-", "experimental", t_grid)
)
overall_curves <- bm_pos_curves %>%
  select(time, endpoint, treatment, surv_pos = surv) %>%
  left_join(
    bm_neg_curves %>% select(time, endpoint, treatment, surv_neg = surv),
    by = c("time", "endpoint", "treatment")
  ) %>%
  mutate(
    surv = bm_prev * surv_pos + (1 - bm_prev) * surv_neg,
    stratum = "Overall",
    population = "Overall"
  ) %>%
  select(time, surv, endpoint, stratum, treatment, population)

theor_curves <- bind_rows(stratum_curves, overall_curves)
```

```{r, fig.width=10, fig.asp=1}
theor_curves$endpoint <- factor(theor_curves$endpoint, levels = c("PFS", "OS"))
theor_curves$population <- factor(theor_curves$population,
                                   levels = c("By stratum", "Overall"))

# Median labels: by stratum row shows BM+/BM- medians, overall row shows overall medians
median_tbl <- theor_curves %>%
  group_by(endpoint, population, stratum, treatment) %>%
  summarise(median = time[which.min(abs(surv - 0.5))], .groups = "drop") %>%
  mutate(label = sprintf("%s %s: %.1f mo", stratum, treatment, median))

median_labels <- median_tbl %>%
  group_by(endpoint, population) %>%
  summarise(label = paste(label, collapse = "\n"), .groups = "drop")
median_labels$endpoint <- factor(median_labels$endpoint, levels = c("PFS", "OS"))
median_labels$population <- factor(median_labels$population,
                                    levels = c("By stratum", "Overall"))

ggplot(theor_curves, aes(x = time, y = surv,
                          color = treatment, linetype = stratum)) +
  geom_line(linewidth = 0.8) +
  facet_grid(population ~ endpoint) +
  scale_color_manual(values = trt_colors, name = "Treatment") +
  scale_linetype_manual(values = c("BM+" = "solid", "BM-" = "dashed", "Overall" = "solid"),
                        name = "Stratum") +
  scale_x_continuous(breaks = seq(0, 36, by = 6)) +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, 1, by = 0.2), limits = c(0, 1)) +
  geom_text(data = median_labels,
            aes(x = 20, y = 0.95, label = label),
            inherit.aes = FALSE, hjust = 0, vjust = 1, size = 2.8,
            lineheight = 0.9) +
  labs(title = "Theoretical survival from illness-death model",
       x = "Time (months)", y = "Survival probability") +
  theme_bw() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold", size = 12))
```

### Simulate one trial

```{r}
set.seed(20250313)
n_per_arm <- ceiling(n_bm_pos + n_bm_neg)
n_sim <- n_per_arm * 2
steady_rate <- n_sim / 11  # 14-month enrollment: 0.25*2 + 0.5*2 + 0.75*2 + 1*8 = 11 rate-months

sim_data <- sim_illness_death(
  n = n_sim,
  stratum = data.frame(stratum = c("BM+", "BM-"), p = c(bm_prev, 1 - bm_prev)),
  block = c("control", "control", "experimental", "experimental"),
  enroll_rate = data.frame(
    rate = steady_rate * c(0.25, 0.50, 0.75, 1.00),
    duration = c(2, 2, 2, 8)
  ),
  transition_rate = transition_rate,
  dropout_rate = 0.001
)

fpe <- max(sim_data$ENRLTIME)
cat("Total N:", nrow(sim_data), "\n")
cat("N BM+:", sum(sim_data$STRATUM == "BM+"),
    "N BM-:", sum(sim_data$STRATUM == "BM-"), "\n")
cat("FPE:", round(fpe, 1), "months\n")
```

### Determine analysis cut dates

```{r}
# Extract target event counts from designs
target_pfs_events_sub <- ceiling(max(pfssub$analysis$event))
target_os_events_sub <- ceiling(max(ossub$analysis$event))

analyses <- list(
  list(min_followup = 6, endpoint = NULL, event_target = NULL,
       target_stratum = NULL, max_followup = NULL),
  list(min_followup = 14, endpoint = "PFS", event_target = target_pfs_events_sub,
       target_stratum = "BM+", max_followup = 17),
  list(min_followup = 24, endpoint = "OS", event_target = target_os_events_sub,
       target_stratum = "BM+", max_followup = 30)
)

cut_dates <- get_analysis_dates(sim_data, analyses)

cat("Analysis cut dates:\n")
for (i in seq_along(cut_dates)) {
  cat(sprintf("  Analysis %d: %.1f months (FPE + %.1f)\n",
              i, cut_dates[i], cut_dates[i] - fpe))
}
```

### Cut data and compute event counts

```{r}
adtte <- lapply(seq_along(cut_dates), function(i) {
  d <- cut_illness_death(sim_data, cut_dates[i])
  d$ANALYSIS <- i
  d
})
names(adtte) <- paste0("IA", seq_along(cut_dates))

# Event counts per analysis
event_summary <- do.call(rbind, lapply(seq_along(adtte), function(i) {
  adtte[[i]] %>%
    filter(PARAMCD %in% c("OS", "PFS")) %>%
    group_by(PARAMCD, STRATUM) %>%
    summarise(events = sum(CNSR == 0), .groups = "drop") %>%
    mutate(Analysis = i)
}))

kable(event_summary %>% tidyr::pivot_wider(
  names_from = Analysis, values_from = events, names_prefix = "Analysis "
), caption = "Event counts by analysis") %>%
  kable_styling()
```

### Compute nominal p-values {.tabset}

One-sided p-values from the logrank test (OS, PFS) and asymptotic rate difference test (ORR).
For overall population hypotheses, analyses are stratified by BM+/BM-.

```{r}
# Helper: 1-sided logrank p-value (H1: experimental better)
logrank_pval <- function(data, stratified = FALSE) {
  data$event <- 1L - data$CNSR
  if (stratified) {
    fit <- survdiff(Surv(AVAL, event) ~ TRT + strata(STRATUM), data = data)
  } else {
    fit <- survdiff(Surv(AVAL, event) ~ TRT, data = data)
  }
  # survdiff orders groups alphabetically: "control" before "experimental"
  # With strata, obs/exp are matrices (groups x strata); sum across strata
  obs_ctrl <- if (is.matrix(fit$obs)) sum(fit$obs[1, ]) else fit$obs[1]
  exp_ctrl <- if (is.matrix(fit$exp)) sum(fit$exp[1, ]) else fit$exp[1]
  # Observed - Expected > 0 for control means experimental has fewer events (better)
  z_sign <- sign(obs_ctrl - exp_ctrl)

  z <- z_sign * sqrt(fit$chisq)
  pnorm(-z)  # 1-sided: small when experimental is better
}

# Helper: 1-sided p-value for rate difference (H1: p_exp > p_ctrl)
# If stratified = TRUE, combines stratum-specific rate differences
# using sample size (Cochran-Mantel-Haenszel) weights
rd_pval <- function(data, stratified = FALSE) {
  if (!stratified) {
    n_exp <- sum(data$TRT == "experimental")
    n_ctrl <- sum(data$TRT == "control")
    resp_exp <- sum(data$AVAL[data$TRT == "experimental"])
    resp_ctrl <- sum(data$AVAL[data$TRT == "control"])
    p_exp <- resp_exp / n_exp
    p_ctrl <- resp_ctrl / n_ctrl
    p_pool <- (resp_exp + resp_ctrl) / (n_exp + n_ctrl)
    se <- sqrt(p_pool * (1 - p_pool) * (1 / n_exp + 1 / n_ctrl))
    if (se == 0) return(0.5)
    z <- (p_exp - p_ctrl) / se
    return(pnorm(-z))  # 1-sided: small when experimental is better
  }

  # Stratified risk difference with sample size weights
  strata <- unique(data$STRATUM)
  n_s <- rd_s <- var_s <- numeric(length(strata))
  for (k in seq_along(strata)) {
    ds <- data[data$STRATUM == strata[k], ]
    n_exp <- sum(ds$TRT == "experimental")
    n_ctrl <- sum(ds$TRT == "control")
    n_s[k] <- n_exp + n_ctrl
    p_exp <- sum(ds$AVAL[ds$TRT == "experimental"]) / n_exp
    p_ctrl <- sum(ds$AVAL[ds$TRT == "control"]) / n_ctrl
    rd_s[k] <- p_exp - p_ctrl
    var_s[k] <- p_exp * (1 - p_exp) / n_exp + p_ctrl * (1 - p_ctrl) / n_ctrl
  }
  w <- n_s / sum(n_s)  # sample size weights
  rd_combined <- sum(w * rd_s)
  se_combined <- sqrt(sum(w^2 * var_s))
  if (se_combined == 0) return(0.5)
  z <- rd_combined / se_combined
  pnorm(-z)  # 1-sided: small when experimental is better
}
```

```{r}
# Compute p-values for each hypothesis at each analysis
pvals <- list()

for (i in seq_along(cut_dates)) {
  d <- adtte[[i]]

  # H1: OS BM+ (analyses 1-3) — unstratified logrank
  pvals[[paste0("H1_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS", STRATUM == "BM+"))

  # H2: OS Overall (analyses 1-3) — stratified logrank
  pvals[[paste0("H2_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "OS"), stratified = TRUE)

  # H3: PFS BM+ (analyses 1-2) — unstratified logrank
  if (i <= 2) {
    pvals[[paste0("H3_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS", STRATUM == "BM+"))
  }

  # H4: PFS Overall (analyses 1-2) — stratified logrank
  if (i <= 2) {
    pvals[[paste0("H4_", i)]] <- logrank_pval(d %>% filter(PARAMCD == "PFS"), stratified = TRUE)
  }

  # H5: ORR BM+ (analysis 1 only) — unstratified rate difference
  if (i == 1) {
    pvals[[paste0("H5_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR", STRATUM == "BM+"))
  }

  # H6: ORR Overall (analysis 1 only) — stratified rate difference
  if (i == 1) {
    pvals[[paste0("H6_", i)]] <- rd_pval(d %>% filter(PARAMCD == "ORR"), stratified = TRUE)
  }
}
```

### Build results table

Assemble event counts and p-values into the format required for sequential testing.
Spending time uses the BM+ information fraction for all hypotheses.

```{r, warnings=FALSE, message=FALSE}
# Extract event counts from ADTTE
get_events <- function(analysis_idx, paramcd, stratum = NULL) {
  d <- adtte[[analysis_idx]]
  if (!is.null(stratum)) d <- d %>% filter(STRATUM == stratum)
  sum(d$PARAMCD == paramcd & d$CNSR == 0, na.rm = TRUE)
}

events_os_sub <- sapply(1:3, function(i) get_events(i, "OS", "BM+"))
events_os_all <- sapply(1:3, function(i) get_events(i, "OS"))
events_pfs_sub <- sapply(1:2, function(i) get_events(i, "PFS", "BM+"))
events_pfs_all <- sapply(1:2, function(i) get_events(i, "PFS"))

inputResults <- tibble(
  H = c(rep(1, 3), rep(2, 3), rep(3, 2), rep(4, 2), 5, 6),
  Pop = c(
    rep("BM+", 3), rep("Overall", 3),
    rep("BM+", 2), rep("Overall", 2),
    "BM+", "Overall"
  ),
  Endpoint = c(rep("OS", 6), rep("PFS", 4), rep("ORR", 2)),
  nominalP = c(
    pvals$H1_1, pvals$H1_2, pvals$H1_3,
    pvals$H2_1, pvals$H2_2, pvals$H2_3,
    pvals$H3_1, pvals$H3_2,
    pvals$H4_1, pvals$H4_2,
    pvals$H5_1,
    pvals$H6_1
  ),
  Analysis = c(1:3, 1:3, 1:2, 1:2, 1, 1),
  events = c(events_os_sub, events_os_all, events_pfs_sub, events_pfs_all, NA, NA),
  spendingTime = c(
    events_os_sub / max(events_os_sub),
    events_os_sub / max(events_os_sub),
    events_pfs_sub / max(events_pfs_sub),
    events_pfs_sub / max(events_pfs_sub),
    NA, NA
  )
)
kable(inputResults, caption = "Results from simulated trial data") %>%
  kable_styling()
```

### Graphical summaries {.tabset}

```{r, include=FALSE}
# Helper: Kaplan-Meier plot with HR, 95% CI, logrank p-value, and at-risk table
km_plot <- function(data, title, stratified = FALSE) {
  data$event <- 1L - data$CNSR
  fit <- survfit(Surv(AVAL, event) ~ TRT, data = data)

  # HR and 95% CI from Cox model
  if (stratified) {
    cox_fit <- coxph(Surv(AVAL, event) ~ TRT + strata(STRATUM), data = data)
  } else {
    cox_fit <- coxph(Surv(AVAL, event) ~ TRT, data = data)
  }
  hr <- exp(coef(cox_fit))
  ci <- exp(confint(cox_fit))

  # Logrank p-value
  pval <- logrank_pval(data, stratified = stratified)

  # Build annotation
  hr_text <- sprintf(
    "HR = %.3f (95%% CI: %.3f, %.3f)\n1-sided logrank p = %.4f",
    hr, ci[1], ci[2], pval
  )

  # Convert survfit to data frame for ggplot
  sdf <- data.frame(
    time = fit$time,
    surv = fit$surv,
    upper = fit$upper,
    lower = fit$lower,
    strata = rep(names(fit$strata), fit$strata)
  )
  sdf$TRT <- gsub("TRT=", "", sdf$strata)

  # X-axis breaks every 6 months
  max_time <- max(sdf$time)
  x_breaks <- seq(0, ceiling(max_time / 6) * 6, by = 6)

  # At-risk table: number at risk at each x_break
  trt_levels <- c("control", "experimental")
  at_risk <- expand.grid(time = x_breaks, TRT = trt_levels,
                         stringsAsFactors = FALSE)
  at_risk$n_risk <- mapply(function(t, arm) {
    d_arm <- data[data$TRT == arm, ]
    sum(d_arm$AVAL >= t)
  }, at_risk$time, at_risk$TRT)

  # Main KM plot
  p_km <- ggplot(sdf, aes(x = time, y = surv, color = TRT)) +
    geom_step(linewidth = 0.8) +
    scale_color_manual(values = trt_colors, name = "Treatment") +
    scale_x_continuous(breaks = x_breaks, limits = c(0, max(x_breaks))) +
    scale_y_continuous(labels = scales::percent,
                       breaks = seq(0, 1, by = 0.2), limits = c(0, 1)) +
    labs(title = title, x = NULL, y = "Survival probability") +
    annotate("text", x = max(x_breaks) * 0.98, y = 0.95,
             label = hr_text, hjust = 1, vjust = 1, size = 3.2) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 5.5, 0, 5.5)
    )

  # At-risk table as a plot
  p_risk <- ggplot(at_risk, aes(x = time, y = TRT, label = n_risk, color = TRT)) +
    geom_text(size = 3) +
    scale_color_manual(values = trt_colors, guide = "none") +
    scale_x_continuous(breaks = x_breaks, limits = c(0, max(x_breaks))) +
    scale_y_discrete(limits = rev(trt_levels)) +
    labs(x = "Time (months)", y = NULL) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      panel.border = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = margin(0, 5.5, 5.5, 5.5)
    )

  # Stack KM plot and at-risk table
  gridExtra::grid.arrange(p_km, p_risk, ncol = 1, heights = c(4, 1))
}

# Helper: ORR bar chart for all populations
orr_plot <- function(data_bm_pos, data_all, data_bm_neg) {

  get_rates <- function(d, pop_label) {
    resp_ctrl <- sum(d$AVAL[d$TRT == "control"])
    n_ctrl <- sum(d$TRT == "control")
    resp_exp <- sum(d$AVAL[d$TRT == "experimental"])
    n_exp <- sum(d$TRT == "experimental")
    tibble(
      Population = pop_label,
      Treatment = c("control", "experimental"),
      Responses = c(resp_ctrl, resp_exp),
      N = c(n_ctrl, n_exp),
      Rate = c(resp_ctrl / n_ctrl, resp_exp / n_exp)
    )
  }

  rates <- bind_rows(
    get_rates(data_bm_pos, "BM+"),
    get_rates(data_all, "Overall"),
    get_rates(data_bm_neg, "BM-")
  )
  rates$Population <- factor(rates$Population, levels = c("BM+", "Overall", "BM-"))

  # Compute risk differences and p-values per population
  rd_stats <- rates %>%
    group_by(Population) %>%
    summarise(
      rd = Rate[Treatment == "experimental"] - Rate[Treatment == "control"],
      rate_exp = Rate[Treatment == "experimental"],
      rate_ctrl = Rate[Treatment == "control"],
      .groups = "drop"
    )

  pval_bm_pos <- rd_pval(data_bm_pos, stratified = FALSE)
  pval_overall <- rd_pval(data_all, stratified = TRUE)
  pval_bm_neg <- rd_pval(data_bm_neg, stratified = FALSE)
  rd_stats$pval <- c(pval_bm_pos, pval_overall, pval_bm_neg)

  rd_stats$label <- sprintf(
    "RD = %.1f%%\n1-sided p = %.4f",
    rd_stats$rd * 100, rd_stats$pval
  )

  ggplot(rates, aes(x = Population, y = Rate, fill = Treatment)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(
      aes(label = sprintf("%.1f%%", Rate * 100)),
      position = position_dodge(width = 0.7),
      vjust = -0.5, size = 3
    ) +
    geom_text(
      data = rd_stats,
      aes(x = Population, y = pmax(rate_exp, rate_ctrl) + 0.08,
          label = label, fill = NULL),
      size = 3, lineheight = 0.9
    ) +
    scale_fill_manual(values = trt_colors, name = "Treatment") +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1),
                       expand = expansion(mult = c(0, 0.05))) +
    labs(title = "Objective Response Rate by Population",
         x = NULL, y = "Response rate") +
    theme_bw() +
    theme(legend.position = "bottom")
}
```

#### ORR (Analysis 1)

```{r, fig.asp=0.7}
d1 <- adtte[[1]]
orr_plot(
  data_bm_pos = d1 %>% filter(PARAMCD == "ORR", STRATUM == "BM+"),
  data_all = d1 %>% filter(PARAMCD == "ORR"),
  data_bm_neg = d1 %>% filter(PARAMCD == "ORR", STRATUM == "BM-")
)
```

```{r, include=FALSE}
# Pre-compute all KM plot objects
# Ordering: Analysis first, then Endpoint, then Population
endpoints <- c("OS", "PFS")
populations <- list(
  "BM+" = list(filter_expr = "BM+", stratified = FALSE),
  "Overall (stratified)" = list(filter_expr = NULL, stratified = TRUE),
  "BM-" = list(filter_expr = "BM-", stratified = FALSE)
)

km_plots <- list()
for (ai in 1:3) {
  for (ep in endpoints) {
    n_analyses <- if (ep == "OS") 3 else 2
    if (ai > n_analyses) next
    for (pop_name in names(populations)) {
      pop <- populations[[pop_name]]
      d <- adtte[[ai]]
      d_ep <- d %>% filter(PARAMCD == ep)
      if (!is.null(pop$filter_expr)) {
        d_ep <- d_ep %>% filter(STRATUM == pop$filter_expr)
      }
      title <- sprintf("%s \u2014 Analysis %d \u2014 %s (N = %d, events = %d)",
                        ep, ai, pop_name, nrow(d_ep), sum(d_ep$CNSR == 0))
      key <- sprintf("A%d_%s_%s", ai, ep, gsub(" ", "_", pop_name))
      km_plots[[key]] <- km_plot(d_ep, title, stratified = pop$stratified)
    }
  }
}
```

```{r, results='asis', fig.asp=0.85}
# Generate tabs: Analysis 1 first, then 2, then 3
for (ai in 1:3) {
  for (ep in endpoints) {
    n_analyses <- if (ep == "OS") 3 else 2
    if (ai > n_analyses) next
    for (pop_name in names(populations)) {
      key <- sprintf("A%d_%s_%s", ai, ep, gsub(" ", "_", pop_name))
      cat(" \n")
      cat(sprintf("#### %s A%d %s \n\n", ep, ai, pop_name))
      print(km_plots[[key]])
      cat(" \n\n\n")
    }
  }
}
```

## Testing hypotheses

### Compute sequential p-values for each hypothesis

```{r, message=FALSE}
### USER SHOULD NOT NEED TO MODIFY THIS CODE
EOCtab <- NULL
EOCtab <- inputResults %>%
  group_by(H) %>%
  slice(1) %>%
  ungroup() %>%
  select("H", "Pop", "Endpoint", "nominalP")
EOCtab$seqp <- .9999
for (EOCtabline in 1:nHypotheses) {
  EOCtab$seqp[EOCtabline] <-
    ifelse(is.null(gsD2list[[EOCtabline]]), EOCtab$nominalP[EOCtabline], {
      tem <- filter(inputResults, H == EOCtabline)
      gsDesign2::sequential_pval(
        gs_design = gsD2list[[EOCtabline]],
        event = tem$events,
        z = -stats::qnorm(tem$nominalP),
        ustime = tem$spendingTime,
        interval = c(1e-05, 0.9999)
      )
    })
}
EOCtab <- EOCtab %>% select(-"nominalP")
```

### Evaluate hypothesis rejection using graphicalMCP

```{r, message=FALSE, warning=FALSE}
library(graphicalMCP)

g0 <- graphicalMCP::graph_create(
  hypotheses = alphaHypotheses / fwer,
  transitions = m,
  hyp_names = nameHypotheses
)
```

```{r}
result <- graphicalMCP::graph_test_shortcut(
  graph = g0,
  p = EOCtab$seqp,
  alpha = fwer,
  verbose = TRUE
)

adj_p <- result$outputs$adjusted_p
rej <- result$outputs$rejected

if (is.numeric(rej)) {
  rej_logical <- rep(FALSE, nHypotheses)
  rej_logical[rej] <- TRUE
} else {
  rej_logical <- as.logical(rej)
}

EOCtab$Rejected <- rej_logical
EOCtab$adjPValues <- adj_p
```

## Verification of hypotheses rejected

```{r}
rejected_order <- which(EOCtab$Rejected)[order(EOCtab$adjPValues[EOCtab$Rejected])]

graphs <- list(g0)
if (length(rejected_order) > 0) {
  gu <- graphicalMCP::graph_update(g0, delete = rejected_order)
  graphs <- gu$intermediate_graphs
}
ngraphs <- length(graphs)

lastWeights <- as.numeric(graphs[[ngraphs]]$hypotheses)
lastGraph <- rep(ngraphs, nrow(EOCtab))

if (length(rejected_order) > 0) {
  for (j in seq_along(rejected_order)) {
    h <- rejected_order[j]
    lastWeights[h] <- as.numeric(graphs[[j]]$hypotheses[h])
    lastGraph[h] <- j
  }
}

EOCtab$lastAlpha <- fwer * lastWeights
EOCtab$lastGraph <- lastGraph
EOCtabx <- EOCtab
names(EOCtabx) <- c(
  "Hypothesis", "Population", "Endpoint", "Sequential p",
  "Rejected", "Adjusted p", "Max alpha allocated", "Last Graph"
)
```

### Multiplicity graph sequence from graphicalMCP {.tabset}

```{r, message=FALSE, warning=FALSE, results='asis', fig.asp=0.95}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
for (i in 1:ngraphs) {
  cat(" \n")
  cat("####", paste(" Graph", as.character(i), " \n\n"))
  gi_alpha <- fwer * as.numeric(graphs[[i]]$hypotheses)
  gi_labels <- paste(
    names(graphs[[i]]$hypotheses),
    sprintf("\u03b1 = %s", format(gi_alpha, scientific = FALSE)),
    sep = "\n"
  )
  plot(graphs[[i]],
    layout = layout6,
    vertex.size = 45,
    vertex.label = gi_labels,
    vertex.label.cex = 0.9,
    vertex.color = unname(vertex_color_map[names(graphs[[i]]$hypotheses)]),
    margin = 0.25
  )
  cat(" \n\n\n")
}
```

### Comparison of sequential p-values to multiplicity graphs

```{r}
EOCtabx %>%
  select(c(1:4, 7, 5:6, 8)) %>%
  kable() %>%
  kable_styling()
```

### Bounds at final alpha allocated for group sequential tests {.tabset}

```{r, results='asis'}
for (i in 1:nHypotheses) {
  cat("####", paste(" Hypothesis", as.character(i), " \n"))
  hresults <- inputResults %>% filter(H == i)
  xx <- paste("Max alpha allocated from above table: ",
    as.character(EOCtab$lastAlpha[i]),
    sep = ""
  )
  d2 <- gsD2list[[i]]
  if (is.null(d2)) {
    cat("Maximum alpha allocated: ")
    cat(EOCtab$lastAlpha[i])
    cat("\n\n")
    cat("Nominal p-value for hypothesis test: ")
    cat(hresults$nominalP)
    cat("\n\n")
  }
  if (!is.null(gsD2list[[i]])) {
    cat("Nominal p-values at each analysis for comparison to bounds in table below:\n\n")
    cat(hresults$nominalP)
    cat("\n\n")
    n.I <- hresults$events
    usTime <- hresults$spendingTime
    if (EOCtab$lastAlpha[i] == 0) {
      cat("Maximum alpha allocated: 0\n\n")
      cat("No testing required\n\n")
    }
    if (EOCtab$lastAlpha[i] > 0) {
      d2_upd <- gsDesign2::gs_update_ahr(
        x = d2,
        alpha = EOCtab$lastAlpha[i],
        ustime = usTime,
        event_tbl = data.frame(analysis = hresults$Analysis, event = n.I)
      )

      d2_upd$bound %>%
        dplyr::filter(.data$bound == "upper") %>%
        dplyr::left_join(
          data.frame(analysis = hresults$Analysis, spending_time = usTime),
          by = "analysis"
        ) %>%
        dplyr::transmute(
          Analysis = .data$analysis,
          Z = .data$z,
          `Nominal p (1-sided)` = .data$`nominal p`,
          `Spending time` = .data$spending_time
        ) %>%
        kable(caption = xx, row.names = FALSE) %>%
        kable_styling() %>%
        cat()
      cat("\n\n")
    }
  }
}
```

```{r}
options(old)
```

## Weighted parametric group sequential design (wpgsd)

The analysis above uses the Bonferroni-based shortcut (via `graphicalMCP::graph_test_shortcut()`)
which ignores correlations between test statistics.
The [wpgsd](https://merck.github.io/wpgsd/) package (Anderson et al., 2022)
accounts for two sources of correlation:

1. **Group sequential correlation** — between interim and final analyses of the same hypothesis,
   driven by the information fraction.
2. **Population-induced correlation** — between subgroup (BM+) and overall population tests for the
   same endpoint, because BM+ patients are a subset of the overall population.

We apply wpgsd separately for OS (H1, H2; 3 analyses) and PFS (H3, H4; 2 analyses).

```{r, message=FALSE}
library(wpgsd)
```

### Correlation matrices

The correlation between test statistics arises from two sources:

- **Group sequential**: $\text{corr}(Z_{H_i,A_s}, Z_{H_i,A_t}) = \sqrt{d_{i,\min(s,t)} / d_{i,\max(s,t)}}$
  where $d_{i,k}$ is the event count for hypothesis $i$ at analysis $k$.
- **Nested populations**: $\text{corr}(Z_{H_{\text{sub}},A_s}, Z_{H_{\text{all}},A_t}) = d_{\text{sub},\min(s,t)} / \sqrt{d_{\text{sub},s} \cdot d_{\text{all},t}}$
  since the intersection of BM+ and Overall events equals the BM+ event count.

We construct the event-count matrix $D$ where $D_{ij}$ contains the number of events shared
between test statistics $i$ and $j$, then convert to correlation via
$\text{corr} = B \, D \, B$ where $B = \text{diag}(1/\sqrt{d_{ii}})$.

For PFS (2 analyses), we use `wpgsd::generate_corr()` directly.
For OS (3 analyses), we construct the correlation matrix manually.

```{r}
# PFS: use generate_corr (works correctly for k = 2)
event_pfs <- tibble::tribble(
  ~H1, ~H2, ~Analysis, ~Event,
  # Analysis 1
  1, 1, 1, events_pfs_sub[1],
  2, 2, 1, events_pfs_all[1],
  1, 2, 1, events_pfs_sub[1],
  # Analysis 2
  1, 1, 2, events_pfs_sub[2],
  2, 2, 2, events_pfs_all[2],
  1, 2, 2, events_pfs_sub[2]
)

corr_pfs <- generate_corr(event_pfs)
```

```{r}
# OS: build 6x6 correlation matrix manually (2 hypotheses x 3 analyses)
# Ordering: H1_A1, H2_A1, H1_A2, H2_A2, H1_A3, H2_A3
d_os <- c(
  events_os_sub[1], events_os_all[1],
  events_os_sub[2], events_os_all[2],
  events_os_sub[3], events_os_all[3]
)
D_os <- diag(d_os)

# Within-hypothesis across-analysis: events at min(s,t)
# H1 (BM+) at positions 1, 3, 5
D_os[1, 3] <- D_os[3, 1] <- events_os_sub[1]  # A1<->A2
D_os[1, 5] <- D_os[5, 1] <- events_os_sub[1]  # A1<->A3
D_os[3, 5] <- D_os[5, 3] <- events_os_sub[2]  # A2<->A3
# H2 (Overall) at positions 2, 4, 6
D_os[2, 4] <- D_os[4, 2] <- events_os_all[1]
D_os[2, 6] <- D_os[6, 2] <- events_os_all[1]
D_os[4, 6] <- D_os[6, 4] <- events_os_all[2]

# Cross-hypothesis: intersection at min(s,t) = BM+ events at min(s,t)
# Same analysis
D_os[1, 2] <- D_os[2, 1] <- events_os_sub[1]
D_os[3, 4] <- D_os[4, 3] <- events_os_sub[2]
D_os[5, 6] <- D_os[6, 5] <- events_os_sub[3]
# Cross-analysis
D_os[1, 4] <- D_os[4, 1] <- events_os_sub[1]  # H1_A1<->H2_A2
D_os[1, 6] <- D_os[6, 1] <- events_os_sub[1]  # H1_A1<->H2_A3
D_os[2, 3] <- D_os[3, 2] <- events_os_sub[1]  # H2_A1<->H1_A2
D_os[2, 5] <- D_os[5, 2] <- events_os_sub[1]  # H2_A1<->H1_A3
D_os[3, 6] <- D_os[6, 3] <- events_os_sub[2]  # H1_A2<->H2_A3
D_os[4, 5] <- D_os[5, 4] <- events_os_sub[2]  # H2_A2<->H1_A3

B_os <- diag(1 / sqrt(diag(D_os)))
corr_os <- B_os %*% D_os %*% B_os
colnames(corr_os) <- paste0(
  rep(c("H1", "H2"), 3), "_A", rep(1:3, each = 2)
)
```

The OS correlation matrix is $6 \times 6$ (2 hypotheses $\times$ 3 analyses)
and the PFS correlation matrix is $4 \times 4$ (2 hypotheses $\times$ 2 analyses).
Off-diagonal blocks capture the population-induced correlation from the nested BM+ $\subset$ Overall structure.

```{r}
cat("OS correlation matrix (H1: BM+, H2: Overall; Analyses 1-3):\n")
round(corr_os, 3)

cat("\nPFS correlation matrix (H3: BM+, H4: Overall; Analyses 1-2):\n")
round(corr_pfs, 3)
```

### WPGSD bounds vs Bonferroni bounds {.tabset}

We compare bounds from the weighted parametric approach (which uses the correlation) to
standard Bonferroni bounds (which ignore it). For each endpoint, we use a 2-hypothesis
sub-graph with full reallocation between BM+ and Overall, and Lan-DeMets O'Brien-Fleming
spending (`sfLDOF`).

```{r}
# Sub-graph for 2 hypotheses: full reallocation
m_sub <- matrix(c(0, 1, 1, 0), nrow = 2, byrow = TRUE)

# --- OS ---
# Weights proportional to initial alpha for H1, H2
w_os <- alphaHypotheses[1:2] / sum(alphaHypotheses[1:2])

# Information fractions (spending time) for OS
t_os_sub <- events_os_sub / max(events_os_sub)
t_os_all <- events_os_all / max(events_os_all)

# Bonferroni bounds (type = 0)
bound_os_bonf <- generate_bounds(
  type = 0, k = 3, w = w_os, m = m_sub, corr = corr_os,
  alpha = sum(alphaHypotheses[1:2]),
  sf = list(gsDesign::sfLDOF, gsDesign::sfLDOF),
  sfparm = list(0, 0),
  t = list(t_os_sub, t_os_sub)
)

# WPGSD bounds (type = 3: separate alpha spending with parametric adjustment)
bound_os_wpgsd <- generate_bounds(
  type = 3, k = 3, w = w_os, m = m_sub, corr = corr_os,
  alpha = sum(alphaHypotheses[1:2]),
  sf = list(gsDesign::sfLDOF, gsDesign::sfLDOF),
  sfparm = list(0, 0),
  t = list(t_os_sub, t_os_sub)
)
```

```{r}
# --- PFS ---
# Use equal weights for the PFS sub-graph (both hypotheses get positive alpha)
# The actual alpha allocation from the full graph is handled separately
w_pfs <- c(0.5, 0.5)

t_pfs_sub <- events_pfs_sub / max(events_pfs_sub)
t_pfs_all <- events_pfs_all / max(events_pfs_all)

# Bonferroni bounds
bound_pfs_bonf <- generate_bounds(
  type = 0, k = 2, w = w_pfs, m = m_sub, corr = corr_pfs,
  alpha = sum(alphaHypotheses[3:4]),
  sf = list(gsDesign::sfLDOF, gsDesign::sfLDOF),
  sfparm = list(0, 0),
  t = list(t_pfs_sub, t_pfs_sub)
)

# WPGSD bounds
bound_pfs_wpgsd <- generate_bounds(
  type = 3, k = 2, w = w_pfs, m = m_sub, corr = corr_pfs,
  alpha = sum(alphaHypotheses[3:4]),
  sf = list(gsDesign::sfLDOF, gsDesign::sfLDOF),
  sfparm = list(0, 0),
  t = list(t_pfs_sub, t_pfs_sub)
)
```

#### OS bounds comparison

```{r}
# Combine Bonferroni and WPGSD bounds for OS
bounds_os <- dplyr::left_join(
  bound_os_bonf, bound_os_wpgsd,
  by = c("Hypotheses", "Analysis"),
  suffix = c(".Bonf", ".WPGSD")
)

# Convert to Z-scale for easier interpretation
bounds_os <- bounds_os %>%
  dplyr::mutate(
    `Z H1 Bonf` = round(-qnorm(H1.Bonf), 3),
    `Z H2 Bonf` = round(-qnorm(H2.Bonf), 3),
    `Z H1 WPGSD` = round(-qnorm(H1.WPGSD), 3),
    `Z H2 WPGSD` = round(-qnorm(H2.WPGSD), 3)
  )

bounds_os %>%
  dplyr::select(Analysis, Hypotheses,
                `Z H1 Bonf`, `Z H1 WPGSD`,
                `Z H2 Bonf`, `Z H2 WPGSD`) %>%
  kable(caption = "OS bounds: Bonferroni vs WPGSD (Z-scale, lower = less conservative)") %>%
  kable_styling()
```

#### PFS bounds comparison

```{r}
bounds_pfs <- dplyr::left_join(
  bound_pfs_bonf, bound_pfs_wpgsd,
  by = c("Hypotheses", "Analysis"),
  suffix = c(".Bonf", ".WPGSD")
)

bounds_pfs <- bounds_pfs %>%
  dplyr::mutate(
    `Z H3 Bonf` = round(-qnorm(H1.Bonf), 3),
    `Z H4 Bonf` = round(-qnorm(H2.Bonf), 3),
    `Z H3 WPGSD` = round(-qnorm(H1.WPGSD), 3),
    `Z H4 WPGSD` = round(-qnorm(H2.WPGSD), 3)
  )

bounds_pfs %>%
  dplyr::select(Analysis, Hypotheses,
                `Z H3 Bonf`, `Z H3 WPGSD`,
                `Z H4 Bonf`, `Z H4 WPGSD`) %>%
  kable(caption = "PFS bounds: Bonferroni vs WPGSD (Z-scale, lower = less conservative)") %>%
  kable_styling()
```

### Closed testing with WPGSD bounds

We use `closed_test()` to formally test the observed nominal p-values against the WPGSD bounds.

```{r}
# Observed p-values for OS (H1, H2) at each analysis
p_obs_os <- tibble::tibble(
  Analysis = 1:3,
  H1 = c(pvals$H1_1, pvals$H1_2, pvals$H1_3),
  H2 = c(pvals$H2_1, pvals$H2_2, pvals$H2_3)
)

# Observed p-values for PFS (H3, H4) at analyses 1-2
p_obs_pfs <- tibble::tibble(
  Analysis = 1:2,
  H1 = c(pvals$H3_1, pvals$H3_2),
  H2 = c(pvals$H4_1, pvals$H4_2)
)

cat("Observed nominal p-values for OS:\n")
kable(p_obs_os, digits = 6) %>% kable_styling()
```

```{r}
cat("Observed nominal p-values for PFS:\n")
kable(p_obs_pfs, digits = 6) %>% kable_styling()
```

```{r}
# Closed testing for OS
ct_os <- closed_test(bound_os_wpgsd, p_obs_os)
kable(ct_os, caption = "WPGSD closed testing results for OS (H1: BM+, H2: Overall)") %>%
  kable_styling()
```

```{r}
# Closed testing for PFS
ct_pfs <- closed_test(bound_pfs_wpgsd, p_obs_pfs)
kable(ct_pfs, caption = "WPGSD closed testing results for PFS (H1: BM+, H2: Overall)") %>%
  kable_styling()
```

The WPGSD approach provides less conservative bounds than the Bonferroni-based approach by
exploiting the known correlation between BM+ and Overall population test statistics.
The Z-boundary tables above show the WPGSD bounds are consistently lower (less conservative)
than the Bonferroni bounds, particularly for the intersection hypotheses.

## Session information

You can use `sessionInfo()` to document the versions of R and R packages used to render this document.

## References

```{js, echo=FALSE}
// Fix dynamic tabsets code block rendering issue under pkgdown
(function () {
  unstyledCode = document.querySelectorAll("button code");
  if (!unstyledCode.length) {
    return;
  }
  for (let i = 0; i < unstyledCode.length; i++) {
    blockOld = unstyledCode[i].outerHTML;
    blockNew =
      "<pre class='downlit sourceCode r' style='text-align:left;white-space:pre-wrap;'>" +
      blockOld +
      "</pre>";
    unstyledCode[i].outerHTML = blockNew;
  }
  var tabElements = [
    "ul#multiplicity-graph-sequence-from-graphicalmcp li:nth-child(2) button",
    "ul#bounds-at-final-alpha-allocated-for-group-sequential-tests li:nth-child(2) button",
  ];
  for (let i = 0; i < tabElements.length; i++) {
    var el = document.querySelector(tabElements[i]);
    var tab = new bootstrap.Tab(el);
    tab.show();
  }
})();
```
