---
title: "New Trial Design: Group sequential design using graphicalMCP and gsDesign2"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
bibliography: gMCPLite.bib
vignette: >
  %\VignetteIndexEntry{New Trial Design: Group sequential design using graphicalMCP and gsDesign2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1,
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)
```

## Overview

This document evaluates statistical significance for graphical multiplicity control with group sequential design following the Maurer-Bretz (2013) framework. The combination of graphicalMCP and gsDesign2 packages is used for design and analysis.

The trial tests 3 endpoints across 2 populations, yielding 6 hypotheses:

| Hypothesis | Endpoint | Population | Analyses |
|------------|----------|------------|----------|
| H1 | OS | BM+ | 3 |
| H2 | OS | Overall | 3 |
| H3 | PFS | BM+ | 2 |
| H4 | PFS | Overall | 2 |
| H5 | ORR | BM+ | 1 |
| H6 | ORR | Overall | 1 |

Key design features:

- Enrollment over 14 months with a 6-month ramp-up.
- BM+ prevalence assumed at 60%.

### Analysis timing rules

All timing is relative to final patient enrolled (FPE). Enrollment is planned over 14 months.

| Analysis | Timing rule (after FPE) | Max extension | Endpoints assessed |
|----------|------------------------|---------------|-------------------|
| IA1 | 6 months after FPE | None | ORR (final for both populations), PFS and OS (interim) |
| IA2 | 14 months after FPE AND targeted PFS event count in BM+ reached | Up to 17 months after FPE (+3 months) | PFS (final for both populations), OS (interim) |
| Final | 24 months after FPE AND targeted OS event count in BM+ reached | Up to 30 months after FPE (+6 months) | OS (final for both populations) |

Notes on timing rules:

- IA1 is calendar-driven to ensure 6 months minimum follow-up for ORR assessment in all patients. PFS and OS are analyzed with whatever event counts are available at the cutoff.
- IA2 is both calendar- and event-driven: we require at least 14 months after FPE to ensure minimum follow-up of nearly 3x the assumed control median PFS, but also require the targeted BM+ PFS event count. If the targeted events are not reached by 14 months after FPE, the analysis may be delayed up to 3 months.
- The final analysis is event-driven for OS in BM+. If the targeted event count is not reached by 24 months after FPE, the analysis may be delayed up to 6 months (30 months after FPE).

## Design

```{r, message=FALSE, warning=FALSE}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
old <- options(scipen = 999)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(gt)
library(ggplot2)
library(gsDesign)
library(gsDesign2)
library(graphicalMCP)

if (!"sequential_pval" %in% getNamespaceExports("gsDesign2")) {
  stop(
    "This vignette requires a version of gsDesign2 that exports `sequential_pval()`.\n",
    "Install the GitHub version from `Merck/gsDesign2`."
  )
}
```

### Multiplicity diagram for hypothesis testing

Alpha is split equally between BM+ and Overall populations for OS, with smaller allocations to PFS and ORR. Reallocation flows from BM+ to Overall within each endpoint, and from Overall to the next endpoint's populations.

```{r, warning=FALSE, message=FALSE, fig.asp=0.9}
### MODIFY HYPOTHESES, ALPHA, AND TRANSITIONS FOR YOUR STUDY
nameHypotheses <- c(
  "H1: OS\n BM+",
  "H2: OS\n Overall",
  "H3: PFS\n BM+",
  "H4: PFS\n Overall",
  "H5: ORR\n BM+",
  "H6: ORR\n Overall"
)
nHypotheses <- length(nameHypotheses)

# Transition weights for alpha reallocation
m <- matrix(c(
  0, 1, 0, 0, 0, 0,
  0, 0, .5, .5, 0, 0,
  0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, .5, .5,
  0, 0, 0, 0, 0, 1,
  .5, .5, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)

# Initial Type I error assigned to each hypothesis (one-sided)
alphaHypotheses <- c(.01, .01, .004, 0.000, 0.0005, .0005)
fwer <- sum(alphaHypotheses)

fill_idx <- c(2, 2, 4, 4, 3, 3)
vertex_color_map <- stats::setNames(cbPalette[fill_idx], nameHypotheses)

g0 <- graphicalMCP::graph_create(
  hypotheses = stats::setNames(alphaHypotheses / fwer, nameHypotheses),
  transitions = m,
  hyp_names = nameHypotheses
)

layout6 <- rbind(
  c(0.28, 0.74),
  c(0.72, 0.74),
  c(0.18, 0.50),
  c(0.82, 0.50),
  c(0.28, 0.26),
  c(0.72, 0.26)
)

# Build vertex labels with alpha character
alphaLabels <- sprintf("\u03b1 = %s", format(alphaHypotheses, scientific = FALSE))
vertexLabels <- paste(nameHypotheses, alphaLabels, sep = "\n")

plot(g0,
  layout = layout6,
  vertex.size = 45,
  vertex.label = vertexLabels,
  vertex.label.cex = 0.9,
  vertex.color = unname(vertex_color_map[names(g0$hypotheses)]),
  margin = 0.25
)
```

### Common design parameters

BM+ prevalence is 60%. Enrollment occurs over 14 months with a 6-month ramp-up.

```{r}
### MODIFY THESE PARAMETERS FOR YOUR STUDY
bm_prev <- 0.60
osmedian <- 12  # Control median OS (months)
pfsmedian <- 5  # Control median PFS (months)
```

### Group sequential designs for each hypothesis {.tabset}

H1 (OS in BM+) is the sample-size-driving hypothesis, designed with `gs_design_ahr()` targeting 90% power.
All other hypotheses use the enrollment rates derived from H1 and compute power with `gs_power_ahr()` (time-to-event) or `fixed_design_rd()` (ORR).

#### H1: OS, BM+ (sample-size-driving)

BM+ subgroup only. Control median OS = `r osmedian` months, HR = 0.65. Target 90% power.
This design determines the BM+ enrollment rate and sample size for all other hypotheses.

```{r}
fail_rate_os_sub <- gsDesign2::define_fail_rate(
  duration = Inf,
  fail_rate = log(2) / osmedian,
  hr = 0.65,
  dropout_rate = 0.001
)

# Initial BM+ enrollment rate for design (will be updated by gs_design_ahr)
enroll_rate_sub_init <- gsDesign2::define_enroll_rate(
  duration = c(2, 2, 2, 8),
  rate = c(0.25, 0.50, 0.75, 1.00)  # Relative rates; gs_design_ahr will scale
)

ossub <- gsDesign2::gs_design_ahr(
  enroll_rate = enroll_rate_sub_init,
  fail_rate = fail_rate_os_sub,
  alpha = alphaHypotheses[1],
  beta = 0.1,
  binding = FALSE,
  analysis_time = c(20, 28, 38),  # FPE+6, FPE+14, FPE+24 (14 mo enrollment + follow-up)
  info_frac = NULL,                # Derived from analysis_time + enroll/fail rates
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[1]),
  test_lower = FALSE
)

gsDesign2::gs_bound_summary(ossub) %>%
  gt::gt() %>%
  gt::tab_header(title = "Design for OS in BM+ (gsDesign2)")
```

#### Derive enrollment rates and sample sizes from H1

The BM+ enrollment rate from H1 determines all other designs.
The BM- enrollment rate is derived from the BM+ prevalence.

```{r}
# Extract BM+ enrollment rate from H1 design
enroll_rate_sub <- ossub$enroll_rate
n_bm_pos <- sum(enroll_rate_sub$rate * enroll_rate_sub$duration)
n_bm_neg <- n_bm_pos * (1 - bm_prev) / bm_prev
n_total <- n_bm_pos + n_bm_neg

cat("BM+ sample size (per arm):", round(n_bm_pos), "\n")
cat("BM- sample size (per arm):", round(n_bm_neg), "\n")
cat("Total sample size (per arm):", round(n_total), "\n")

# Build stratified enrollment rate for overall population designs
enroll_rate_overall <- gsDesign2::define_enroll_rate(
  stratum = rep(c("BM+", "BM-"), each = nrow(enroll_rate_sub)),
  duration = rep(enroll_rate_sub$duration, 2),
  rate = c(
    enroll_rate_sub$rate,                                    # BM+ rates from H1
    enroll_rate_sub$rate * (1 - bm_prev) / bm_prev          # BM- rates scaled by prevalence
  )
)
```

#### H2: OS, Overall

Stratified power calculation using enrollment from H1.

- BM+ (60%): Control median OS = `r osmedian` months, HR = 0.65.
- BM- (40%): Control median OS = `r osmedian` months, HR = 0.85.

```{r}
fail_rate_os_overall <- gsDesign2::define_fail_rate(
  stratum = c("BM+", "BM-"),
  duration = c(Inf, Inf),
  fail_rate = log(2) / osmedian,
  hr = c(0.65, 0.85),
  dropout_rate = 0.001
)

os <- gsDesign2::gs_power_ahr(
  enroll_rate = enroll_rate_overall,
  fail_rate = fail_rate_os_overall,
  ratio = 1,
  event = NULL,
  analysis_time = c(20, 28, 38),  # FPE+6, FPE+14, FPE+24
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[2]),
  test_lower = FALSE,
  binding = FALSE
)

gsDesign2::gs_bound_summary(os) %>%
  gt::gt() %>%
  gt::tab_header(title = "Power for OS in Overall population (gsDesign2)")
```

#### H3: PFS, BM+

Power calculation for BM+ subgroup using enrollment from H1. Control median PFS = `r pfsmedian` months, HR = 0.70.

```{r}
fail_rate_pfs_sub <- gsDesign2::define_fail_rate(
  duration = Inf,
  fail_rate = log(2) / pfsmedian,
  hr = 0.70,
  dropout_rate = 0.001
)

pfssub <- gsDesign2::gs_power_ahr(
  enroll_rate = enroll_rate_sub,
  fail_rate = fail_rate_pfs_sub,
  ratio = 1,
  event = NULL,
  analysis_time = c(20, 28),  # FPE+6, FPE+14
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[3]),
  test_lower = FALSE,
  binding = FALSE
)

gsDesign2::gs_bound_summary(pfssub) %>%
  gt::gt() %>%
  gt::tab_header(title = "Power for PFS in BM+ (gsDesign2)")
```

#### H4: PFS, Overall

Stratified power calculation using enrollment from H1.

- BM+ (60%): Control median PFS = `r pfsmedian` months, HR = 0.70.
- BM- (40%): Control median PFS = `r pfsmedian` months, HR = 0.90.

```{r}
fail_rate_pfs_overall <- gsDesign2::define_fail_rate(
  stratum = c("BM+", "BM-"),
  duration = c(Inf, Inf),
  fail_rate = log(2) / pfsmedian,
  hr = c(0.70, 0.90),
  dropout_rate = 0.001
)

# H4 starts with alpha = 0 but may receive alpha through reallocation.
# Use PFS BM+ alpha for bounds structure; actual testing uses reallocated alpha.
pfs <- gsDesign2::gs_power_ahr(
  enroll_rate = enroll_rate_overall,
  fail_rate = fail_rate_pfs_overall,
  ratio = 1,
  event = NULL,
  analysis_time = c(20, 28),  # FPE+6, FPE+14
  info_scale = "h0_info",
  upper = gsDesign2::gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = alphaHypotheses[3]),  # Use H3's alpha for bounds
  test_lower = FALSE,
  binding = FALSE
)

gsDesign2::gs_bound_summary(pfs) %>%
  gt::gt() %>%
  gt::tab_header(title = "Power for PFS in Overall population (gsDesign2)")
```

#### H5: ORR, BM+

Fixed design for rate difference in BM+ subgroup. Control ORR = 30%, experimental ORR = 45%.

```{r}
orr_ctrl_bm_pos <- 0.30
orr_exp_bm_pos <- 0.45

orr_sub <- gsDesign2::fixed_design_rd(
  alpha = alphaHypotheses[5],
  power = NULL,
  p_c = orr_ctrl_bm_pos,
  p_e = orr_exp_bm_pos,
  rd0 = 0,
  n = ceiling(n_bm_pos) * 2  # Total N (both arms) in BM+
)

summary(orr_sub) %>%
  gt::gt() %>%
  gt::fmt_number(columns = "Bound", decimals = 2) %>%
  gt::fmt_number(columns = "Power", decimals = 3) %>%
  gt::tab_header(title = "Power for ORR in BM+ (fixed_design_rd)")
```

#### H6: ORR, Overall

Fixed design for rate difference in overall population, stratified by biomarker status.

- BM+ (60%): Control ORR = 30%, experimental ORR = 45%.
- BM- (40%): Control ORR = 20%, experimental ORR = 24%.

```{r}
orr_ctrl_bm_neg <- 0.20
orr_exp_bm_neg <- 0.24

orr_ctrl_overall <- orr_ctrl_bm_pos * bm_prev + orr_ctrl_bm_neg * (1 - bm_prev)
orr_exp_overall <- orr_exp_bm_pos * bm_prev + orr_exp_bm_neg * (1 - bm_prev)

orr_all <- gsDesign2::fixed_design_rd(
  alpha = alphaHypotheses[6],
  power = NULL,
  p_c = orr_ctrl_overall,
  p_e = orr_exp_overall,
  rd0 = 0,
  n = ceiling(n_total) * 2  # Total N (both arms) overall
)

summary(orr_all) %>%
  gt::gt() %>%
  gt::fmt_number(columns = "Bound", decimals = 2) %>%
  gt::fmt_number(columns = "Power", decimals = 3) %>%
  gt::tab_header(title = "Power for ORR in Overall population (fixed_design_rd)")
```

### Design list

```{r}
gsD2list <- list(ossub, os, pfssub, pfs, NULL, NULL)
```

### Enrollment and analysis timing

Enrollment occurs over 14 months with a 6-month ramp-up:

- Months 1-2: 25% of steady-state rate
- Months 3-4: 50% of steady-state rate
- Months 5-6: 75% of steady-state rate
- Months 7-14: Full steady-state enrollment

Analysis timing relative to final patient enrolled (FPE):

- **IA1 (FPE + 6 months)**: ORR assessed for all patients. Interim PFS and OS.
- **IA2 (FPE + 14 months)**: Final PFS analysis, requires targeted BM+ PFS events (max extension to FPE + 17 months). Interim OS.
- **Final (FPE + 24 months)**: Final OS analysis, requires targeted BM+ OS events (max extension to FPE + 30 months).

## Results entry at time of analysis

### Timing of analyses and resulting event counts

```{r}
### MODIFY EVENT COUNTS TO MATCH YOUR STUDY
# PFS, overall population
events_pfs_all <- c(675, 750)
# PFS, BM+
events_pfs_sub <- c(265, 310)
# OS, overall population
events_os_all <- c(529, 700, 800)
# OS, BM+
events_os_sub <- c(185, 245, 295)
```

### Nominal p-values for each analysis

Enter nominal 1-sided p-values for each analysis. For analyses not yet performed, use dummy values (e.g., 0.99).
Spending time uses the BM+ information fraction for all hypotheses.

```{r, warnings=FALSE, message=FALSE}
### MODIFY P-VALUES TO MATCH YOUR STUDY
inputResults <- tibble(
  H = c(rep(1, 3), rep(2, 3), rep(3, 2), rep(4, 2), 5, 6),
  Pop = c(
    rep("BM+", 3), rep("Overall", 3),
    rep("BM+", 2), rep("Overall", 2),
    "BM+", "Overall"
  ),
  Endpoint = c(rep("OS", 6), rep("PFS", 4), rep("ORR", 2)),
  nominalP = c(
    .03, .0001, .000001,
    .2, .15, .1,
    .2, .001,
    .3, .2,
    .00001,
    .1
  ),
  Analysis = c(1:3, 1:3, 1:2, 1:2, 1, 1),
  events = c(events_os_sub, events_os_all, events_pfs_sub, events_pfs_all, NA, NA),
  spendingTime = c(
    events_os_sub / max(events_os_sub),
    events_os_sub / max(events_os_sub),
    events_pfs_sub / max(events_pfs_sub),
    events_pfs_sub / max(events_pfs_sub),
    NA, NA
  )
)
kable(inputResults, caption = "DUMMY RESULTS — replace with actual data.") %>%
  kable_styling() %>%
  add_footnote("Dummy results", notation = "none")
```

## Testing hypotheses

### Compute sequential p-values for each hypothesis

```{r, message=FALSE}
### USER SHOULD NOT NEED TO MODIFY THIS CODE
EOCtab <- NULL
EOCtab <- inputResults %>%
  group_by(H) %>%
  slice(1) %>%
  ungroup() %>%
  select("H", "Pop", "Endpoint", "nominalP")
EOCtab$seqp <- .9999
for (EOCtabline in 1:nHypotheses) {
  EOCtab$seqp[EOCtabline] <-
    ifelse(is.null(gsD2list[[EOCtabline]]), EOCtab$nominalP[EOCtabline], {
      tem <- filter(inputResults, H == EOCtabline)
      gsDesign2::sequential_pval(
        gs_design = gsD2list[[EOCtabline]],
        event = tem$events,
        z = -stats::qnorm(tem$nominalP),
        ustime = tem$spendingTime,
        interval = c(1e-05, 0.9999)
      )
    })
}
EOCtab <- EOCtab %>% select(-"nominalP")
```

### Evaluate hypothesis rejection using graphicalMCP

```{r, message=FALSE, warning=FALSE}
library(graphicalMCP)

g0 <- graphicalMCP::graph_create(
  hypotheses = alphaHypotheses / fwer,
  transitions = m,
  hyp_names = nameHypotheses
)
```

```{r}
result <- graphicalMCP::graph_test_shortcut(
  graph = g0,
  p = EOCtab$seqp,
  alpha = fwer,
  verbose = TRUE
)

adj_p <- result$outputs$adjusted_p
rej <- result$outputs$rejected

if (is.numeric(rej)) {
  rej_logical <- rep(FALSE, nHypotheses)
  rej_logical[rej] <- TRUE
} else {
  rej_logical <- as.logical(rej)
}

EOCtab$Rejected <- rej_logical
EOCtab$adjPValues <- adj_p
```

## Verification of hypotheses rejected

```{r}
rejected_order <- which(EOCtab$Rejected)[order(EOCtab$adjPValues[EOCtab$Rejected])]

graphs <- list(g0)
if (length(rejected_order) > 0) {
  gu <- graphicalMCP::graph_update(g0, delete = rejected_order)
  graphs <- gu$intermediate_graphs
}
ngraphs <- length(graphs)

lastWeights <- as.numeric(graphs[[ngraphs]]$hypotheses)
lastGraph <- rep(ngraphs, nrow(EOCtab))

if (length(rejected_order) > 0) {
  for (j in seq_along(rejected_order)) {
    h <- rejected_order[j]
    lastWeights[h] <- as.numeric(graphs[[j]]$hypotheses[h])
    lastGraph[h] <- j
  }
}

EOCtab$lastAlpha <- fwer * lastWeights
EOCtab$lastGraph <- lastGraph
EOCtabx <- EOCtab
names(EOCtabx) <- c(
  "Hypothesis", "Population", "Endpoint", "Sequential p",
  "Rejected", "Adjusted p", "Max alpha allocated", "Last Graph"
)
```

### Multiplicity graph sequence from graphicalMCP {.tabset}

```{r, message=FALSE, warning=FALSE, results='asis', fig.asp=0.95}
### THERE SHOULD BE NO NEED TO MODIFY THIS CODE SECTION
for (i in 1:ngraphs) {
  cat(" \n")
  cat("####", paste(" Graph", as.character(i), " \n\n"))
  gi_alpha <- fwer * as.numeric(graphs[[i]]$hypotheses)
  gi_labels <- paste(
    names(graphs[[i]]$hypotheses),
    sprintf("\u03b1 = %s", format(gi_alpha, scientific = FALSE)),
    sep = "\n"
  )
  plot(graphs[[i]],
    layout = layout6,
    vertex.size = 45,
    vertex.label = gi_labels,
    vertex.label.cex = 0.9,
    vertex.color = unname(vertex_color_map[names(graphs[[i]]$hypotheses)]),
    margin = 0.25
  )
  cat(" \n\n\n")
}
```

### Comparison of sequential p-values to multiplicity graphs

```{r}
EOCtabx %>%
  select(c(1:4, 7, 5:6, 8)) %>%
  kable() %>%
  kable_styling()
```

### Bounds at final alpha allocated for group sequential tests {.tabset}

```{r, results='asis'}
for (i in 1:nHypotheses) {
  cat("####", paste(" Hypothesis", as.character(i), " \n"))
  hresults <- inputResults %>% filter(H == i)
  xx <- paste("Max alpha allocated from above table: ",
    as.character(EOCtab$lastAlpha[i]),
    sep = ""
  )
  d2 <- gsD2list[[i]]
  if (is.null(d2)) {
    cat("Maximum alpha allocated: ")
    cat(EOCtab$lastAlpha[i])
    cat("\n\n")
    cat("Nominal p-value for hypothesis test: ")
    cat(hresults$nominalP)
    cat("\n\n")
  }
  if (!is.null(gsD2list[[i]])) {
    cat("Nominal p-values at each analysis for comparison to bounds in table below:\n\n")
    cat(hresults$nominalP)
    cat("\n\n")
    n.I <- hresults$events
    usTime <- hresults$spendingTime
    if (EOCtab$lastAlpha[i] == 0) {
      cat("Maximum alpha allocated: 0\n\n")
      cat("No testing required\n\n")
    }
    if (EOCtab$lastAlpha[i] > 0) {
      d2_upd <- gsDesign2::gs_update_ahr(
        x = d2,
        alpha = EOCtab$lastAlpha[i],
        ustime = usTime,
        event_tbl = data.frame(analysis = hresults$Analysis, event = n.I)
      )

      d2_upd$bound %>%
        dplyr::filter(.data$bound == "upper") %>%
        dplyr::left_join(
          data.frame(analysis = hresults$Analysis, spending_time = usTime),
          by = "analysis"
        ) %>%
        dplyr::transmute(
          Analysis = .data$analysis,
          Z = .data$z,
          `Nominal p (1-sided)` = .data$`nominal p`,
          `Spending time` = .data$spending_time
        ) %>%
        kable(caption = xx, row.names = FALSE) %>%
        kable_styling() %>%
        cat()
      cat("\n\n")
    }
  }
}
```

```{r}
options(old)
```

## Session information

You can use `sessionInfo()` to document the versions of R and R packages used to render this document.

## References

```{js, echo=FALSE}
// Fix dynamic tabsets code block rendering issue under pkgdown
(function () {
  unstyledCode = document.querySelectorAll("button code");
  if (!unstyledCode.length) {
    return;
  }
  for (let i = 0; i < unstyledCode.length; i++) {
    blockOld = unstyledCode[i].outerHTML;
    blockNew =
      "<pre class='downlit sourceCode r' style='text-align:left;white-space:pre-wrap;'>" +
      blockOld +
      "</pre>";
    unstyledCode[i].outerHTML = blockNew;
  }
  var tabElements = [
    "ul#multiplicity-graph-sequence-from-graphicalmcp li:nth-child(2) button",
    "ul#bounds-at-final-alpha-allocated-for-group-sequential-tests li:nth-child(2) button",
  ];
  for (let i = 0; i < tabElements.length; i++) {
    var el = document.querySelector(tabElements[i]);
    var tab = new bootstrap.Tab(el);
    tab.show();
  }
})();
```
